% Part: counterfactuals
% Chapter: introduction
% Section: strict-conditional

\documentclass[../../../include/open-logic-section]{subfiles}

\begin{document}

\olfileid{cnt}{int}{str}

\olsection{The Strict Conditional}

Lewis introduced the \emph{strict conditional}~$\strictif$ and argued
that it, not the material conditional, corresponds to implication.  In
alethic modal logic, $!A \strictif !B$ can be defined as $\Box(!A \lif
!B)$. A strict conditional is thus true (at a world) iff the
corresponding material conditional is necessary.

How does the strict conditional fare vis-a-vis the paradoxes of the
material conditional?  A strict conditional with a false antecedent
and one with a true consequent, may be true, or it may be
false. Moreover, $(!A \strictif !B) \lor (!B \strictif !A)$ is not
valid. The strict conditional $!A \strictif !B$ is also not equivalent
to $\lnot !A \lor !B$, so it is not truth functional.

We have:
\begin{align}
  !A \strictif !B & \Entails \lnot !A \lor !B
  \text{ but:}\\
  \lnot !A \lor !B & \Entails/ !A \strictif !B\\
  !B & \Entails/ !A \strictif !B\\
  \lnot !A & \Entails/ !A \strictif !B\\
  \lnot(!A \lif !B) & \Entails/ !A \land \lnot !B
  \text{ but:}\\
  !A \land \lnot !B & \Entails \lnot(!A \strictif !B)
  \intertext{However, the strict conditional still supports modus ponens:}
  !A, !A \strictif !B & \Entails !B
  \intertext{The strict conditional agglomerates:}
  !A \strictif !B,  !A \strictif !C & \Entails !A \strictif (!B \land !C)
  \intertext{Antecedent strengthening holds for the strict conditional:}
  !A \strictif !B & \Entails (!A \land !C) \strictif !B
  \intertext{The strict conditional is also transitive:}
  !A \strictif !B, !B \strictif !C & \Entails !A \strictif !C
  \intertext{Finally, the strict conditional is equivalent to its
    contrapositive:}
  !A \strictif !B & \Entails \lnot !B \strictif \lnot !A\\
  \lnot !B \strictif \lnot !A & \Entails !A \strictif !B
\end{align}

\begin{prob}
  Give \Log{S5}-counterexamples to the entailment relations which do not hold
  for the strict conditional, i.e., for:
  \begin{enumerate}
  \item $\lnot p \Entails/ \Box(p \lif q)$
  \item $q \Entails/ \Box(p \lif q)$
  \item $\lnot\Box(p \lif q) \Entails/ p \land \lnot q$
  \item $\Entails/ \Box(p \lif q) \lor \Box(q \lif p)$
  \end{enumerate}
\end{prob}

\begin{prob}
  Show that the valid entailment relations hold for the strict
  conditional by giving \Log{S5}-proofs of:
  \begin{enumerate}
  \item  $\Box(!A \lif !B) \Entails \lnot !A \lor !B$
  \item $!A \land \lnot !B \Entails \lnot\Box(!A \lif !B)$
  \item $!A, \Box(!A \lif !B) \Entails !B$
  \item $\Box(!A \lif !B), \Box(!A \lif !C) \Entails \Box(!A \lif (!B
    \land !C))$
  \item $\Box(!A \lif !B) \Entails \Box((!A \land !C) \lif !B)$
  \item $\Box(!A \lif !B), \Box(!B \lif !C) \Entails \Box(!A
    \lif !C)$
  \item $\Box(!A \lif !B) \Entails \Box(\lnot !B \lif \lnot !A)$
  \item $\Box(\lnot !B \lif \lnot !A) \Entails \Box(!A \lif !B)$
  \end{enumerate}
  \end{prob}

However, the strict conditional still has its own ``paradoxes.'' Just
as a material conditional with a false antecedent or a true consequent
is true, a strict conditional with a \emph{necessarily} false
antecedent or a necessarily true consequent is true. Moreover, any
true strict conditional is necessarily true, and any false strict
conditional is necessarily false. In other words, we have
\begin{align}
  \Box \lnot !A & \Entails !A \strictif !B\\
  \Box !B & \Entails !A \strictif !B\\
  !A \strictif !B& \Entails \Box(!A \strictif !B)\\
  \lnot(!A \strictif !B) & \Entails \Box\lnot(!A \strictif !B)
\end{align}
These are not problems if you think of $\strictif$ as ``implies.''
Logical entailment relationships are, after all, mathematical facts and so
can't be contingent. But they do raise issues if you want to use
$\strictif$ as a logical connective that is supposed to capture ``if
\dots then \dots,'' especially the last two. For surely there are ``if
\dots then \dots'' statements that are contingently true or
contingently false---in fact, they generally are neither necessary nor
impossible.

\begin{prob}
  Give proofs in \Log{S5} of:
  \begin{enumerate}
    \item  $\Box \lnot !A \Entails !A \strictif !B$
    \item $!A \strictif !B \Entails \Box(!A \strictif !B)$
    \item $\lnot(!A \strictif !B)  \Entails \Box\lnot(!A \strictif !B)$
  \end{enumerate}
  Use the definition of $\strictif$ to do so.
\end{prob}

\end{document}
