% Part: incompleteness
% Chapter: incompleteness-provability
% Section: introduction

\documentclass[../../include/open-logic-section]{subfiles}

\begin{document}

\olfileid{inc}{inp}{int}

\olsection{Introduction}

Hilbert thought that a system of axioms for a mathematical structure,
such as the natural numbers, is inadequate unless it allows one to
derive all true statements about the structure. Combined with his
later interest in formal systems of deduction, this suggests that one
should try to guarantee that, say, the formal system one is using to
reason about the natural numbers is not only consistent, but also
\emph{complete}, i.e., every statement is either provable or
refutable. G\"odel's first incompleteness theorem shows that no such
system of axioms exists: there is no complete, consistent, effectively
axiomatized formal system for arithmetic.  In fact, no ``sufficiently
strong,'' consistent, effectively axiomatized mathematical theory is
complete.

A more important goal of Hilbert's, the centerpiece of his program for
the justification of modern (``classical'') mathematics, was to find
finitary consistency proofs for formal systems representing classical
reasoning.  With regard to Hilbert's program, then, G\"odel's second
incompleteness theorem was a much bigger blow.

The second incompleteness theorem can be stated in vague terms, like
the first incompleteness theorem. Roughly speaking, then, it says that
no sufficiently strong theory of arithmetic can prove its own
consistency. We will have to take ``sufficiently strong'' to include a
little bit more than $\Th{Q}$.

The idea behind G\"odel's original proof of the incompleteness theorem
can be found in the Epimenides paradox. Epimenides, a Cretin, asserted
that all Cretans are liars; a more direct form of the paradox is the
assertion ``this sentence is false.'' Essentially, by replacing truth
with provability, G\"odel was able to formalize a sentence which, in
essence, asserts ``this sentence is not provable.'' Assuming
$\omega$-consistency---a property stronger than consistency---G\"odel
was able to show that this sentence is neither provable nor refutable
from the system of axioms he was considering.

The first challenge is to understand how one can construct a sentence
that refers to itself. For every formula $!A$ in the language of $\Th{Q}$,
let $\gn{!A}$ denote the numeral corresponding to $\Gn{!A}$. Think
about what this means: $!A$ is a formula in the language of $\Th{Q}$,
$\Gn{!A}$ is a natural number, and $\gn{!A}$ is a \emph{term} in the
language of $\Th{Q}$. So every formula $!A$ in the language of $\Th{Q}$ has a
\emph{name}, $\gn{!A}$, which is a term in the language of $\Th{Q}$; this
provides us with a conceptual framework in which formulas in the
language of $\Th{Q}$ can ``say'' things about other formulas. The following
lemma is known as G\"odel's fixed-point lemma.

\begin{lem}
  Let $\Th{T}$ be any theory extending $\Th{Q}$, and let $!B(x)$ be any
  formula with free variable $x$. Then there is a sentence $!A$ such
  that $\Th{T}$ proves $!A \liff !B(\gn{!A})$.
\end{lem}

The lemma asserts that given any property $!B(x)$, there is a
sentence $!A$ that asserts ``$!B(x)$ is true of me.''

How can we construct such a sentence? Consider the following version
of the Epimenides paradox, due to Quine:
\begin{quote}
``Yields falsehood when preceded by its quotation'' yields falsehood
when preceded by its quotation.
\end{quote}
This sentence is not directly self-referential. It simply makes an
assertion about the syntactic objects between quotes, and, in doing
so, it is on par with sentences like
\begin{enumerate}
\item ``Robert'' is a nice name.
\item ``I ran.'' is a short sentence.
\item ``Has three words'' has three words.
\end{enumerate}
But what happens when one takes the phrase ``yields falsehood when
preceded by its quotation,'' and precedes it with a quoted version of
itself? Then one has the original sentence! In short, the sentence
asserts that it is false.

\end{document}
