% Part: incompleteness
% Chapter: incompleteness-provability
% Section: provability-conditions

\documentclass[../../include/open-logic-section]{subfiles}

\begin{document}

\olfileid{inc}{inp}{prc}

\olsection{The Provability Conditions for $\Th{PA}$}

Peano arithmetic, or $\Th{PA}$, is the theory extending $\Th{Q}$ with
induction axioms for all formulas. In other words, one adds to $\Th{Q}$
axioms of the form
\[
(!A(0) \land \lforall[x][(!A(x) \lif !A(x'))]) \lif \lforall[x][!A(x)]
\]
for every formula $!A$. Notice that this is really a {\em schema},
which is to say, infinitely many axioms (and it turns out that
$\Th{PA}$ is {\em not} finitely axiomatizable). But since one can
effectively determine whether or not a string of symbols is an
instance of an induction axiom, the set of axioms for $\Th{PA}$ is
computable. $\Th{PA}$ is a much more robust theory than $\Th{Q}$. For
example, one can easily prove that addition and multiplication are
commutative, using induction in the usual way. In fact, most finitary
number-theoretic and combinatorial arguments can be carried out in
$\Th{PA}$.

Since $\Th{PA}$ is computably axiomatized, the provability predicate
$\Prf[PA](x,y)$ is computable and hence represented in $\Th{Q}$ (and
so, in $\Th{PA}$). As before, I will take $\OPrf[PA](x,y)$ to denote
the formula representing the relation.  Let $\OProv[PA](y)$ be the
formula $\lexists[x][\Prf[PA](x,y)]$, which, intuitively says, ``$y$ is
provable from the axioms of $\Th{PA}$.''  The reason we need a little
bit more than the axioms of $\Th{Q}$ is we need to know that the
theory we are using is strong enough to prove a few basic facts about
this provability predicate. In fact, what we need are the following
facts:
\begin{enumerate}
\item If $\Th{PA} \Proves !A$, then $\Th{PA} \Proves
  \OProv[PA](\gn{!A})$
\item For every formula $!A$ and $!B$, $\Th{PA} \Proves
  \OProv[PA](\gn{!A \lif !B}) \lif (\OProv[PA](\gn{!A}) \lif
  \OProv[PA](\gn{!B}))$
\item For every formula $!A$, $\Th{PA} \Proves \OProv[PA](\gn{!A})
\lif \OProv[PA](\gn{\OProv[PA](\gn{!A})})$.
\end{enumerate}
The only way to verify that these three properties hold is to describe
the formula $\OProv[PA](y)$ carefully and use the axioms of $\Th{PA}$ to
describe the relevant formal proofs. Clauses 1 and 2 are easy; it is
really clause 3 that requires work. (Think about what kind of work it
entails\dots) Carrying out the details would be tedious and
uninteresting, so here we will ask you to take it on faith the
$\Th{PA}$ has the three properties listed above. A reasonable choice
of $\OProv[PA](y)$ will also satisfy
\begin{enumerate}
\item[4.] If $\Th{PA}$ proves $\OProv[PA](\gn{!A})$, then $\Th{PA}$ proves
  $!A$.
\end{enumerate}
But we will not need this fact.

\begin{digress}
Incidentally, G\"odel was lazy in the same way we are
being now. At the end of the 1931 paper, he sketches the proof of the
second incompleteness theorem, and promises the details in a later
paper. He never got around to it; since everyone who understood the
argument believed that it could be carried out, he did not need to
fill in the details.)
\end{digress}

\end{document}
