% Part: first-order-logic
% Chapter: natural-deduction
% Section: provability-consistency

\documentclass[../../../include/open-logic-section]{subfiles}

\begin{document}

\iftag{FOL}
      {\olfileid{fol}{ntd}{prv}}
      {\olfileid{pl}{ntd}{prv}}
      
\olsection{\usetoken{S}{derivability} and Consistency}

We will now establish a number of properties of the !!{derivability}
relation.  They are independently interesting, but each will play a
role in the proof of the completeness theorem.

\begin{prop}\ollabel{prop:provability-contr}
  If $\Gamma \Proves !A$ and $\Gamma \cup \{!A\}$ is inconsistent,
  then $\Gamma$ is inconsistent.
\end{prop}

\begin{proof}
Let the !!{derivation} of~$!A$ from~$\Gamma$ be~$\delta_1$ and the
!!{derivation} of~$\lfalse$ from $\Gamma \cup \{!A\}$
be~$\delta_2$. We can then !!{derive}:
\begin{prooftree}
\AxiomC{$\Gamma, \Discharge{!A}{1}$}
\RightLabel{$\delta_2$}
\DeduceC{$\lfalse$}
\DischargeRule{\Intro{\lnot}}{1}
\UnaryInfC{$\lnot !A$}
\AxiomC{$\Gamma$}
\RightLabel{$\delta_1$}
\DeduceC{$!A$}
\RightLabel{\Elim{\lnot}}
\BinaryInfC{$\lfalse$}
\end{prooftree}
In the new !!{derivation}, the assumption~$!A$ is !!{discharged}, so it is
!!a{derivation} from~$\Gamma$.
\end{proof}

\begin{prop}
\ollabel{prop:prov-incons}
$\Gamma \Proves !A$ iff $\Gamma \cup \{\lnot !A\}$ is inconsistent.
\end{prop}

\begin{proof}
First suppose $\Gamma \Proves !A$, i.e., there is
!!a{derivation}~$\delta_0$ of~$!A$ from !!{undischarged}
assumptions~$\Gamma$. We obtain !!a{derivation} of $\lfalse$ from
$\Gamma \cup \{\lnot !A\}$ as follows:
\begin{prooftree}
  \AxiomC{$\lnot !A$}
  \AxiomC{$\Gamma$}
  \RightLabel{$\delta_0$}
  \DeduceC{$!A$}
  \RightLabel{\Elim{\lnot}}
  \BinaryInfC{$\lfalse$}
\end{prooftree}

Now assume $\Gamma \cup \{\lnot !A\}$ is inconsistent, and let
$\delta_1$ be the corresponding derivation of $\lfalse$ from
!!{undischarged} assumptions in~$\Gamma \cup \{\lnot !A\}$. We obtain
!!a{derivation} of~$!A$ from~$\Gamma$ alone by using~$\FalseCl$:
\begin{prooftree}
  \AxiomC{$\Gamma, \Discharge{\lnot !A}{1}$}
  \RightLabel{$\delta_1$}
  \DeduceC{$\lfalse$}
  \RightLabel{\FalseCl}
  \DischargeRule{\FalseCl}{1}
  \UnaryInfC{$!A$}
\end{prooftree}
\end{proof}

\begin{prob}
Prove that $\Gamma \Proves \lnot !A$ iff $\Gamma \cup \{!A\}$ is
inconsistent.
\end{prob}

\begin{prop}\ollabel{prop:explicit-inc}
  If $\Gamma \Proves !A$ and $\lnot !A \in \Gamma$, then $\Gamma$ is
  inconsistent.
\end{prop}

\begin{proof}
  Suppose $\Gamma \Proves !A$ and $\lnot !A \in \Gamma$.  Then there
  is !!a{derivation}~$\delta$ of~$!A$ from~$\Gamma$. Consider this
    simple application of the $\Elim{\lnot}$ rule:
  \begin{prooftree}
    \AxiomC{$\lnot !A$}
    \AxiomC{$\Gamma$}
    \RightLabel{$\delta$}
    \DeduceC{$!A$}
    \RightLabel{\Elim{\lnot}}
    \BinaryInfC{$\lfalse$}
  \end{prooftree}
  Since $\lnot !A \in \Gamma$, all !!{undischarged} assumptions are
  in~$\Gamma$, this shows that $\Gamma \Proves \lfalse$.
\end{proof}

\begin{prop}\ollabel{prop:provability-exhaustive}
  If $\Gamma \cup \{!A\}$ and $\Gamma \cup \{\lnot !A\}$ are both
  inconsistent, then $\Gamma$ is inconsistent.
\end{prop}

\begin{proof}
There are !!{derivation}s $\delta_1$ and $\delta_2$ of~$\lfalse$ from
  $\Gamma \cup \{ !A \}$ and $\lfalse$ from $\Gamma \cup \{ \lnot !A
  \}$, respectively. We can then !!{derive}
\begin{prooftree}
\AxiomC{$\Gamma, \Discharge{\lnot !A}{2}$}
\RightLabel{$\delta_2$}
\DeduceC{$\lfalse$}
\DischargeRule{\Intro{\lnot}}{2}
\UnaryInfC{$\lnot \lnot !A$}
\AxiomC{$\Gamma, \Discharge{!A}{1}$}
\RightLabel{$\delta_1$}
\DeduceC{$\lfalse$}
\DischargeRule{\Intro{\lnot}}{1}
\UnaryInfC{$\lnot !A$}
\RightLabel{\Elim{\lnot}}
\BinaryInfC{$\lfalse$}
\end{prooftree}
Since the assumptions $!A$ and $\lnot !A$ are !!{discharged}, this is
!!a{derivation} of~$\lfalse$ from~$\Gamma$ alone. Hence $\Gamma$ is
inconsistent.
\end{proof}

\end{document}
