% Part: first-order-logic
% Chapter: natural-deduction
% Section: proving-things

\documentclass[../../../include/open-logic-section]{subfiles}

\begin{document}

\olfileid{fol}{ntd}{pro}

\olsection{Examples of \usetoken{P}{derivation}}

\begin{ex}
Let's give a !!{derivation} of the !!{formula} $(!A \land !B) \lif !A$.

We begin by writing the desired end-!!{formula} at the bottom of the 
!!{derivation}.
\[
\AxiomC{}
\UnaryInfC{$(!A\land !B) \lif !A$}
\DisplayProof
\]

Next, we need to figure out what kind of inference could result in
!!a{formula} of this form. The !!{main operator} of the
end-!!{formula} is $\lif$, so we'll try to arrive at the
end-!!{formula} using the \Intro{\lif} rule. It is best to write down
the assumptions involved and label the inference rules as you
progress, so it is easy to see whether all assumptions have been
!!{discharged} at the end of the proof.
\[
\AxiomC{$\Discharge{!A \land !B}{1}$}
\DeduceC{$!A$}
\DischargeRule{\Intro{\lif}}{1} 
\UnaryInfC{$(!A\land !B) \lif !A$}
\DisplayProof
\]

We now need to fill in the steps from the assumption $!A \land !B$ to $!A$.
Since we only have one connective to deal with, $\land$, we must
use the $\land$ elim rule. This gives us the following proof:
\[
\AxiomC{$\Discharge{!A \land !B}{1}$}
\RightLabel{\Elim{\land}}
\UnaryInfC{$!A$}
\DischargeRule{\Intro{\lif}}{1} 
\UnaryInfC{$(!A\land !B) \lif !A$}
\DisplayProof
\]

We now have a correct !!{derivation} of the formula $(!A \land
!B) \lif !A$.
\end{ex}

\begin{ex}
Now let's give a !!{derivation} of the !!{formula} $(\lnot !A \lor !B)
\lif (!A \lif !B)$.

We begin by writing the desired end-!!{formula} at the bottom of the 
derivation.
\[
\AxiomC{}
\UnaryInfC{$(\lnot !A \lor !B) \lif (!A \lif !B)$}
\DisplayProof
\]
To find a logical rule that could give us this end-!!{formula}, we
look at the logical connectives in the end-!!{formula}: $\lnot$,
$\lor$, and $\lif$. We only care at the moment about the first
occurence of $\lif$ because it is the !!{main operator} of the
!!{sentence} in the end-sequent, while $\lnot$, $\lor$ and the second
occurence of $\lif$ are inside the scope of another connective, so we
will take care of those later. We therefore start with the
\Intro{\lif} rule.  A correct application must look as follows:
\[
\AxiomC{$\Discharge{\lnot !A \lor !B}{1}$}
\DeduceC{$!A \lif !B$}
\DischargeRule{\Intro{\lif}}{1}
\UnaryInfC{$(\lnot !A \lor !B) \lif (!A \lif !B)$}
\DisplayProof
\]

This leaves us with two possibilities to continue. Either we can
keep working from the bottom up and look for another application
of the \Intro{\lif} rule, or we can work from the top down and apply a
\Elim{\lor} rule. Let us apply the latter. We will use the assumption
$\lnot !A \lor !B$ as the leftmost premise of \Elim{\lor}.  For a valid
application of \Elim{\lor}, the other two premises must be identical
to the conclusion $!A \lif !B$, but each may be derived in turn from
another assumption, namely the two disjuncts of $\lnot !A \lor !B$.
So our !!{derivation} will look like this:
\[
\AxiomC{$\Discharge{\lnot !A \lor !B}{1}$}
\AxiomC{$\Discharge{\lnot !A}{2}$}
\DeduceC{$!A \lif !B$}
\AxiomC{$\Discharge{!B}{2}$}
\DeduceC{$!A \lif !B$}
\DischargeRule{\Elim{\lor}}{2}
\TrinaryInfC{$!A \lif !B$}
\DischargeRule{\Intro{\lif}}{1} 
\UnaryInfC{$(\lnot !A \lor !B) \lif (!A \lif !B)$}
\DisplayProof
\]

In each of the two branches on the right, we want to !!{derive} $!A
\lif !B$, which is best done using \Intro{\lif}.
\[
\AxiomC{$\Discharge{\lnot !A \lor !B}{1}$}
\AxiomC{$\Discharge{\lnot !A}{2}, \Discharge{!A}{3}$}
\DeduceC{$!B$}
\DischargeRule{\Intro{\lif}}{3}
\UnaryInfC{$!A \lif !B$}
\AxiomC{$\Discharge{!B}{2}, \Discharge{!A}{4}$}
\DeduceC{$!B$}
\DischargeRule{\Intro{\lif}}{4}
\UnaryInfC{$!A \lif !B$}
\DischargeRule{\Elim{\lor}}{2}
\TrinaryInfC{$!A \lif !B$}
\DischargeRule{\Intro{\lif}}{1} 
\UnaryInfC{$(\lnot !A \lor !B) \lif (!A \lif !B)$}
\DisplayProof
\]

For the two missing parts of the !!{derivation}, we need
!!{derivation}s of $!B$ from $\lnot !A$ and $!A$ in the middle, and
from $!A$ and $!B$ on the left.  Let's take the former first. $\lnot
!A$ and $!A$ are the two premises of \Elim{\lnot}:
\[
\AxiomC{$\Discharge{\lnot !A}{2}$}
\AxiomC{$\Discharge{!A}{3}$}
\RightLabel{\Elim{\lnot}}
\BinaryInfC{$\lfalse$}
\DeduceC{$!B$}
\DisplayProof
\]
By using \FalseInt, we can obtain $!B$ as a conclusion and
complete the branch.
\[
\AxiomC{$\Discharge{\lnot !A \lor !B}{1}$}
\AxiomC{$\Discharge{\lnot !A}{2}$}
\AxiomC{$\Discharge{!A}{3}$}
\RightLabel{\Intro{\lfalse}}
\BinaryInfC{$\lfalse$}
\RightLabel{\FalseInt}
\UnaryInfC{$!B$}
\DischargeRule{\Intro{\lif}}{3}
\UnaryInfC{$!A \lif !B$}
\AxiomC{$\Discharge{!B}{2}, \Discharge{!A}{4}$}
\DeduceC{$!B$}
\DischargeRule{\Intro{\lif}}{4}
\UnaryInfC{$!A \lif !B$}
\DischargeRule{\Elim{\lor}}{2}
\TrinaryInfC{$!A \lif !B$}
\DischargeRule{\Intro{\lif}}{1} 
\UnaryInfC{$(\lnot !A \lor !B) \lif (!A \lif !B)$}
\DisplayProof
\]

Let's now look at the rightmost branch.  Here it's important to
realize that the definition of !!{derivation} \emph{allows assumptions
  to be discharged} but \emph{does not require} them to be.  In other
words, if we can derive $!B$ from one of the assumptions $!A$ and $!B$
without using the other, that's ok.  And to !!{derive} $!B$ from~$!B$
is trivial: $!B$ by itself is such a !!{derivation}, and no inferences
are needed.  So we can simply delete the assumtion $!A$.
\[
\AxiomC{$\Discharge{\lnot !A \lor !B}{1}$}
\AxiomC{$\Discharge{\lnot !A}{2}$}
\AxiomC{$\Discharge{!A}{3}$}
\RightLabel{\Elim{\lnot}}
\BinaryInfC{$\lfalse$}
\RightLabel{\FalseInt}
\UnaryInfC{$!B$}
\DischargeRule{\Intro{\lif}}{3}
\UnaryInfC{$!A \lif !B$}
\AxiomC{$\Discharge{!B}{2}$}
\RightLabel{\Intro{\lif}}
\UnaryInfC{$!A \lif !B$}
\DischargeRule{\Elim{\lor}}{2}
\TrinaryInfC{$!A \lif !B$}
\DischargeRule{\Intro{\lif}}{1} 
\UnaryInfC{$(\lnot !A \lor !B) \lif (!A \lif !B)$}
\DisplayProof
\]
Note that in the finished !!{derivation}, the rightmost \Intro{\lif}
inference does not actually discharge any assumptions.
\end{ex}

\begin{ex}
So far we have not needed the \FalseCl{} rule. It is special in that
it allows us to discharge !!a{formula} that isn't a sub-!!{formula} of
the conclusion of the rule.  It is closely related to the \FalseInt{}
rule; in fact, the \FalseInt{} rule is a special case of the
\FalseCl{} rule---there is a logic called ``intuitionistic logic'' in
which only \FalseInt{} is allowed.  The \FalseCl{} rule is a last
resort when nothing else works.  For instance, suppose we want to
!!{derive} $!A \lor \lnot !A$. Our usual strategy would be to attempt
to !!{derive} $!A \lor \lnot !A$ using $\Intro{\lor}$. But this would
require us to !!{derive} either $!A$ or $\lnot !A$ from no
assumptions, and this can't be done. \FalseCl{} to the rescue!
\begin{prooftree}
  \AxiomC{$\Discharge{\lnot(!A \lor \lnot !A)}{1}$}
  \DeduceC{$\lfalse$}
  \DischargeRule{\FalseCl}{1}
  \UnaryInfC{$!A \lor \lnot !A$}
\end{prooftree}
Now we're looking for !!a{derivation} of $\lfalse$ from $\lnot(!A \lor
\lnot !A)$. Since $\lfalse$ is the conclusion of $\Elim{\lnot}$ we
might try that:
\begin{prooftree}
  \AxiomC{$\Discharge{\lnot(!A \lor \lnot !A)}{1}$}
  \DeduceC{$\lnot !A$}
  \AxiomC{$\Discharge{\lnot(!A \lor \lnot !A)}{1}$}
  \DeduceC{$!A$}
  \RightLabel{\Elim{\lnot}}
  \BinaryInfC{$\lfalse$}
  \DischargeRule{\FalseCl}{1}
  \UnaryInfC{$!A \lor \lnot !A$}
\end{prooftree}
Our strategy for finding a !!{derivation} of~$\lnot !A$ calls for an
application of~$\Intro{\lnot}$:
\begin{prooftree}
  \AxiomC{$\Discharge{\lnot(!A \lor \lnot !A)}{1}, \Discharge{!A}{2}$}
  \DeduceC{$\lfalse$}
  \DischargeRule{\Intro{\lnot}}{2}
  \UnaryInfC{$\lnot !A$}
  \AxiomC{$\Discharge{\lnot(!A \lor \lnot !A)}{1}$}
  \DeduceC{$!A$}
  \RightLabel{\Elim{\lnot}}
  \BinaryInfC{$\lfalse$}
  \DischargeRule{\FalseCl}{1}
  \UnaryInfC{$!A \lor \lnot !A$}
\end{prooftree}
Here, we can get $\lfalse$ easily by applying $\Elim{\lnot}$ to the
assumption $\lnot(!A \lor \lnot !A)$ and $!A \lor \lnot !A$ which
follows from our new assumption $!A$ by~$\Intro{\lor}$:
\begin{prooftree}
  \AxiomC{$\Discharge{\lnot(!A \lor \lnot !A)}{1}$}
  \AxiomC{$\Discharge{!A}{2}$}
  \RightLabel{\Intro{\lor}}
  \UnaryInfC{$!A \lor \lnot !A$}
  \RightLabel{\Elim{\lnot}}
  \BinaryInfC{$\lfalse$}
  \DischargeRule{\Intro{\lnot}}{2}
  \UnaryInfC{$\lnot !A$}
  \AxiomC{$\Discharge{\lnot(!A \lor \lnot !A)}{1}$}
  \DeduceC{$!A$}
  \RightLabel{\Elim{\lnot}}
  \BinaryInfC{$\lfalse$}
  \DischargeRule{\FalseCl}{1}
  \UnaryInfC{$!A \lor \lnot !A$}
\end{prooftree}
On the right side we use the same strategy, except we get $!A$ by \FalseCl:
\begin{prooftree}
  \AxiomC{$\Discharge{\lnot(!A \lor \lnot !A)}{1}$}
  \AxiomC{$\Discharge{!A}{2}$}
  \RightLabel{\Intro{\lor}}
  \UnaryInfC{$!A \lor \lnot !A$}
  \RightLabel{\Elim{\lnot}}
  \BinaryInfC{$\lfalse$}
  \DischargeRule{\Intro{\lnot}}{2}
  \UnaryInfC{$\lnot !A$}
  \AxiomC{$\Discharge{\lnot(!A \lor \lnot !A)}{1}$}
  \AxiomC{$\Discharge{\lnot !A}{3}$}
  \RightLabel{\Intro{\lor}}
  \UnaryInfC{$!A \lor \lnot !A$}
  \RightLabel{\Elim{\lnot}}
  \BinaryInfC{$\lfalse$}
  \DischargeRule{\FalseCl}{3}
  \UnaryInfC{$!A$}
  \RightLabel{\Elim{\lnot}}
  \BinaryInfC{$\lfalse$}
  \DischargeRule{\FalseCl}{1}
  \UnaryInfC{$!A \lor \lnot !A$}
\end{prooftree}
\end{ex}

\begin{ex}
When dealing with quantifiers, we have to make sure not to violate the
eigenvariable condition, and sometimes this requires us to play around
with the order of carrying out certain inferences. In general, it
helps to try and take care of rules subject to the eigenvariable
condition first (they will be lower down in the finished proof).

Let's see how we'd give a !!{derivation} of the !!{formula}
$\lexists[x][\lnot !A(x)] \lif \lnot \lforall[x][!A(x)]$.
Starting as usual, we write
\[
\AxiomC{}
\UnaryInfC{$\lexists[x][\lnot !A(x)]\lif \lnot \lforall[x][!A(x)]$}
\DisplayProof
\]
We start by writing down what it would take to justify that last step
using the \Intro{\lif} rule.
\[
\AxiomC{$\Discharge{\lexists[x][\lnot !A(x)]}{1}$}
\DeduceC{$\lnot \lforall[x][!A(x)]$}
\RightLabel{\Intro{\lif}}
\UnaryInfC{$\lexists[x][\lnot !A(x)]\lif \lnot \lforall[x][!A(x)]$}
\DisplayProof
\]
Since there is no obvious rule to apply to $\lnot \lforall[x][!A(x)]$,
we will proceed by setting up the !!{derivation} so we can use the
\Elim{\lexists} rule. Here we must pay attention to the eigenvariable
condition, and choose a constant that does not appear in
$\lexists[x][!A(x)]$ or any assumptions that it depends on.
(Since no !!{constant}s appear, however, any choice will do fine.)
\[
\AxiomC{$\Discharge{\lexists[x][\lnot !A(x)]}{1}$}
\AxiomC{$\Discharge{\lnot !A(a)}{2}$}
\DeduceC{$\lnot \lforall[x][!A(x)]$}
\DischargeRule{\Elim{\lexists}}{2}
\BinaryInfC{$\lnot \lforall[x][!A(x)]$}
\RightLabel{\Intro{\lif}}
\UnaryInfC{$\lexists[x][\lnot !A(x)] \lif \lnot \lforall[x][!A(x)]$}
\DisplayProof
\]
In order to derive $\lnot \lforall[x][!A(x)]$, we will attempt to use
the \Intro{\lnot} rule: this requires that we derive a contradiction,
possibly using $\lforall[x][!A(x)]$ as an additional assumption. Of
coursem, this contradiction may involve the assumption $\lnot !A(a)$
which will be discharged by the \Intro{\lif} inference. We can set it
up as follows:
\[
\AxiomC{$\Discharge{\lexists[x][\lnot !A(x)]}{1}$}
\AxiomC{$\Discharge{\lnot !A(a)}{2}, \Discharge{\lforall[x][!A(x)]}{3}$}
\DeduceC{$\lfalse$}
\DischargeRule{\Intro{\lnot}}{3}
\UnaryInfC{$\lnot \lforall[x][!A(x)]$}
\DischargeRule{\Elim{\lexists}}{2}
\BinaryInfC{$\lnot \lforall[x][!A(x)]$}
\RightLabel{\Intro{\lif}}
\UnaryInfC{$\lexists[x][\lnot !A(x)]\lif \lnot \lforall[x][!A(x)]$}
\DisplayProof
\]
It looks like we are close to getting a contradiction. The easiest
rule to apply is the \Elim{\lforall}, which has no eigenvariable
conditions. Since we can use any term we want to replace the
universally quantified~$x$, it makes the most sense to continue
using~$a$ so we can reach a contradiction.
\[
\AxiomC{$\Discharge{\lexists[x][\lnot !A(x)]}{1}$}
\AxiomC{$\Discharge{\lnot !A(a)}{2}$}
\AxiomC{$\Discharge{\lforall[x][!A(x)]}{3}$}
\RightLabel{\Elim{\lforall}}
\UnaryInfC{$!A(a)$}
\RightLabel{\Elim{\lnot}}
\BinaryInfC{$\lfalse$}
\DischargeRule{\Intro{\lnot}}{3}
\UnaryInfC{$\lnot \lforall[x][!A(x)]$}
\DischargeRule{\Elim{\lexists}}{2}
\BinaryInfC{$\lnot \lforall[x][!A(x)]$}
\RightLabel{\Intro{\lif}}
\UnaryInfC{$\lexists[x][\lnot !A(x)]\lif \lnot \lforall[x][!A(x)]$}
\DisplayProof
\]

It is important, especially when dealing with quantifiers, to double
check at this point that the eigenvariable condition has not been
violated. Since the only rule we applied that is subject to the
eigenvariable condition was \Elim{\exists}, and the eigenvariable~$a$
does not occur in any assumptions it depends on, this is a
correct derivation.
\end{ex}

\begin{prob}
Give !!{derivation}s of the following !!{formula}s:
\begin{enumerate}
\item $\lnot(!A \lif !B) \lif (!A \land \lnot !B)$
\item $\lforall[x][(!A(x) \lif !B)] \lif (\lexists[y][!A(y)] \lif !B)$
\end{enumerate}
\end{prob}

\begin{ex}
Sometimes we may derive a !!{formula} from other !!{formula}s.
In these cases, we may have undischarged assumptions. It is 
important to keep track of our assumptions as well
as the end goal.

Let's see how we'd give a !!{derivation} of the !!{formula}
$\lexists[x][!C(x,b)]$ from the assumptions $\lexists[x][(!A(x) 
\land !B(x))]$ and $\lforall[x][(!B(x) \lif !C(x,b)]$.
Starting as usual, we write the end-!!{formula} at the
bottom.
\[
\AxiomC{}
\UnaryInfC{$\lexists[x][!C(x,b)]$}
\DisplayProof
\]

We have two premises to work with. To use the first, i.e., try to find
a !!{derivation} of $\lexists[x][!C(x, b)]$ from $\lexists[x][(!A(x)
  \land !B(x))]$ we would use the \Elim{\lexists} rule. Since it has
an eigenvariable condition, we will apply that rule first. We get the
following:
\[
\AxiomC{$\lexists[x][(!A(x) \land !B(x))]$}
\AxiomC{$\Discharge{!A(a) \land !B(a)}{1}$}
\DeduceC{$\lexists[x][!C(x,b)]$}
\DischargeRule{\Elim{\lexists}}{1}
\BinaryInfC{$\lexists[x][!C(x,b)]$}
\DisplayProof
\]
The two assumptions we are working with share~$!B$.  It may be useful
at this point to apply \Elim{\land} to separate out~$!B(a)$.
\[
\AxiomC{$\lexists[x][(!A(x) \land !B(x)])$}
\AxiomC{$\Discharge{!A(a) 
\land !B(a)}{1}$}
\RightLabel{\Elim{\land}}
\UnaryInfC{$!B(a)$}
\DeduceC{$\lexists[x][!C(x,b)]$}
\DischargeRule{\Elim{\lexists}}{1}
\BinaryInfC{$\lexists[x][!C(x,b)]$}
\DisplayProof
\]

The second assumption we have to work with is~$\lforall[x][(!B(x) \lif
  !C(x,b)]$. Since there is no eigenvariable condition we can
instantiate $x$ with the !!{constant}~$a$ using \Elim{\lforall} to get
$!B(a) \lif !C(a, b)$.  We now have both $!B(a) \lif !C(a,b)$ and
$!B(a)$. Our next move should be a straightforward application of the
\Elim{\lif} rule.
\[
\AxiomC{$\lexists[x][(!A(x) \land !B(x))]$}
\AxiomC{$\lforall[x][(!B(x) \lif !C(x,b))]$}
\RightLabel{\Elim{\lforall}}
\UnaryInfC{$!B(a) \lif !C(a,b)$}
\AxiomC{$\Discharge{!A(a) 
\land !B(a)}{1}$}
\RightLabel{\Elim{\land}}
\UnaryInfC{$!B(a)$}
\RightLabel{\Elim{\lif}}
\BinaryInfC{$!C(a,b)$}
\DeduceC{$\lexists[x][!C(x,b)]$}
\DischargeRule{\Elim{\lexists}}{1}
\BinaryInfC{$\lexists[x][!C(x,b)]$}
\DisplayProof
\]
We are so close!{} One application of \Intro{\lexists} and we
have reached our goal.
\[
\AxiomC{$\lexists[x][(!A(x) \land !B(x))]$}
\AxiomC{$\lforall[x][(!B(x) \lif !C(x,b))]$}
\RightLabel{\Elim{\lforall}}
\UnaryInfC{$!B(a) \lif !C(a,b)$}
\AxiomC{$\Discharge{!A(a) 
\land !B(a)}{1}$}
\RightLabel{\Elim{\land}}
\UnaryInfC{$!B(a)$}
\RightLabel{\Elim{\lif}}
\BinaryInfC{$!C(a,b)$}
\RightLabel{\Intro{\lexists}}
\UnaryInfC{$\lexists[x][!C(x,b)]$}
\DischargeRule{\Elim{\lexists}}{1}
\BinaryInfC{$\lexists[x][!C(x,b)]$}
\DisplayProof
\]
Since we ensured at each step that the eigenvariable
conditions were not violated, we can be confident that this
is a correct derivation.
\end{ex}

\begin{ex}
Give a !!{derivation} of the !!{formula}
$\lnot\lforall[x][!A(x)]$ from the assumptions $\lforall[x][!A(x)] 
\lif \lexists[y][!B(y)]$ and $\lnot\lexists[y][!B(y)]$.
Starting as usual, we write the target !!{formula} at the bottom.
\[
\AxiomC{}
\UnaryInfC{$\lnot\lforall[x][!A(x)]$}
\DisplayProof
\]
The last line of the !!{derivation} is a negation, so let's try using
\Intro{\lnot}. This will require that we figure out how to !!{derive}
a contradiction.
\[
\AxiomC{$\Discharge{\lforall[x][!A(x)]}{1}$}
\DeduceC{$\lfalse$}
\DischargeRule{\Intro{\lnot}}{1}
\UnaryInfC{$\lnot\lforall[x][!A(x)]$}
\DisplayProof
\]
So far so good. We can use \Elim{\lforall} but it's not obvious
if that will help us get to our goal. Instead, let's use one of our 
assumptions. $\lforall[x][!A(x)] \lif \lexists[y][!B(y)]$ together
with $\lforall[x][!A(x)]$ will allow us to use the \Elim{\lif} rule.
\[
\AxiomC{$\lforall[x][!A(x)] \lif \lexists[y][!B(y)]$}
\AxiomC{$\Discharge{\lforall[x][!A(x)]}{1}$}
\RightLabel{\Elim{\lif}}
\BinaryInfC{$\lexists[y][!B(y)]$}
\DeduceC{$\lfalse$}
\DischargeRule{\Intro{\lnot}}{1}
\UnaryInfC{$\lnot\lforall[x][!A(x)]$}
\DisplayProof
\]
We now have one final assumption to work with,
and it looks like this will help us reach a contradiction
by using \Elim{\lnot}.
\[
\AxiomC{$\lnot\lexists[y][!B(y)]$}
\AxiomC{$\lforall[x][!A(x)] \lif \lexists[y][!B(y)]$}
\AxiomC{$\Discharge{\lforall[x][!A(x)]}{1}$}
\RightLabel{\Elim{\lif}}
\BinaryInfC{$\lexists[y][!B(y)]$}
\RightLabel{\Elim{\lnot}}
\BinaryInfC{$\lfalse$}
\DischargeRule{\Intro{\lnot}}{1}
\UnaryInfC{$\lnot\lforall[x][!A(x)]$}
\DisplayProof
\]
\end{ex}

\end{document}
