% Part: first-order-logic
% Chapter: natural-deduction
% Section: rules-and-proofs

\documentclass[../../../include/open-logic-section]{subfiles}

\begin{document}

\olfileid{fol}{ntd}{rul}

\olsection{Rules and \usetoken{P}{derivation}}

\begin{editorial}
  This section collects all the rules propositional connectives and
  quantifiers, but not for identity.  It is planned to divide this
  into separate sections on connectives and quantifiers so that proofs
  for propositional logic can be treated separately
  (\gitissue{77}).
\end{editorial}

\begin{explain}
Natural deduction systems are meant to closely parallel the informal
reasoning used in mathematical proof (hence it is somewhat
``natural''). Natural deduction proofs begin with assumptions.
Inference rules are then applied. Assumptions are ``!!{discharged}''
by the \Intro{\lnot}, \Intro{\lif}, \Elim{\lor} and \Elim{\lexists}
inference rules, and the label of the !!{discharged} assumption is
placed beside the inference for clarity.
\end{explain}

\begin{defn}[Initial !!^{formula}]
An \emph{initial !!{formula}} or \emph{assumption} is any !!{formula}
in the topmost position of any branch.
\end{defn}

\begin{defn}[Inference]
An \emph{inference} is an expression of the form
\[
\AxiomC{$!A$}
\UnaryInfC{$!C$}
\DisplayProof
\quad
\textrm{  or  }
\quad
\AxiomC{$!A$}
\AxiomC{$!B$}
\BinaryInfC{$!C$}
\DisplayProof
\]
where $!A, !B$, and $!C$ are !!{formula}s. $!A$ and $!B$ are called
the \emph{upper !!{formula}s} or \emph{premises} and $!C$ the
\emph{lower !!{formula}s} or \emph{conclusion} of the inference.
\end{defn}

The rules for natural deduction are divided into two main types:
\emph{propositional} rules (quantifier-free) and \emph{quantifier}
rules.  The rules come in pairs, an introduction and an elimination
rule for each !!{operator}. They introduced !!a{operator} in the
conclusion or remove !!a{operator} from a premise of the rule.  Some
of the rules allow an assumption of a certain type to be
\emph{!!{discharged}}. To indicate which assumption is !!{discharged}
by which inference, we also assign labels to both the assumption and
the inference.  This is indicated by writing the assumption as
``$\Discharge{!A}{n}$''.

It is customary to consider rules for all !!{operator}s, even for
those (if any) that we consider as defined.

\subsection{Propositional Rules}

\subsubsection{Rules for $\land$}

\begin{defish}
\AxiomC{$!A$}
\AxiomC{$!B$}
\RightLabel{\Intro{\land}}
\BinaryInfC{$!A \land !B$}
\DisplayProof
\hfill
\begin{tabular}{r}
\AxiomC{$!A \land !B$}
\RightLabel{\Elim{\land}}
\UnaryInfC{$!A$}
\DisplayProof
\\[3ex]
\AxiomC{$!A \land !B$}
\RightLabel{\Elim{\land}}
\UnaryInfC{$!B$}
\DisplayProof
\end{tabular}
\end{defish}

\subsubsection{Rules for $\lor$}

\begin{defish}
\begin{tabular}{r}
\AxiomC{$!A$}
\RightLabel{\Intro{\lor}}
\UnaryInfC{$!A \lor !B$}
\DisplayProof
\\[3ex]
\AxiomC{$!B$}
\RightLabel{\Intro{\lor}}
\UnaryInfC{$!A \lor !B$}
\DisplayProof
\end{tabular}
\hfill
\AxiomC{$!A \lor !B$}
\AxiomC{$\Discharge{!A}{n}$}
\DeduceC{$!C$}
\AxiomC{$\Discharge{!B}{n}$}
\DeduceC{$!C$}
\DischargeRule{\Elim{\lor}}{n}
\TrinaryInfC{$!C$}
\DisplayProof
\end{defish}

\subsubsection{Rules for $\lif$}

\begin{defish}
\AxiomC{$\Discharge{!A}{n}$}
\DeduceC{$!B$}
\DischargeRule{\Intro{\lif}}{n}
\UnaryInfC{$!A \lif !B$}
\DisplayProof
\hfill
\AxiomC{$!A \lif !B$}
\AxiomC{$!A$}
\RightLabel{\Elim{\lif}}
\BinaryInfC{$!B$}
\DisplayProof
\end{defish}

\subsubsection{Rules for $\lnot$}

\begin{defish}
\AxiomC{$\Discharge{!A}{n}$}
\noLine
\DeduceC{$\lfalse$}
\DischargeRule{\Intro{\lnot}}{n}
\UnaryInfC{$\lnot !A$}
\DisplayProof
\hfill
\AxiomC{$\lnot !A$}
\AxiomC{$!A$}
\RightLabel{\Elim{\lnot}}
\BinaryInfC{$\lfalse$}
\DisplayProof
\end{defish}

\subsubsection{Rules for $\lfalse$}

\begin{defish}
\AxiomC{$\lfalse$}
\RightLabel{\FalseInt}
\UnaryInfC{$!A$}
\DisplayProof
\hfill
\AxiomC{$\Discharge{\lnot !A}{n}$}
\DeduceC{$\lfalse$}
\DischargeRule{\FalseCl}{n}
\UnaryInfC{$!A$}
\DisplayProof
\end{defish}

\subsection{Quantifier Rules}

\subsubsection{Rules for $\lforall$}

\begin{defish}
\AxiomC{$!A(a)$}
\RightLabel{\Intro{\lforall}}
\UnaryInfC{$\lforall[x][!A(x)]$}
\DisplayProof
\hfill
\AxiomC{$\lforall[x][!A(x)]$}
\RightLabel{\Elim{\lforall}}
\UnaryInfC{$!A(t)$}
\DisplayProof
\end{defish}

In the rules for~$\lforall$, $t$ is a ground term (a term that does
not contain any variables), and $a$~is a !!{constant} which does not
occur in~$!A$, or in any assumption which is !!{undischarged} in the
!!{derivation} ending with the premise~$!A$. We call $a$ the
\emph{eigenvariable} of the \Intro{\lforall} inference.

\subsubsection{Rules for $\lexists$}

\begin{defish}
\AxiomC{$\Atom{!A}{t}$}
\RightLabel{\Intro{\lexists}}
\UnaryInfC{$\lexists[x][\Atom{!A}{x}]$}
\DisplayProof
\hfill
\AxiomC{$\lexists[x][\Atom{!A}{x}]$}
\AxiomC{[$\Atom{!A}{a}$]$^n$}
\DeduceC{$!C$}
\DischargeRule{\Elim{\lexists}}{n}
\BinaryInfC{$!C$}
\DisplayProof
\end{defish}

Again, $t$ is a ground term, and $a$ is a constant which does not occur
in the premise $\lexists[x][!A(x)]$, in $!C$, or any assumption which
is !!{undischarged} in the !!{derivation}s ending with the two
premises~$!C$ (other than the assumptions $!A(a)$).  We call $a$ the
\emph{eigenvariable} of the \Elim{\lexists} inference.

The condition that an eigenvariable neither occur in the premises nor in
any assumption that is !!{undischarged} in the !!{derivation}s leading 
to the premises for the \Intro{\lforall} or \Elim{\lexists} inference is 
called the \emph{eigenvariable condition}.

\begin{explain}
We use the term ``eigenvariable'' even though $a$ in the above rules
is a constant. This has historical reasons.

In \Intro{\lexists} and \Elim{\lforall} there are no restrictions, and
the term~$t$ can be anything, so we do not have to worry about any
conditions. On the other hand, in the \Elim{\lexists} and \Intro{\lforall}
rules, the eigenvariable condition requires that $a$ does not
occur anywhere else in the formula. Thus, if the upper !!{formula} is
valid, the truth values of the formulas other than $\Atom{!A}{a}$ are
independent of~$a$.
\end{explain}

\subsection{!!^{derivation}s}

\begin{defn}[!!^{derivation}]
A \emph{!!{derivation}} of !!a{formula} $!A$ from assumptions~$\Gamma$
is a tree of !!{formula}s satisfying the following conditions:
\begin{enumerate}
\item The topmost !!{formula}s of the tree are either in $\Gamma$ or
  are !!{discharged} by an inference in the tree.
\item Every !!{formula} in the tree is an upper !!{formula} of an
  inference whose lower !!{formula} stands directly below that !!{formula} in
  the tree.
\end{enumerate}
We then say that $!A$ is the \emph{end-!!{formula}} of the
!!{derivation} and that $!A$ is \emph{!!{derivable}} from~$\Gamma$.
\end{defn}

\end{document}
