% Part: computability
% Chapter: recursive-functions
% Section: sequences

\documentclass[../../include/open-logic-section]{subfiles}

\begin{document}

\olfileid{cmp}{rec}{seq}
\olsection{Sequences}

The set of primitive recursive functions is
remarkably robust. But we will be able to do even more once we have
developed an adequate means of handling \emph{sequences}.  We will
identify finite sequences of natural numbers with natural numbers, in
the following way: the sequence $\langle a_0, a_1, a_2, \dots, a_k \rangle$
corresponds to the number
\[
p_0^{a_0+1} \cdot p_1^{a_1+1} \cdot p_2^{a_2+1} \cdot \dots \cdot
p_k^{a_k+1}.
\]
We add one to the exponents to guarantee that, for example,
the sequences $\langle 2, 7, 3\rangle$ and $\langle 2, 7, 3, 0, 0 \rangle$ have
distinct numeric codes. We will take both 0 and 1 to code the empty
sequence; for concreteness, let $\emptyseq$ denote~$0$.

Let us define the following functions:
\begin{enumerate}
\item $\len{s}$, which returns the length of the sequence $s$:
\[
\len{s} = 
\begin{cases} 
0 & \text{if $s = 0$ or $s = 1$} \\
\fn{min} \; i < s \; (p_i | s \land \lforall[j < s][(j > i \lif
    p_j \not | s)])+1 & \text{otherwise}
\end{cases}
\]
Note that we need to bound the search on $i$; clearly $s$ provides an
acceptable bound.

\item $\fn{append}(s,a)$, which returns the result of appending $a$ to
  the sequence $s$:
\[
\fn{append}(s,a) = 
\begin{cases}
  2^{a+1} & \text{if $s = 0$ or $s = 1$} \\
s \cdot p_{\len{s}}^{a+1} & \text{otherwise}
\end{cases}
\]
\item $\fn{element}(s,i)$, which returns the $i$th element of $s$
  (where the initial element is called the $0$th), or $0$ if $i$ is
  greater than or equal to the length of $s$:
\[
\fn{element}(s,i) = 
\begin{cases}
  0 & \mbox{if $i \geq \len{s}$} \\
  \fn{min} \; {j < s} \; (p_i^{j+2} \not | s) - 1 & \text{otherwise}
\end{cases}
\]
\end{enumerate}

Instead of using the official names for the functions defined above,
we introduce a more compact notation.  We will use $(s)_i$ instead of
$\fn{element}(s,i)$, and $\tuple{s_0, \dots, s_k}$ to abbreviate
$\fn{append}(\fn{append}(\dots \fn{append}(\emptyseq,s_0)
\dots),s_k)$. Note that if $s$ has length~$k$, the elements of $s$ are
$(s)_0$, \dots,~$(s)_{k-1}$.

It will be useful for us to be able to bound the numeric code of a
sequence, in terms of its length and its largest element. Suppose $s$
is a sequence of length $k$, each element of which is less than equal
to some number $x$. Then $s$ has at most $k$ prime factors, each at
most $p_{k-1}$, and each raised to at most $x+1$ in the prime
factorization of $s$. In other words, if we define
\[
\fn{sequenceBound}(x,k) = p_{k-1}^{k \cdot (x+1)},
\]
then the numeric code of the sequence~$s$ described above, is at
most~$\fn{sequenceBound}(x,k)$.

Having such a bound on sequences gives us a way of defining new
functions using bounded search. For example, suppose we want to
define the function $\fn{concat}(s,t)$, which concatenates two
sequences. One first option is to define a ``helper'' function
$\fn{hconcat}(s,t,n)$ which concatenates the first $n$ symbols of $t$
to~$s$. This function can be defined by primitive recursion, as
follows:
\begin{enumerate}
\item $\fn{hconcat}(s,t,0) = s$
\item $\fn{hconcat}(s,t,n+1) = \fn{append}(\fn{hconcat}(s,t,n),(t)_n)$
\end{enumerate}
Then we can define $\fn{concat}$ by 
\[
\fn{concat}(s,t) = \fn{hconcat}(s,t,\len{t}).
\]
But using bounded search, we can be lazy. All we need to do is write
down a primitive recursive \emph{specification} of the object (number)
we are looking for, and a bound on how far to look. The following works:
\begin{multline*}
  \fn{concat}(s,t) = \fn{min} \; {v < \fn{sequenceBound}(s+t,\len{s} +
    \len{t})} \\
  (\len{v} = \len{s} + \len{t} \land \\
  \lforall[i < \len{s}][((v)_i = (s)_i) \land \lforall[j <
    \len{t}][((v)_{\len{s}+j} = (t)_j)])]
\end{multline*}
We will write $s \concat t$ instead of $\fn{concat}(s,t)$.

\begin{prob}
Show that there is a primitive recursive function~$\fn{sconcat}(s)$
whith the property that
\[
\fn{sconcat}(\tuple{s_0, \dots, s_k}) = s_0 \concat \dots \dots \concat s_k.
\]
\end{prob}

\end{document}

