% Part: normal-modal-logic
% Chapter: filtrations
% Section: filtrations-def

\documentclass[../../../include/open-logic-section]{subfiles}

\begin{document}

\olfileid{nml}{fil}{fil}

\olsection{Filtrations}

Rather than define ``the'' filtration of $\mModel{M}$ through
$\Gamma$, we define when a model~$\mModel{M^*}$ counts as a filtration
of~$\mModel{M}$. All filtrations have the same set of worlds~$W^*$ and
the same valuation~$V^*$. But different filtrations may have different
accessibility relations~$R^*$. To count as a filtration, $R^*$ has to
satisfy a number of conditions, however. These conditions are exactly
what we'll require to prove the main result, namely that
$\mSat{M}{!A}[w]$ iff $\mSat{M^*}{!A}[{[w]}]$, provided $!A \in
\Gamma$.

\begin{defn}\ollabel{defn:filtration}
  Let $\Gamma$ be closed under subformulas and $\mModel{M} = \tuple{W,
    R, V}$. A \emph{filtration of $\mModel{M}$ through $\Gamma$} is
  any model $\mModel{M^*} = \tuple{W^*,R^*,V^*}$, where:
  \begin{enumerate}
  \item $W^* = \Setabs{[w]}{w \in W}$;
  \item \ollabel{defn:filtration-R}%
    For any $u,v \in W$:
    \begin{enumerate}
    \item \ollabel{defn:filtration-R1}%
      If $Ruv$ then $R^*[u][v]$;
    \tagitem{prvBox}{\ollabel{defn:filtration-R2}%
      If $R^*[u][v]$ then for any $\Box !A \in \Gamma$, if
      $\mSat{M}{\Box!A}[u]$ then $\mSat{M}{!A}[v]$}{};
    \tagitem{prvDiamond}{\ollabel{defn:filtration-R3}%
      If $R^*[u][v]$ then for any $\Diamond !A \in \Gamma$, if
      $\mSat{M}{!A}[v]$ then $\mSat{M}{\Diamond!A}[u]$.}{}
    \end{enumerate}
  \item $V^*(p) = \Setabs{[u]}{u \in V(p)}$.
  \end{enumerate}
\end{defn}

It's worthwhile thinking about what $V^*(p)$ is: the set consisting of
the equivalence classes~$[w]$ of all worlds~$w$ where $p$ is true
in~$\mModel{M}$. On the one hand, if $w \in V(p)$, then $[w] \in
V^*(p)$ by that definition. However, it is not necessarily the case
that if $[w] \in V^*(p)$, then $w \in V(p)$. If $[w] \in V^*(p)$ we
are only guaranteed that $[w] = [u]$ for \emph{some}~$u \in V(p)$. Of
course, $[w] = [u]$ means that $w \equiv u$.  So, when $[w] \in
V^*(p)$ we can (only) conclude that $w \equiv u$ for some $u \in
V(p)$.

\begin{thm}\ollabel{thm:filtrations}
  If $\mModel{M^*}$ is a filtration of $\mModel{M}$ through $\Gamma$,
  then for every $!A \in \Gamma$ and $w \in W$, we have
  $\mSat{M}{!A}[w]$ if and only if $\mSat{M^*}{!A}[{[w]}]$.
\end{thm}

\begin{proof}
  By induction on $!A$, using the fact that $\Gamma$ is closed under
  subformulas. Since $!A \in \Gamma$ and $\Gamma$ is closed under
  sub-!!{formula}s, all sub-!!{formula}s of~$!A$ are also $\in
  \Gamma$. Hence in each inductive step, the induction hypothesis
  applies to the sub-!!{formula}s of~$!A$.
  \begin{enumerate}
    \tagitem{prvFalse}{\indcase{!A}{\lfalse}{Neither
        $\mSat{M}{\indfrm}[w]$ nor $\mSat{M^*}{\indfrm}[{[w]}]$.}}{}
    \tagitem{prvTrue}{\indcase{!A}{\ltrue}{Both
        $\mSat{M}{\indfrm}[w]$ and $\mSat{M^*}{\indfrm}[{[w]}]$.}}{}
    \item \indcase{!A}{p}{The left-to-right direction is immediate, as
      $\mSat{M}{\indfrm}[w]$ only if $w \in V(p)$, which implies $[w]
      \in V^*(p)$, i.e., $\mSat{M^*}{\indfrm}[{[w]}]$. Conversely,
      suppose $\mSat{M^*}{\indfrm}[{[w]}]$, i.e., $[w] \in
      V^*(p)$. Then for some $v \in V(p)$, $w \equiv v$. Of course
      then also $\mSat{M}{p}[v]$. Since $w \equiv v$, $w$ and $v$ make
      the same !!{formula}s from~$\Gamma$ true. Since by assumption $p
      \in \Gamma$ and $\mSat{M}{p}[v]$, $\mSat{M}{\indfrm}[w]$.}

    \tagitem{prvNot}{\iftag{probNot}{Exercise.}{\indcase{!A}{\lnot
          !B}{$\mSat{M}{\indfrm}[w]$ iff $\mSat/{M}{!B}[w]$. By
          induction hypothesis, $\mSat/{M}{!B}[w]$ iff
          $\mSat/{M^*}{!B}[{[w]}]$.  Finally, $\mSat/{M^*}{!B}[{[w]}]$
          iff $\mSat{M^*}{\indfrm}[{[w]}]$.}}}{}
        
    \tagitem{prvAnd}{\iftag{probAnd}{Exercise.}{\indcase{!A}{(!B \land
          !C)}{$\mSat{M}{\indfrm}[w]$ iff $\mSat{M}{!B}[w]$ and
          $\mSat{M}{!C}[w]$. By induction hypothesis, $\mSat{M}{!B}[w]$
          iff $\mSat{M^*}{!B}[{[w]}]$, and $\mSat{M}{!C}[w]$ iff
          $\mSat{M^*}{!C}[{[w]}]$. And $\mSat{M^*}{\indfrm}[{[w]}]$
          iff $\mSat{M^*}{!B}[{[w]}]$ and
          $\mSat{M^*}{!C}[{[w]}]$.}}}{}

    \tagitem{prvOr}{\iftag{probOr}{Exercise.}{\indcase{!A}{(!B \lor
          !C)}{$\mSat{M}{\indfrm}[w]$ iff $\mSat{M}{!B}[w]$ or
          $\mSat{M}{!C}[w]$. By induction hypothesis, $\mSat{M}{!B}[w]$
          iff $\mSat{M^*}{!B}[{[w]}]$, and $\mSat{M}{!C}[w]$ iff
          $\mSat{M^*}{!C}[{[w]}]$. And $\mSat{M^*}{\indfrm}[{[w]}]$
          iff $\mSat{M^*}{!B}[{[w]}]$ or
          $\mSat{M^*}{!C}[{[w]}]$.}}}{}

    \tagitem{prvIf}{\iftag{probIf}{Exercise.}{\indcase{!A}{(!B \lif
          !C)}{$\mSat{M}{\indfrm}[w]$ iff $\mSat/{M}{!B}[w]$ or
          $\mSat{M}{!C}[w]$. By induction hypothesis, $\mSat{M}{!B}[w]$
          iff $\mSat{M^*}{!B}[{[w]}]$, and $\mSat{M}{!C}[w]$ iff
          $\mSat{M^*}{!C}[{[w]}]$. And $\mSat{M^*}{\indfrm}[{[w]}]$
          iff $\mSat/{M^*}{!B}[{[w]}]$ or
          $\mSat{M^*}{!C}[{[w]}]$.}}}{}

    \tagitem{prvIff}{\iftag{probIff}{Exercise.}{\indcase{!A}{(!B \liff
          !C)}{$\mSat{M}{\indfrm}[w]$ iff $\mSat{M}{!B}[w]$ and
          $\mSat{M}{!C}[w]$, or $\mSat/{M}{!B}[w]$ and
          $\mSat/{M}{!C}[w]$. By induction hypothesis, $\mSat{M}{!B}[w]$
          iff $\mSat{M^*}{!B}[{[w]}]$, and $\mSat{M}{!C}[w]$ iff
          $\mSat{M^*}{!C}[{[w]}]$. And $\mSat{M^*}{\indfrm}[{[w]}]$
          iff $\mSat{M^*}{!B}[{[w]}]$ and
          $\mSat{M^*}{!C}[{[w]}]$, or $\mSat/{M^*}{!B}[{[w]}]$ and
          $\mSat/{M^*}{!C}[{[w]}]$.}}}{}

    \tagitem{prvBox}{\iftag{probBox}{Exercise.}{\indcase{!A}{\Box
          !B}{Suppose $\mSat{M}{\indfrm}[w]$; to show that
          $\mSat{M^*}{\indfrm}[{[w]}]$, let $v$ be such that
          $R^*[w][v]$. From
          \olref{defn:filtration}\olref{defn:filtration-R2}, we have
          that $\mSat{M}{!B}[v]$, and by inductive hypothesis
          $\mSat{M^*}{!B}[{[v]}]$. Since $v$ was arbitrary,
          $\mSat{M^*}{\indfrm}[{[w]}]$ follows.

          Conversely, suppose $\mSat{M^*}{\indfrm}[{[w]}]$ and let $v$
          be arbitrary such that $Rwv$. From
          \olref{defn:filtration}\olref{defn:filtration-R1}, we have
          $R^*[w][v]$, so that $\mSat{M^*}{!B}[{[v]}]$; by inductive
          hypothesis $\mSat{M}{!B}[v]$, and since $v$ was arbitrary,
          $\mSat{M}{\indfrm}[w]$.}}}{}
    
    \tagitem{prvDiamond}{\iftag{probDiamond}{Exercise.}{\indcase{!A}{\Diamond
          !B}{Suppose $\mSat{M}{\indfrm}[w]$. Then for some $v \in W$,
          $Rwv$ and $\mSat{M}{!B}[v]$.  By inductive hypothesis
          $\mSat{M^*}{!B}[{[v]}]$, and by
          \olref{defn:filtration}\olref{defn:filtration-R1}, we have
          $R^*[w][v]$. Thus, $\mSat{M^*}{\indfrm}[{[w]}]$.

          Now suppose $\mSat{M^*}{\indfrm}[{[w]}]$. Then for some $[v]
          \in W^*$ with $R^*[w][v]$, $\mSat{M^*}{!B}[{[v]}]$.  By
          inductive hypothesis $\mSat{M}{!B}[v]$.  By
          \olref{defn:filtration}\olref{defn:filtration-R3}, we have
          that $\mSat{M}{\indfrm}[w]$.}}}{}
  \end{enumerate}
\end{proof}

\begin{prob}
  Complete the proof of \olref[nml][fil][fil]{thm:filtrations}
\end{prob}

What holds for truth at worlds in a model also holds for truth in a
model and validity in a class of models.

\begin{cor}
  Let $\Gamma$ be closed under subformulas. Then:
  \begin{enumerate}
  \item If $\mModel{M^*}$ is a filtration of $\mModel{M}$ through
    $\Gamma$ then for any $!A \in \Gamma$: $\mSat{M}{!A}$ if and only
    if $\mSat{M^*}{!A}$.
  \item If $\mClass{C}$ is a class of models and $\Gamma(\mClass{C})$
    is the class of $\Gamma$-filtrations of models in $\mClass{C}$,
    then any !!{formula}~$!A \in \Gamma$ is valid in $\mClass{C}$ if
    and only if it is valid in $\Gamma(\mClass{C})$.
  \end{enumerate}
\end{cor}



\end{document}
