% Part: normal-modal-logic
% Chapter: filtrations
% Section: more-filtrations

\documentclass[../../../include/open-logic-section]{subfiles}

\begin{document}

\olfileid{nml}{fil}{acc}

\olsection{Filtrations and Properties of Accessibility}

As noted, filtrations of universal, serial, and reflexive models are
always also universal, serial, or reflexive. But not every filtration
of a symmetric or transitive model is symmetric or transitive,
respectively. In some cases, however, it is possible to define
filtrations so that this does hold. In order to do so, we proceed as
in the definition of the coarsest filtration, but add additional
conditions to the definition of~$R^*$.  Let $\Gamma$ be closed under
sub-!!{formula}s. Consider the relations~$C_i(u,v)$ in
\olref{tab:Cn-filtrations} between worlds $u$, $v$ in a model
$\mModel{M} =\tuple{W, R, V}$. We can define $R^*[u][v]$ on the basis
of combinations of these conditions. For instance, if we stipulate
that $R^*[u][v]$ iff the condition $C_1(u,v)$ holds, we get exactly
the coarsest filtration. If we stipulate $R^*[u][v]$ iff both
$C_1(u,v)$ and $C_2(u, v)$ hold, we get a different filtration. It is
``finer'' than the coarsest since fewer pairs of worlds satisfy
$C_1(u,v)$ and $C_2(u,v)$ than $C_1(u,v)$ alone.

\begin{table}[ht]
  \centering
  \begin{tabular}{|ll|}
    \hline
    \multirow{2}{*}{$C_1(u,v)$:}
    \iftag{prvBox}{&
      if $\Box!A \in \Gamma$ and $\mSat{M}{\Box!A}[u]$
      then $\mSat{M}{!A}[v]$;\iftag{prvDiamond}{ and}{}\\}{}
    \iftag{prvDiamond}{&
      if $\Diamond!A \in \Gamma$  and $\mSat{M}{!A}[v]$
      then $\mSat{M}{\Diamond!A}[u]$; \\}{}
    \hline
    \multirow{2}{*}{$C_2(u,v)$:} 
    \iftag{prvBox}{&
      if $\Box!A \in \Gamma$ and $\mSat{M}{\Box!A}[v]$
      then $\mSat{M}{!A}[u]$;\iftag{prvDiamond}{ and}{}\\}{}
    \iftag{prvDiamond}{&
      if $\Diamond!A \in \Gamma$ and $\mSat{M}{!A}[u]$
      then $\mSat{M}{\Diamond!A}[v]$; \\}{}
    \hline
    \multirow{2}{*}{$C_3(u,v)$:}
    \iftag{prvBox}{&
      if $\Box!A \in \Gamma$ and $\mSat{M}{\Box!A}[u]$
      then $\mSat{M}{\Box!A}[v]$;\iftag{prvDiamond}{ and}{}\\}{}
    \iftag{prvDiamond}{&
      if $\Diamond!A \in \Gamma$ and $\mSat{M}{\Diamond!A}[v]$
      then $\mSat{M}{\Diamond!A}[u]$; \\}{}
    \hline
    \multirow{2}{*}{$C_4(u,v)$:}
    \iftag{prvBox}{&
      if $\Box!A \in \Gamma$ and $\mSat{M}{\Box!A}[v]$
      then $\mSat{M}{\Box!A}[u]$;\iftag{prvDiamond}{ and}{}\\}{}
    \iftag{prvDiamond}{&
      if $\Diamond!A \in \Gamma$ and $\mSat{M}{\Diamond!A}[u]$
      then $\mSat{M}{\Diamond!A}[v]$; \\}{}
    \hline
    \end{tabular}
  \caption{Conditions on possible worlds for defining
    filtrations.}
  \ollabel{tab:Cn-filtrations}
\end{table}

\begin{thm}\ollabel{thm:more-filtrations}
  Let $\mModel{M} =\tuple{W,R,V}$ be a model, $\Gamma$ closed under
  sub-!!{formula}s. Let $W^*$ and $V^*$ be defined as in
  \olref[fil]{defn:filtration}. Then:
  \begin{enumerate}
  \item Suppose $R^*[u][v]$ if and only if $C_1(u, v) \land
    C_2(u,v)$. Then $R^*$ is symmetric, and $\mModel{M^*} =
    \tuple{W^*,R^*,V^*}$ is a filtration if $\mModel{M}$ is symmetric.
  \item Suppose $R^*[u][v]$ if and only if $C_1(u, v)
    \land C_3(u,v)$. Then $R^*$ is transitive, and
    $\mModel{M^*}=\tuple{W^*,R^*,V^*}$ is a filtration if $\mModel{M}$
    is transitive.
  \item Suppose $R^*[u][v]$ if and only if $C_1(u, v) \land C_2(u,v)
    \land C_3(u,v) \land C_4(u,v)$. Then $R^*$ is symmetric and
    transitive, and $\mModel{M^*}=\tuple{W^*,R^*,V^*}$ is a filtration
    if $\mModel{M}$ is symmetric and transitive.
  \item Suppose $R^*$ is defined as $R^*[u][v]$ if and only if $C_1(u,
    v) \land C_3(u,v) \land C_4(u,v)$. Then $R^*$ is transitive and
    euclidean, and $\mModel{M^*}=\tuple{W^*,R^*,V^*}$ is a filtration
    if $\mModel{M}$ is transitive and euclidean.
  \end{enumerate}
\end{thm}

\begin{proof}
  \begin{enumerate}
    \item It's immediate that $R^*$ is symmetric, since $C_1(u,v)
      \Leftrightarrow C_2(v,u)$ and $C_2(u,v) \Leftrightarrow
      C_1(v,u)$. So it's left to show that if $\mModel{M}$ is
      symmetric then $\mModel{M^*}$ is a filtration
      through~$\Gamma$. Condition $C_1(u,v)$ guarantees that
      \iftag{prvBox}{%
        \iftag{prvDiamond}{%
          \olref[fil]{defn:filtration-R2} and
          \olref[fil]{defn:filtration-R3} of
          \olref[fil]{defn:filtration} are}{%
          \olref[fil]{defn:filtration}\olref[fil]{defn:filtration-R2} is}}{%
        \olref[fil]{defn:filtration}\olref[fil]{defn:filtration-R3} is}
      satisfied. So we just have to verify
      \olref[fil]{defn:filtration}\olref[fil]{defn:filtration-R1},
      i.e., that $Ruv$ implies $R^*[u][v]$.

      So suppose $Ruv$. To show $R^*[u][v]$ we need to establish that
      $C_1(u,v)$ and $C_2(u,v)$. For $C_1$: \iftag{prvBox}{if $\Box!A
        \in\Gamma$ and $\mSat{M}{\Box!A}[u]$ then also
        $\mSat{M}{!A}[v]$ (since $Ruv$).\iftag{prvDiamond}{ Similarly,
        }{}}{}\iftag{prvDiamond}{if $\Diamond!A \in \Gamma$ and
        $\mSat{M}{!A}[v]$ then $\mSat{M}{\Diamond!A}[u]$ since
        $Ruv$.}{} For $C_2$: \iftag{prvBox}{if $\Box!A \in \Gamma$ and
        $\mSat{M}{\Box!A}[v]$ then $Ruv$ implies $Rvu$ by symmetry, so
        that $\mSat{M}{!A}[u]$.\iftag{prvDiamond}{ Similarly,
        }{}}{}\iftag{prvDiamond}{if $\Diamond!A \in\Gamma$ and
        $\mSat{M}{!A}[u]$ then $\mSat{M}{\Diamond!A}[v]$ (since $Rvu$
        by symmetry).}{}
    \item Exercise.
    \item Exercise.
    \item Exercise.
  \end{enumerate}
\end{proof}

\begin{prob}
  Complete the proof of \olref[nml][fil][acc]{thm:more-filtrations}.
\end{prob}

\end{document}
