% Part: normal-modal-logic
% Chapter: filtrations
% Section: S5-fmp

\documentclass[../../../include/open-logic-section]{subfiles}

\begin{document}

\olfileid{nml}{fil}{fmp}

\olsection{\Log{K} and \Log{S5} have the Finite Model Property}

\begin{defn}
  A system $\Sigma$ of modal logic is said to have the \emph{finite
    model property} if whenever !!a{formula}~$!A$ is true at a world
  in a model of $\Sigma$ then $!A$ is true at a world in a
  \emph{finite} model of~$\Sigma$.
\end{defn}

\begin{prop}\ollabel{prop:K-fmp}
  \Log{K} has the finite model property.
\end{prop}

\begin{proof}
  \Log{K} is the set of valid !!{formula}s, i.e., any model is a model
  of~\Log{K}.  By \olref[fil]{thm:filtrations}, if
  $\mSat{M}{!A}[w]$, then $\mSat{M^*}{!A}[w]$ for any filtration
  of~$\mModel{M}$ through the set $\Gamma$ of sub-!!{formula}s
  of~$!A$. Any !!{formula} only has finitely many sub-!!{formula}s, so
  $\Gamma$ is finite. By \olref[fin]{prop:filt-are-finite},
  $\card{W^*} \le 2^n$, where $n$ is the number of !!{formula}s
  in~$\Gamma$. And since \Log{K} imposes no restriction on models,
  $\mModel{M^*}$ is a \Log{K}-model.
\end{proof}

To show that a logic~\Log{L} has the finite model property via
filtrations it is essential that the filtration of an \Log{L}-model is
itself a \Log{L}-model. Often this requires a fair bit of work, and
not any filtration yields a \Log{L}-model. However, for universal
models, this still holds.

\begin{prop}\ollabel{prop:univ-fin}
  Let $\mClass{U}$ be the class of universal models (see
  \olref[frd][es5]{prop:S5=univ}) and $\mClass{U}_\mathrm{Fin}$ the
  class of all finite universal models. Then any !!{formula}~$!A$ is
  valid in $\mClass{U}$ if and only if it is valid in
  $\mClass{U}_\mathrm{Fin}$.
\end{prop}

\begin{proof}
  Finite universal models are universal models, so the left-to-right
  direction is trivial. For the right-to left direction, suppose
  that~$!A$ is false at some world $w$ in a universal model
  $\mModel{M}$. Let $\Gamma$ contain $!A$ as well as all of its
  subformulas; clearly $\Gamma$ is finite. Take a filtration
  $\mModel{M^*}$ of $\mModel{M}$; then $\mModel{M^*}$ is finite by
  \olref[fin]{prop:filt-are-finite}, and by
  \olref[fil]{thm:filtrations}, $!A$ is false at $[w]$ in
  $\mModel{M^*}$. It remains to observe that $\mModel{M^*}$ is also
  universal: given $u$ and $v$, by hypothesis $Ruv$ and by
  \olref[fil]{defn:filtration}\olref[fil]{defn:filtration-R},
  also $R^*[u][v]$.
\end{proof}

\begin{cor}\ollabel{cor:S5fmp}
  \Log{S5} has the finite model property.
\end{cor}

\begin{proof}
  By \olref[frd][es5]{prop:S5=univ}, if $!A$ is true at a world in
  some reflexive and euclidean model then it is true at a world in a
  universal model. By \olref{prop:univ-fin}, it is true at a world in
  a finite universal model (namely the filtration of the model through
  the set of sub-!!{formula}s of~$!A$). Every universal model is also
  reflexive and euclidean; so $!A$ is true at a world in a finite
  reflexive euclidean model.
\end{proof}

\begin{prob}
  Show that any filtration of a serial or reflexive model is also
  serial or reflexive (respectively).
\end{prob}

\begin{prob}
  Find a non-symmetric (non-transitive, non-euclidean) filtration of a
  symmetric (transitive, euclidean) model.
\end{prob}

\end{document}
