\chapter*{About the Open Logic Project}

The \textit{Open Logic Text} is an open-source, collaborative textbook
of formal meta-logic and formal methods, starting at an intermediate level
(i.e., after an introductory formal logic course). It is aimed at a
non-mathematical audience (in particular, students of philosophy and
computer science), but is completely rigorous.

The \textit{Open Logic Text} is a collaborative project, and is under
active development.  Coverage of some topics currently included may
not yet be complete, and many sections still require substantial
revisions. We plan to expand the text to cover additional topics in
the future. We also plan to add features to the text, such as a
glossary, further reading, historical notes, pictures, better
explanations, sections explaining the relevance of results to
philosophy, computer science, and mathematics, and more problems and
examples.  If you find an error or have a suggestion,
\href{https://github.com/OpenLogicProject/OpenLogic/wiki/Contributing}{please
  let the project team know}.

The project operates in the spirit of open source. Not only is the
texts available for free, we provide it in source LaTeX format under a
Creative Commons Attribution license, which gives anyone the right to
download, use, modify, re-arrange, convert, and re-distribute our
work, as long as they give appropriate credit.

Please see the Open Logic Project website at
\href{http://openlogicproject.org/}{openlogicproject.org} for
additional information.
