% Part: first-order-logic
% Chapter: syntax-and-semantics
% Section: free-vars-sentences

\documentclass[../../include/open-logic-section]{subfiles}

\begin{document}

\olfileid{fol}{syn}{fvs}

\olsection{Free Variables and Sentences}

\begin{defn}[Free occurrences of a !!{variable}]
The \emph{free} occurrences of a !!{variable} in a !!{formula} are defined
inductively:
\begin{enumerate}
\item If $!A$ is atomic, all !!{variable} occurrences in $!A$ are free.

\tagitem{prvNot}{If $!A$ is a !!{formula}, then the free !!{variable}
  occurrences of $\lnot !A$ are exactly those of $!A$.}{}

\tagitem{prvAnd}{If $!A$ and $!B$ are !p{formula}, then the free
  !!{variable} occurrences in $(!A \land !B)$ are those in $!A$
  together with those in $!B$.}{}

\tagitem{prvOr}{If $!A$ and $!B$ are !p{formula}, then the free
  !!{variable} occurrences in $(!A \lor !B)$ are those in $!A$
  together with those in $!B$.}

\tagitem{prvIf}{If $!A$ and $!B$ are !p{formula}, then the free
  !!{variable} occurrences in $(!A \lif !B)$ are those in $!A$
  together with those in $!B$.}{}

\tagitem{prvIff}{If $!A$ and $!B$ are !p{formula}, then the free
  !!{variable} occurrences in $(!A \liff !B)$ are those in $!A$
  together with those in $!B$.}{}

\tagitem{prvAll}{If $!A$ is a !!{formula}, then the free !!{variable}
  occurrences in $\lforall[x][!A]$ are all of those in~$!A$ except for
  occurrences of~$x$.}{}

\tagitem{prvEx}{If $!A$ is a !!{formula}, then the free !!{variable}
  occurrences in $\lexists[x][!A]$ are all of those in~$!A$ except for
  occurrences of~$x$.}{}
\end{enumerate}
\end{defn}

\begin{defn}[Sentence]
A !!{formula} $!A$ is a \emph{sentence} iff it contains no free
occurrences of !p{variable}.
\end{defn}

\begin{explain}
Induction is a commonly-used proof technique of mathematics in logic. 
You may be familiar with it from math, in which you do induction on 
natural numbers. As it turns out, there are lots of different kinds of 
mathematical and logical sets of objects that lend themselves to induction.

In general, what induction allows one to do is prove a universal claim; that 
is, show that every object of a certain kind has some property. In particular, 
induction is often useful where a ``universal introduction'' method of proof 
is too complicated. Induction only works on matehmatical objects that are 
constructed in special ways: if every element in the set is either basic or is 
built up from basic elements, then we can use induction on it. More precisely:
\end{explain}

\begin{defn}
A set $S$ is \emph{closed} under a function $f$ iff whenever $x \in
S$, then $f(x) \in S$.
\end{defn}

\begin{explain}
For example, the natural numbers ($\mathbb{N}$) are closed under the successor 
function $f(x) = x+1$. If $x$ is a natural number, so is $x+1$. The natural numbers 
are not, however, closed under the predecessor function $f(x) = x-1$, since $0$ is a 
natural number, but $-1$ is not.
\end{explain}

\begin{defn}
A property $P$ is said to be \emph{preserved under} a function $f(x)$ when, for any
object $a$ in the domain of $f$, $P(a)$ implies $P(f(a))$ (if $a$ has property $P$, then 
so does $f(a)$).
\end{defn}

\begin{explain}
The property of ``being an even number'' is preserved under the function $f(x) = x+2$, 
since if $x$ is even, so is $x+2$. The property of being an even number is not preserved 
under the successor function, since if $x$ is even, $x+1$ is odd.

Induction works on natural numbers because every natural number can be 
reached from 0 by applying the successor function a finite number of times. This is 
characteristic of any set upon which we can do induction.
\end{explain}

\begin{defn}[Induction on $\mathbb{N}$]
If 0 has some property $P$, and if $P$ is preserved under the successor function, then
every natural number has property $P$.
\end{defn}

\begin{ex} The sum of the first $n$ natural numbers is $\frac{n(n+1)}{2}$. \\

\textbf{Base case.} The sum of the first 0 natural numbers is $0= \frac{0\cdot 1}{2}$.\\

\textbf{Inductive step.} Suppose the property holds for $k$. We will show that it holds 
for $k+1$. In other words, suppose that $P(k)$ holds, that is,
\[ 0 + 1 + \ldots + k = \frac{k(k+1)}{2} \]
We will show that $P(k+1)$ holds, i.e. that
\[ 0 + 1 + \ldots + k + (k+1) = \frac{(k+1)(k+2)}{2} \]
by using the assumption that $P(k)$.

\begin{align*}
0 + 1 + \ldots + k &= \frac{k(k+1)}{2} \tag{Assumption} \\
0 + 1 + \ldots + k + (k+1) &= \frac{k(k+1)}{2} + (k+1) \tag{Add $k+1$ to both sides} \\
&= \frac{k(k+1) + 2(k+1)}{2} \\
&= \frac{2k + k + 2k +2}{2} \\
&=\frac{(k+1)(k+2)}{2}
\end{align*}
Hence, the inductive step goes through, and we have proved the claim.
\end{ex}

\begin{explain}
For !p{formula}, the basic elements are the atomic !p{formula}. More complex !p{formula} 
are obtained by joining together less-complicated !p{formula} using operators. Each 
operator can be seen as corresponding to a function of !p{formula} that returns a new 
!!{formula} joining these two together with the operator in question. For example, the function 
that joins two !p{formula} $!A$ and $!B$ into a conjunction can be written as $\mathcal E _\land 
(!A, !B) = !A \land !B$. We can call these \emph{formula-building functions}. The set of 
!p{formula} is closed these functions: whenever $!A$ and $!B$ are !p{formula}, so are 
$\lnot !A$, $!A \land !B$, etc. 
\end{explain}

\begin{defn}[Induction on !p{formula}]
If every atomic !!{formula} has property $P$ and $P$ is preserved under the !!{formula}-building 
functions, then every !!{formula} has property $P$.
\end{defn}

\begin{explain}
This definition suggests a ``recipe'' for inductive proofs on !p{formula}. If we are asked 
to show that every !!{formula} has property $P$, follow the following steps:\\

\textbf{Base Case.} Let $!A$ be an atomic !!{formula}. [\ldots] Therefore, $!A$ has 
property $P$.\\

\textbf{Inductive Step.} Let $!A$ and $!B$ be !p{formula}, both of which have the property $P$.

Case $\lnot$: [\ldots] Therefore, $\lnot !A$ has property $P$.

Case $\land$: [\ldots] Therefore, $!A \land !B$ has property $P$. 

Case $\lor$: [\ldots] Therefore, $!A \lor !B$ has property $P$. 

Case $\lif$: [\ldots] Therefore, $!A \lif !B$ has property $P$.

Case $\liff$: [\ldots] Therefore, $!A \liff !B$ has property $P$.

Case $\lforall[]$: [\ldots] Therefore, $\lforall[x] !A$ has property $P$.

Case $\lexists[]$: [\ldots] Therefore, $\lexists[x] !A$ has property $P$. \\

Therefore, every !!{formula} has property $P$.
\end{explain}

\begin{defn} 
A term $t$ is \emph{free for $x$ in $!A$} if $x$ does not occur in
$!A$ within the scope of a quantifier $\lforall[z]$ binding a !!{variable}
$z$ occurring in~$t$.
\end{defn}

%Examples of this maybe?


\end{document}
