\documentclass[modal-logic]{subfiles}

\begin{document}

\section{Modal Validity}

Aside from truth in a model~$\Mod M$ at a world $w$, we may consider various  types of \emph{validity} of formulas.  In general, notions of validity are defined by universally quantifying over (components of) models.  The first notion, validity relative to a model, quantifies over all worlds:

\begin{defn}\label{defn:valid-in-model}
A formula~$!A$ is \emph{valid in a model}~$\Mod M = \langle W, R, V\rangle$, $\Mod M \Mmodels !A$, iff $\Mod M, w \Mmodels !A$ for all $w \in W$.
\end{defn}

\begin{ex}
\dots
\end{ex}

So $\Mod M \Mmodels !A$ if $!A$ is true everywhere in~$\Mod M$. If we quantify over all valuations as well, we get the notion of validity in a frame.

\begin{defn}\label{valid-in-frame}
A formula~$!A$ if \emph{valid in a frame}~$\Mod F = \langle W, R\rangle$,
$\Mod F \Mmodels !A$, iff $\Mod M \Mmodels !A$ for every model~$\Mod M = \langle W, R, V\rangle$ based on~$\Mod F$.
\end{defn}

The formulas K and Dual,
\begin{align}
\tag{K} \Box(p \lif q) & \lif (\Box p \lif \Box q) \\
\tag{Dual} \Diamond p & \liff \lnot\Box\lnot p
\end{align}
are true in every model at every world, i.e., they are valid in every frame.  This is the reason why we can use relational models only to deal with normal modal logics. 

\begin{prop}
$\Mod F \Mmodels \mathrm{K}$ for all frames~$\Mod F$.
\end{prop}

\begin{proof}
Assume $\Mod M$ is a model based on $\Mod F$, and that $\Mod M, w \Mmodels \Box(p \lif q)$. We have to show that $\Mod M, w \Mmodels \Box p \lif \Box q$, i.e., that $\Mod M, w \Mmodels \Box q$ provided also $\Mod M, w \Mmodels \Box p$.  By \ref{defn:mmodels}\ref{defn:sub:mmodels-box}, $\Mod M, w \Mmodels \Box q$ iff $\Mod M, v \Mmodels q$ for all $v \in W$ with~$wRv$. From our assumptions we have, however, that $\Mod M, v \Mmodels p \lif q$ and $\Mod M, v \Mmodels p$, and this implies $\Mod M, v \Mmodels q$. 
\end{proof}

\begin{prob}
Show that $\Mod F \Mmodels \mathrm{Dual}$ for all frames~$\Mod F$.
\end{prob}

\begin{prob}
Show that $\Mod F \Mmodels \Box(p \land q) \lif (\Box p \land \Box q)$ for all frames~$\Mod F$.
\end{prob}

\begin{ex}
\dots
\end{ex}

There is an intermediate notion where we quantify over all valuations, but consider validity relative to a frame and world in it.  This notion is important since it can be used to express local properties of the relation~$R$ by formulas.

\begin{defn}
A \formula~$!A$ is \emph{valid in a frame~$\Mod F = \langle W, R\rangle$ at a world~$w \in W$} iff $\Mod M, w \Mmodels !A$ for every model~$\Mod M = \langle W, R, V\rangle$ based on~$\Mod F$.
\end{defn}

\begin{prop}
$\Mod F, w \Mmodels \Box p \lif p$ iff $wRw$.
\end{prop}

\begin{proof}
If: Suppose $wRw$, and let $\Mod M$ be a model based on~$\Mod F$.  We have to show that if $\Mod M, w \Mmodels \Box p$, then $\Mod M, w \Mmodels p$.  As $\Mod M, w \Mmodels \Box p$, we have $\Mod M, v \Mmodels p$ for all $v \in W$ with $wRv$. As $wRw$, $w$ itself is one of these~$v$, and so $\Mod M, w \Mmodels p$.

Only if: Suppose $\Mod F, w \Mmodels \Box p \lif p$, but not $wRw$.  We define a model based on~$\Mod F$ at where $\Box p \lif p$ is false at~$w$.  To this end, let~$V(p, w) = 0$ and $V(p, v) = 1$ for all $v \in W$ with $wRv$.  As not $wRw$ by assumption, $w$ is not among these~$v$ and the definitions do not conflict. Let $\Mod M$ be the model based on $\Mod F$ with this valuation.  Then $\Mod M, w \Mmodels \Box p$ but $\Mod M, w \nMmodels p$, and so $\Mod M, w \nMmodels \Box p \lif p$.
\end{proof}



\end{document}
