% Part:sets-functions-relations
% Chapter: sets
% Section: enumerability

\documentclass[../../../include/open-logic-section]{subfiles}

\begin{document}

\olfileid{sfr}{siz}{enm}

\olsection{\printtoken{S}{enumerable} Sets}

One way of specifying a finite set is by listing its !!{element}s. But
conversely, since there are only finitely many !!{element}s in a set,
every finite set can be enumerated.  By this we mean: its elements can
be put into a list (a list with a beginning, where each !!{element} of
the list other than the first has a unique predecessor).  Some
infinite sets can also be enumerated, such as the set of positive
integers.

\begin{defn}[Enumeration]
Informally, an \emph{enumeration} of a set~$X$ is a list (possibly
infinite) of !!{element}s of~$X$ such that every !!{element} of $X$
appears on the list at some finite position. If $X$ has an
enumeration, then $X$ is said to be \emph{!!{enumerable}}.  If $X$ is
!!{enumerable} and infinite, we say $X$ is !!{denumerable}.
\end{defn}

\begin{explain}
A couple of points about enumerations:
\begin{enumerate}
\item We count as enumerations only lists which have a beginning and
  in which every !!{element} other than the first has a single
  !!{element} immediately preceding it.  In other words, there are
  only finitely many elements between the first !!{element} of the
  list and any other !!{element}. In particular, this means that every
  !!{element} of an enumeration has a finite position: the first
  !!{element} has position~$1$, the second position~$2$, etc.
\item We can have different enumerations of the same set~$X$ which
  differ by the order in which the !!{element}s appear: $4$, $1$,
  $25$, $16$,~$9$ enumerates the (set of the) first five square
  numbers just as well as $1$, $4$, $9$, $16$,~$25$ does.
\item Redundant enumerations are still enumerations: $1$, $1$, $2$,
  $2$, $3$, $3$,~\dots{} enumerates the same set as $1$, $2$,
  $3$,~\dots{} does.
\item Order and redundancy \emph{do} matter when we specify an
  enumeration: we can enumerate the positive integers beginning with
  $1$, $2$, $3$, $1$, \dots{}, but the pattern is easier to see when
  enumerated in the standard way as $1$, $2$, $3$, $4$,~\dots
\item Enumerations must have a beginning: \dots, $3$, $2$, $1$ is not
  an enumeration of the natural numbers because it has no first
  !!{element}. To see how this follows from the informal definition,
  ask yourself, ``at what position in the list does the number 76
  appear?''
\item The following is not an enumeration of the positive integers:
  $1$, $3$, $5$, \dots, $2$, $4$, $6$, \dots\@ The problem is that the
  even numbers occur at places $\infty + 1$, $\infty + 2$, $\infty +
  3$, rather than at finite positions.
\item Lists may be gappy: $2$, $-$, $4$, $-$, $6$, $-$, \dots{}
  enumerates the even positive integers.
\item The empty set is enumerable: it is enumerated by the empty list!{}
\end{enumerate}
\end{explain}

\begin{prop}
  If $X$ has an enumeration, it has an enumeration without gaps or
  repetitions.
\end{prop}

\begin{proof}
  Suppose $X$ has an enumeration $x_1$, $x_2$, \dots{} in which each
  $x_i$ is an !!{element} of~$X$ or a gap.  We can remove repetitions
  from an enumeration by replacing repeated !!{element}s by gaps. For
  instance, we can turn the enumeration into a new one in which $x_i'$
  is $x_i$ if $x_i$ is !!a{element} of~$X$ that is not among $x_1$,
  \dots, $x_{i-1}$ or is $-$ if it is. We can remove gaps by closing
  up the elements in the list. To make precise what ``closing up''
  amounts to is a bit difficult to describe. Roughly, it means that we
  can generate a new enumeration $x_1''$, $x_2''$, \dots, where each
  $x_i''$ is the first !!{element} in the enumeration $x_1'$, $x_2'$,
  \dots{} after $x_{i-1}''$ (if there is one).
\end{proof}

The last argument shows that in order to get a good handle on
enumerations and !!{enumerable} sets and to prove things about them,
we need a more precise definition.  The following provides it.

\begin{defn}[Enumeration]
An \emph{enumeration} of a set $X$ is any !!{surjective} function
$f \colon \Int^+ \to X$.
\end{defn}

\begin{explain}
Let's convince ourselves that the formal definition and the informal
definition using a possibly gappy, possibly infinite list are
equivalent. !!^a{surjective} function (partial or total) from $\Int^+$
to a set $X$ enumerates~$X$. Such a function determines an enumeration
as defined informally above: the list $f(1)$, $f(2)$, $f(3)$,
\dots. Since $f$ is !!{surjective}, every !!{element} of $X$ is
guaranteed to be the value of $f(n)$ for some~$n \in \Int^+$. Hence,
every !!{element} of $X$ appears at some finite position in the
list. Since the function may not be !!{injective}, the list may be
redundant, but that is acceptable (as noted above).

On the other hand, given a list that enumerates all !!{element}s
of~$X$, we can define !!a{surjective} function $f\colon \Int^+ \to X$ by
letting $f(n)$ be the $n$th !!{element} of the list that is not a gap,
or the final !!{element} of the list if there is no $n$th !!{element}.
There is one case in which this does not produce !!a{surjective}
function: if $X$ is empty, and hence the list is empty. So, every
non-empty list determines !!a{surjective} function $f\colon \Int^+ \to
X$.
\end{explain}

\begin{defn}
  \ollabel{defn:enumerable}
  A set~$X$ is !!{enumerable} iff it is empty or has an enumeration.
\end{defn}

\begin{ex}
A function enumerating the positive integers ($\Int^+$) is simply the
identity function given by $f(n) = n$. A function enumerating the
natural numbers $\Nat$ is the function $g(n) = n - 1$.
\end{ex}

\begin{prob}
  According to \olref[sfr][siz][enm]{defn:enumerable}, a set $X$ is
  enumerable iff $X = \emptyset$ or there is !!a{surjective} $f\colon
  \Int^+ \to X$.  It is also possible to define ``!!{enumerable} set''
  precisely by: a set is enumerable iff there is !!a{injective}
  function $g\colon X \to \Int^+$.  Show that the definitions are
  equivalent, i.e., show that there is !!a{injective} function
  $g\colon X \to \Int^+$ iff either $X = \emptyset$ or there is
  !!a{surjective} $f\colon \Int^+ \to X$.
\end{prob}

\begin{ex}
The functions $f\colon \Int^+ \to \Int^+$ and $g \colon \Int^+ \to
\Int^+$ given by
\begin{align*}
f(n) & = 2n \text{ and}\\
g(n) & = 2n+1
\end{align*}
enumerate the even positive integers and the odd positive integers,
respectively. However, neither function is an enumeration of
$\Int^+$, since neither is !!{surjective}.
\end{ex}

\begin{prob}
Define an enumeration of the positive squares $4$, $9$, $16$, \dots
\end{prob}

\begin{ex}
The function $f(n) = (-1)^{n} \lceil \frac{(n-1)}{2}\rceil$ (where
$\lceil x \rceil$ denotes the \emph{ceiling} function, which rounds
$x$ up to the nearest integer) enumerates the set of
integers~$\Int$. Notice how $f$ generates the values of $\Int$ by
``hopping'' back and forth between positive and negative integers:
\[
\begin{array}{c c c c c c c c}
f(1) & f(2) & f(3) & f(4) & f(5) & f(6) & f(7) & \dots \\ \\
- \lceil \tfrac{0}{2} \rceil & \lceil \tfrac{1}{2}\rceil & - \lceil \tfrac{2}{2} \rceil & \lceil \tfrac{3}{2} \rceil & - \lceil \tfrac{4}{2} \rceil  & \lceil \tfrac{5}{2}
\rceil & - \lceil \tfrac{6}{2} \rceil & \dots \\ \\
0 & 1 & -1 & 2 & -2 & 3 & \dots
\end{array}
\]
You can also think of $f$ as defined by cases as follows:
\[
f(n) = \begin{cases}
  0 & \text{if $n = 1$}\\
  n/2 & \text{if $n$ is even}\\
  -(n-1)/2 & \text{if $n$ is odd and $>1$}
  \end{cases}
\]
\end{ex}

\begin{prob}
Show that if $X$ and $Y$ are !!{enumerable}, so is $X \cup Y$.
\end{prob}

\begin{prob}
  Show by induction on $n$ that if $X_1$, $X_2$, \dots, $X_n$ are all
  !!{enumerable}, so is $X_1 \cup \dots \cup X_n$.
\end{prob}

\begin{explain}
That is fine for ``easy'' sets. What about the set of, say, pairs of
natural numbers?{}
\[
\Int^+ \times \Int^+ = \Setabs{\tuple{n,m}}{n,m \in \Int^+}
\]
We can organize the pairs of positive integers 
in an \emph{array}, such as the following:
\[
\begin{array}{ c | c | c | c | c | c}
& \textbf 1 & \textbf 2 & \textbf 3 & \textbf 4 & \dots \\
\hline
\textbf 1 & \tuple{1,1} & \tuple{1,2} & \tuple{1,3} & \tuple{1,4} & \dots \\
\hline
\textbf 2 & \tuple{2,1} & \tuple{2,2} & \tuple{2,3} & \tuple{2,4} & \dots \\
\hline
\textbf 3 & \tuple{3,1} & \tuple{3,2} & \tuple{3,3} & \tuple{3,4} & \dots \\
\hline
\textbf 4 & \tuple{4,1} & \tuple{4,2} & \tuple{4,3} & \tuple{4,4} & \dots \\
\hline
\vdots & \vdots & \vdots & \vdots & \vdots & \ddots\\
\end{array}
\]

Clearly, every ordered pair in $\Int^+ \times \Int^+$ will appear
exactly once in the array. In particular, $\tuple{n,m}$ will appear in
the $n$th column and $m$th row. But how do we organize the elements of
such an array into a one-way list? The pattern in the array below
demonstrates one way to do this:
\[
\begin{array}{ c | c | c | c | c | c }
& & & & & \\
\hline
& 1 & 2 & 4 & 7 & \dots \\
\hline
& 3 & 5 & 8 & \dots & \dots \\
\hline
& 6 & 9 & \dots & \dots & \dots \\
\hline
& 10 & \dots & \dots & \dots & \dots \\
\hline
& \vdots & \vdots & \vdots & \vdots & \ddots\\
\end{array}
\]
This pattern is called \emph{Cantor's zig-zag method}. Other patterns
are perfectly permissible, as long as they ``zig-zag'' through every
cell of the array. By Cantor's zig-zag method, the enumeration for
$\Int^+ \times \Int^+$ according to this scheme would be:
\[
\tuple{1,1}, \tuple{1,2}, \tuple{2,1}, \tuple{1,3}, \tuple{2,2},
\tuple{3,1}, \tuple{1,4}, \tuple{2,3}, \tuple{3,2}, \tuple{4,1}, \dots
\]

What ought we do about enumerating, say, the set of ordered triples
of positive integers?
\[
\Int^+ \times \Int^+ \times \Int^+ = \Setabs{\tuple{n,m,k}}{n,m,k \in \Int^+}
\]
We can think of $\Int^+ \times \Int^+ \times \Int^+$ as the Cartesian
product of $\Int^+ \times \Int^+$ and $\Int^+$, that is,
\[
(\Int^+)^3 = (\Int^+ \times \Int^+) \times \Int^+ =
\Setabs{\tuple{\tuple{n,m},k}}{\tuple{n,m} \in \Int^+ \times \Int^+, k
  \in \Int^+ }
\]
and thus we can enumerate $(\Int^+)^3$ with an array by labelling one
axis with the enumeration of $\Int^+$, and the other axis with the
enumeration of $(\Int^+)^2$:
\[
\begin{array}{ c | c | c | c | c | c}
& \textbf 1 & \textbf 2 & \textbf 3 & \textbf 4 & \dots \\
\hline
\mathbf{\tuple{1,1}} & \tuple{1,1,1} & \tuple{1,1,2} & \tuple{1,1,3} & \tuple{1,1,4} & \dots \\
\hline
\mathbf{\tuple{1,2}} & \tuple{1,2,1} & \tuple{1,2,2} & \tuple{1,2,3} & \tuple{1,2,4} & \dots \\
\hline
\mathbf{\tuple{2,1}} & \tuple{2,1,1} & \tuple{2,1,2} & \tuple{2,1,3} & \tuple{2,1,4} & \dots \\
\hline
\mathbf{\tuple{1,3}} & \tuple{1,3,1} & \tuple{1,3,2} & \tuple{1,3,3} & \tuple{1,3,4} & \dots\\
\hline
\vdots & \vdots & \vdots & \vdots & \vdots & \ddots \\
\end{array}
\]
Thus, by using a method like Cantor's zig-zag method, we may
similarly obtain an enumeration of~$(\Int^+)^3$.
\end{explain}

\begin{prob}
Give an enumeration of the set of all positive rational numbers. (A
positive rational number is one that can be written as a fraction
$n/m$ with $n, m \in \Int^+$).
\end{prob}

\begin{prob}
Show that $\Rat$ is !!{enumerable}. (A rational number is one that can
be written as a fraction $z/m$ with $z \in \Int$, $m \in \Int^+$).
\end{prob}

\begin{prob}
Define an enumeration of $\Bin^*$.
\end{prob}

\begin{prob}
Recall from your introductory logic course that each possible truth
table expresses a truth function. In other words, the truth functions
are all functions from $\Bin^k \to \Bin$ for some~$k$. Prove that the
set of all truth functions is enumerable.
\end{prob}

\begin{prob}
Show that the set of all finite subsets of an arbitrary infinite
enumerable set is enumerable.
\end{prob}

\begin{prob}
A set of positive integers is said to be \emph{cofinite} iff
it is the complement of a finite set of positive integers. Let
\emph{I} be the set that contains all the finite and cofinite sets of
positive integers. Show that \emph{I} is enumerable.
\end{prob}

\begin{prob}
Show that the !!{enumerable} union of !!{enumerable} sets is
!!{enumerable}. That is, whenever $X_1$, $X_2$, \dots{} are sets, and
each $X_i$ is !!{enumerable}, then the union $\bigcup_{i=1}^\infty
X_i$ of all of them is also !!{enumerable}.
\end{prob}

\end{document}
