% Part: first-order-logic
% Chapter: axiomatic-deduction
% Section: soundness

\documentclass[../../include/open-logic-section]{subfiles}

\begin{document}

\olfileid{fol}{axd}{sou}
\olsection{Soundness}

\begin{explain}
A !!{derivation} system, such as axiomatic deduction, is \emph{sound}
if it cannot !!{derive} things that do not actually hold.  Soundness is
thus a kind of guaranteed safety property for !!{derivation} systems.
Depending on which proof theoretic property is in question, we would
like to know for instance, that
\begin{enumerate}
\item every !!{derivable} !!{sentence} is valid;
\item if a !!{sentence} is !!{derivable} from some others, it is also a
  consequence of them;
\item if a set of !!{sentence}s is inconsistent, it is unsatisfiable.
\end{enumerate}
These are important properties of a !!{derivation} system.  If any of them do
not hold, the !!{derivation} system is deficient---it would !!{derive} too much.
Consequently, establishing the soundness of a !!{derivation} system is of the
utmost importance.
\end{explain}

\begin{prop}
If $!A$ is an axiom, then $\Sat{M}{!A}[s]$ for each !!{structure}
$\Struct{M}$ and assignment~$s$.
\end{prop}

\begin{proof}
We first verify that the schemas \textbf{Ax0}--\textbf{Ax5} are
valid. For instance, here is the case for \textbf{Ax1}: suppose $t$ is
!!{free for} $x$ in $!A$, and assume $\Sat{M}{\lforall[x][!A]}[s]$. Then
by definition of satisfaction, for each $a \in \Domain{M}$, also
$\Sat{M}{!A}[\Subst{s}{a}{x}]$, and in particular this holds when $a =
\Value{t}{M}[s]$, i.e., $\Sat{M}{!A}[\Subst{s}{\Value{t}{M}[s]}{x}]$. By
the Substitution Theorem, $\Sat{M}{\Subst{!A}{t}{x}}[s]$. This shows
that $\Sat{M}{(\lforall[x][!A] \lif \Subst{!A}{t}{x})}[s]$. After
verifying the schemas, we see that their universal closures are also
valid, by~\ref{cor:valid}.
\end{proof}

\begin{thm}[Soundness]
\ollabel{thm:Soundness}
If $\Gamma \Proves !A$ then $\Gamma \Entails !A$.
\end{thm}

\begin{proof}
By induction on theorems. By the previous proposition, all the axioms
are valid, and hence if $!A$ is an axiom then $\Gamma \Entails
!A$. Similarly if $!A \in \Gamma$. And if $\Gamma \Proves !B$ and
$\Gamma \Proves !B \lif !A$ then $\Gamma \Entails !A$.
\end{proof}

\end{document}
