\documentclass[../../../include/open-logic-section]{subfiles}

\begin{document}

\olfileid{sth}{ordinals}{wo}
\olsection{Well-Orderings}

The fundamental notion is as follows:

\begin{defn}
The relation $<$ \emph{well-orders} $A$ iff it meets these two conditions:
\begin{enumerate}
	\item $<$ is connected, i.e., for all $a, b \in A$, either $a < b$
	or $a = b$ or $b < a$;
	\item every non-empty subset of $A$ has a $<$-minimal !!{element},
	i.e., if $\emptyset \neq X \subseteq A$ then $(\exists m \in
	X)(\forall z \in X)z \nless m$
\end{enumerate}
\end{defn}

It is easy to see that three examples we just considered were indeed
well-ordering relations. 

\begin{prob}
\Olref[sth][ordinals][idea]{sec} presented three example orderings
on the natural numbers. Check that each is a well-ordering.
\end{prob}

Here are some elementary but extremely important observations
concerning well-ordering.

\begin{prop}\ollabel{wo:strictorder}
If $<$ well-orders $A$, then every non-empty subset of $A$ has a unique
$<$-least member, and $<$ is irreflexive, asymmetric and transitive.
\end{prop}

\begin{proof}
If $X$ is a non-empty subset of $A$, it has a $<$-minimal !!{element}
$m$, i.e., $(\forall z \in X)z \nless m$. Since $<$ is connected,
$(\forall z \in X)m \leq z$. So $m$ is the $<$-least !!{element} of $X$.

For irreflexivity, fix $a \in A$; the $<$-least !!{element} of $\{a\}$ is $a$, so $a \nless a$. For transitivity, if $a < b < c$, then
since $\{a, b, c\}$ has a $<$-least !!{element}, $a < c$. Asymmetry
follows from irreflexivity and transitivity
\end{proof}

\begin{prop}\ollabel{propwoinduction}
If $<$ well-orders $A$, then for any formula $\phi(x)$:
\[
	\text{if }(\forall a \in A)((\forall b < a)\phi(b) \lif 
		\phi(a))\text{, then }(\forall a \in A)\phi(a).
\]
\end{prop}

\begin{proof}
We will prove the contrapositive. Suppose $\lnot(\forall a \in
A)\phi(a)$, i.e., that $X = \Setabs{x \in A}{\lnot\phi(x)} \neq
\emptyset$. Then $X$ has an $<$-minimal !!{element}, $a$. So $(\forall
b < a)\phi(b)$ but $\lnot \phi(a)$. \end{proof}\noindent This last
property should remind you of the principle of strong induction on the
naturals, i.e.: if $(\forall n \in \omega)((\forall m < n)\phi(m)
\lif \phi(n))$, then $(\forall n \in \omega)\phi(n)$. And this
property makes well-ordering into a very \emph{robust} notion.\footnote{A reminder: all formulas can have parameters (unless explicitly stated otherwise).} 

\end{document}