% Part: computability
% Chapter: recursive-functions
% Section: primitive-recursion

\documentclass[../../../include/open-logic-section]{subfiles}

\begin{document}

\olfileid{cmp}{rec}{pre}
\olsection{Primitive Recursion}

\begin{explain}
Suppose we specify that a certain function $l$ from $\Nat$ to $\Nat$
satisfies the following two clauses:
\begin{eqnarray*}
l(0) & = & 1 \\
l(x+1) & = & 2 \cdot l(x).
\end{eqnarray*}
It is pretty clear that there is only one function, $l$, that meets
these two criteria. This is an instance of a \emph{definition by
  primitive recursion}. We can define even more fundamental functions
like addition and multiplication by
\begin{eqnarray*}
f(x,0) & = & x \\
f(x,y+1) & = & f(x,y)+1
\end{eqnarray*}
and
\begin{eqnarray*}
g(x,0) & = & 0 \\
g(x,y+1) & = & f(g(x,y),x).
\end{eqnarray*}
Exponentiation can also be defined recursively, by
\begin{eqnarray*}
h(x,0) & = & 1 \\
h(x,y+1) & = & g(h(x,y),x).
\end{eqnarray*}
We can also compose functions to build more complex ones; for example,
\begin{eqnarray*}
k(x) & = & x^x + (x + 3) \cdot x \\
& = & f(h(x,x),g(f(x,3),x)).
\end{eqnarray*}

Remember that the \emph{arity} of a function is the number of
arguments. Let $\Zero(x)$ be the function that always returns~$0$,
regardless of what $x$~is, and let $\Succ(x) = x+1$ be the successor
function.  The set of \emph{primitive recursive functions} is the set
of functions from $\Nat^n$ to $\Nat$ that you get if you start with
$\Zero$ and $\Succ$ by iterating the two operations above, primitive
recursion and composition. The idea is that primitive recursive
functions are defined in a straightforward and explicit way, so that
it is intuitively clear that each one can be computed using finite
means.
\end{explain}

\begin{defn}
If $f$ is a $k$-ary function and $g_0$, \dots,~$g_{k-1}$ are $l$-ary
functions on the natural numbers, the \emph{composition} of $f$ with
$g_0$, \dots,~$g_{k-1}$ is the $l$-ary function $h$ defined by
\[
h(x_0,\dots,x_{l-1}) =
f(g_0(x_0,\dots,x_{l-1}),\dots,g_{k-1}(x_0,\dots,x_{l-1})) .
\]
\end{defn}

\begin{defn}
If $f$ is a $k$-ary function and
$g$ is a $(k+2)$-ary function, then the function
defined by \emph{primitive recursion from $f$ and $g$} is the
$(k+1)$-ary function~$h$ defined by the equations
\begin{eqnarray*}
h(0,z_0,\dots,z_{k-1}) & = & f(z_0,\dots,z_{k-1}) \\
h(x+1,z_0,\dots,z_{k-1}) & = & g(x,h(x,z_0,\dots,z_{k-1}),z_0,\dots,z_{k-1})
\end{eqnarray*}
\end{defn}

In addition to $\Zero$ and $\Succ$, we will include among primitive
recursive functions the projection functions,
\[
\Proj{n}{i}(x_0,\dots,x_{n-1}) = x_i,
\]
for each natural number $n$ and $i < n$. In the end, we have the
following:

\begin{defn}
  The set of primitive recursive functions is the set of functions of
  various arities from $\Nat^n$ to $\Nat$, defined inductively by the
  following clauses:
\begin{enumerate}
\item $\Zero$ is primitive recursive.
\item $\Succ$ is primitive recursive.
\item Each projection function $\Proj{n}{i}$ is primitive recursive.
\item If $f$ is a $k$-ary primitive recursive function and
  $g_0$, \dots,~$g_{k-1}$ are $l$-ary primitive recursive functions, then
  the composition of $f$ with $g_0$, \dots,~$g_{k-1}$ is primitive
  recursive.
\item If $f$ is a $k$-ary primitive recursive function and $g$ is a
  $k+2$-ary primitive recursive function, then the function defined by
  primitive recursion from $f$ and $g$ is primitive recursive.
\end{enumerate}
\end{defn}

\begin{explain}
Put more concisely, the set of primitive recursive functions is the
smallest set containing $\Zero$, $\Succ$, and the projection
functions~$\Proj{n}{j}$, and which is closed under composition and
primitive recursion.

Another way of describing the set of primitive recursive functions
keeps track of the ``stage'' at which a function enters the set. Let
$S_0$ denote the set of starting functions: $\Zero$, $\Succ$, and the
projections. Once $S_i$ has been defined, let $S_{i+1}$ be the set of
all functions you get by applying a single instance of composition or
primitive recursion to functions in $S_i$. Then
\[
S = \bigcup_{i \in \Nat} S_i
\]
is the set of all primitive recursive functions
\end{explain}

Our definition of composition may seem too rigid, since
$g_0$, \dots,~$g_{k-1}$ are all required to have the same arity, $l$. But
adding the projection functions provides the desired flexibility. For
example, suppose $f$ and~$g$ are ternary functions and $h$~is the
binary function defined by
\[
h(x,y) = f(x,g(x,x,y),y).
\]
The definition of~$h$ can be rewritten with the projection
functions, as
\[
h(x,y) = f(\Proj{2}{0}(x,y), g(\Proj{2}{0}(x,y), \Proj{2}{0}(x,y),
\Proj{2}{1}(x,y)), \Proj{2}{1}(x,y)).
\]
Then $h$ is the composition of $f$ with $\Proj{2}{0}$, $l$, and
$\Proj{2}{1}$, where
\[
l(x,y) = g(\Proj{2}{0}(x,y),\Proj{2}{0}(x,y),\Proj{2}{1}(x,y)),
\]
i.e., $l$ is the composition of $g$ with $\Proj{2}{0}$, $\Proj{2}{0}$,
and~$\Proj{2}{1}$.

For another example, let us again consider addition. This is
described recursively by the following two equations:
\begin{eqnarray*}
x + 0 & = & x \\
x + (y+1) & = & \Succ(x+y).
\end{eqnarray*}
In other words, addition is the function $\Add$ defined recursively by
the equations
\begin{eqnarray*}
\Add(0,x) & = & x \\
\Add(y+1,x) & = & \Succ(\Add(y,x)).
\end{eqnarray*}
But even this is not a strict primitive recursive definition; we need
to put it in the form
\begin{eqnarray*}
\Add(0,x) & = & f(x) \\
\Add(y+1,x) & = & g(y,\Add(y,x),x)
\end{eqnarray*}
for some 1-ary primitive recursive function~$f$ and some 3-ary
primitive recursive function~$g$. We can take $f$ to be $\Proj{1}{0}$, and
we can define $g$ using composition,
\[
g(y,w,x) = \Succ(\Proj{3}{1}(y,w,x)).
\]
The function $g$, being the composition of basic primitive recursive
functions, is primitive recursive; and hence so is~$h$. (Note that,
strictly speaking, we have defined the function $g(y, x)$ meeting the
recursive specification of $x + y$; in other words, the variables are
in a different order. Luckily, addition is commutative, so here the
difference is not important; otherwise, we could define the
function~$g'$ by
\[
g'(x,y) = g(\Proj{2}{1}(y,x)),\Proj{2}{0}(y,x)) = g(y,x),
\]
using composition.


\begin{explain}
One advantage to having the precise description of the primitive
recursive functions is that we can be systematic in describing them.
For example, we can assign a ``notation'' to each such function, as
follows. Use symbols $\Zero$, $\Succ$, and $\Proj{n}{i}$ for zero,
successor, and the projections. Now suppose $f$ is defined by
composition from a $k$-ary function~$h$ and $l$-ary functions $g_0$,
\dots,~$g_{k-1}$, and we have assigned notations $H$, $G_0$,
\dots,~$G_{k-1}$ to the latter functions. Then, using a new symbol
$\fn{Comp}_{k,l}$, we can denote the function $f$ by
$\fn{Comp}_{k,l}[H,G_0,\dots,G_{k-1}]$. For the functions defined by
primitive recursion, we can use analogous notations of the form
$\fn{Rec}_k[G,H]$, where $k$ denotes that arity of the function being
defined. With this setup, we can denote the addition function by
\[
\fn{Rec}_2[\Proj{1}{0},\fn{Comp}_{1,3}[\Succ,\Proj{3}{1}]].
\]
Having these notations sometimes proves useful.
\end{explain}

\begin{prob}
Multiplication satisfies the recursive equations
\begin{align*}
  0 \cdot y & = y\\
  (x+1) \cdot y & = (x \cdot y) + x
\end{align*}
Give the explicit precise definition of the function $\Mult(x, y) =
x\cdot y$, assuming that $\Add(x, y) = x + y$ is already defined. Give
the complete notation for $\Mult$.
\end{prob}

\end{document}
