% Part: incompleteness
% Chapter: representability-in-q
% Section: beta-function

\documentclass[../../include/open-logic-section]{subfiles}

\begin{document}

\olfileid{inc}{req}{bet}
\olsection{The Beta Function Lemma}


In order to show that $C$ can carry out primitive recursion, we need
to develop functions that handle sequences. (If we had exponentiation
as well, our task would be easier.) When we had primitive recursion,
we could define things like the ``$n$th prime,'' and pick a fairly
straightforward coding. But here we do not have primitive recursion,
so we need to be more clever.

\begin{lem}
\ollabel{beta:lemma}
There is a function $\beta(d,i)$ in $C$ such that for every sequence
$a_0$, \dots,~$a_n$ there is a number $d$, such that for every $i$ less
than or equal to $n$, $\beta(d,i) = a_i$.
\end{lem}

Think of $d$ as coding the sequence $\langle a_0,\dots,a_n \rangle$, and
$\beta(d,i)$ returning the $i$th element. The lemma is fairly minimal;
it doesn't say we can concatenate sequences or append elements with
functions in $C$, or even that we can \emph{compute} $d$ from
$a_0,\dots,a_n$ using functions in $C$. All it says is that there is
a ``decoding'' function such that every sequence is ``coded.''

The use of the notation $\beta$ is G\"odel's. To repeat, the hard part
of proving the lemma is defining a suitable $\beta$ using the
seemingly restricted resources in the definition of $C$. There are
various ways to prove this lemma, but one of the cleanest is still
G\"odel's original method, which used a number-theoretic fact called
the Chinese Remainder theorem.

\begin{defn}
  Two natural numbers $a$ and $b$ are \emph{relatively prime} if their
  greatest common divisor is 1; in other words, they have no other
  divisors in common.
\end{defn}

\begin{defn}
$a \equiv b \mod c$ means $c \mid (a-b)$, i.e.\ $a$ and $b$ have the
same remainder when divided by $c$.
\end{defn}

Here is the \emph{Chinese Remainder theorem}:
\begin{thm}
Suppose $x_0,\dots,x_n$ are (pairwise) relatively prime. Let
$y_0$, \dots, $y_n$ be any numbers. Then there is a number $z$ such that
\begin{eqnarray*}
z & \equiv & y_0 \mod x_0 \\
z & \equiv & y_1 \mod x_1 \\
& \vdots & \\
z & \equiv & y_n \mod x_n.
\end{eqnarray*}
\end{thm}

Here is how we will use the Chinese Remainder theorem: if
$x_0,\dots,x_n$ are bigger than $y_0,\dots,y_n$ respectively, then
we can take $z$ to code the sequence $\langle y_0,\dots,y_n\rangle$. To recover
$y_i$, we need only divide $z$ by $x_i$ and take the remainder. To use
this coding, we will need to find suitable values for $x_0,\dots,x_n$.

A couple of observations will help us in this regard. Given
$y_0$, \dots,~$y_n$, let
\[
j = \max(n,y_0,\dots,y_n)+1,
\]
and let
\begin{eqnarray*}
x_0 & = & 1 + j\fac \\
x_1 & = & 1 + 2\cdot j\fac \\
x_2 & = & 1 + 3\cdot j\fac \\
& \vdots & \\
x_n & = & 1 + (n+1) \cdot j\fac
\end{eqnarray*}
Then two things are true:
\begin{enumerate}
\item $x_0$, \dots,~$x_n$ are relatively prime.
\item For each $i$, $y_i < x_i$.
\end{enumerate}
To see that clause 1 is true, note that if $p$ is a prime number and
$p \mid x_i$ and $p \mid x_k$, then $p \mid 1 + (i+1) j\fac$ and $p \mid 1 + (k+1)
j\fac$. But then $p$ divides their difference,
\[
(1 + (i+1)j\fac) - (1+ (k+1)j\fac) = (i-k) j\fac.
\]
Since $p$ divides $1 + (1+1)j\fac$, it can't divide $j\fac$ as well
(otherwise, the first division would leave a remainder of 1). So $p$
divides $i-k$. But $\left| i-k \right|$ is at most $n$, and we have
chosen $j > n$, so this implies that $p \mid j\fac$, again a
contradiction. So there is no prime number dividing both $x_i$ and
$x_k$. Clause 2 is easy: we have $y_i < j < j\fac < x_i$.

Now let us prove the $\beta$ function lemma. Remember that $C$ is the
smallest set containing $0$, successor, plus, times, $\Char{=}$,
projections, and closed under composition and $\mu$ applied to regular
functions. As usual, say a relation is in $C$ if its characteristic
function is. As before we can show that the relations in $C$ are
closed under boolean combinations and bounded quantification; for
example:
\begin{enumerate}
\item $\fn{not}(x) \defis \Char{=}(x,0)$
\item $\mu{x \leq z} \; R(x,y) \defis \mu x \; (R(x,y) \lor x = z)$
\item $\lexists[x \leq z][R(x,y)] \defiff R(\mu {x \leq z} \; R(x,y), y)$
\end{enumerate}
We can then show that all of the following are in $C$:
\begin{enumerate}
\item The pairing function, $J(x,y) = \frac{1}{2}[(x+y)(x+y+1)] + x$
% maybe explain more what is going on here, a bit confusing.
\item Projections
\[
K(z) = \mu {x \leq q} \; (\lexists[y \leq z] \, [z = J(x,y)])
\]
and
\[
L(z) = \mu {y \leq q} \; (\lexists[x \leq z]\, [z = J(x,y)]).
\]
\item $x < y$
\item $x \mid y$
% \item $x \tsub y$
% \item $\fn{Prime}(x)$
% \item Assuming $p$ is prime, the relation ``$x$ is a power of $p$'':
% \[
% \lforall[y \leq x][(y \mid x \lif y = 1 \lor y = x)].
% \]
\item The function $\fn{rem}(x,y)$ which returns the remainder when
  $y$ is divided by $x$
\end{enumerate}
Now define
\[
\beta^*(d_0,d_1,i) = \fn{rem}(1+(i+1) d_1,d_0)
\]
and
\[
\beta(d,i) = \beta^*(K(d),L(d),i).
\]
This is the function we need. Given $a_0,\dots,a_n$, as above, let
\[
j = \max(n,a_0,\dots,a_n)+1,
\]
and let $d_1 = j\fac$. By the observations above, we know that $1+d_1,
1+2 d_1, \dots, 1+(n+1) d_1$ are relatively prime and all are bigger
than $a_0,\dots,a_n$. By the Chinese Remainder theorem there is a
value $d_0$ such that for each $i$,
\[
d_0 \equiv a_i \mod (1+(i+1)d_1)
\]
and so (because $d_1$ is greater than $a_i$),
\[
a_i = \fn{rem}(1+(i+1)d_1,d_0).
\]
Let $d = J(d_0,d_1)$. Then for each $i$ from $0$ to $n$, we have
\begin{eqnarray*}
\beta(d,i) & = & \beta^*(d_0,d_1,i) \\
& = & \fn{rem}(1+(i+1) d_1,d_0) \\
& = & a_i
\end{eqnarray*}
which is what we need. This completes the proof of the
$\beta$-function lemma.

\end{document}
