% Part: incompleteness
% Chapter: representability-in-q
% Section: undecidability

\documentclass[../../include/open-logic-section]{subfiles}

\begin{document}

\olfileid{inc}{req}{und}
\olsection{Undecidability}

We call a theory $\Th{T}$ \emph{undecidable} if there is no
computational procedure which, after finitely many steps and
unfailingly, provides a correct answer to the question ``does $\Th{T}$
prove~$!A$'' for any sentence~$!A$ in the language of~$\Th{T}$.  So
$\Th{Q}$ would be decidable iff there were a computational procedure
which decides, given a sentence~$A$ in the language of arithmetic,
whether $\Th{Q} \Proves !A$ or not.  We can make this more precise by
asking: Is the relation~$\fn{Prov}_{\Th{Q}}(x)$, which holds of~$x$
iff $x$ is the G\"odel number of a sentence provable in~$\Th{Q}$,
recursive?  The answer is: no. 

\begin{thm}
$\Th{Q}$ is undecidable, i.e., the relation 
\[
\fn{Prov}_{\Th{Q}}(x) \defiff \fn{Sent}(x) \land
\lexists[y][\fn{Pr}_{\Th{Q}}(x, y)]
\]
is not recursive.
\end{thm}

\begin{proof}
Suppose it were.  Then we could solve the halting problem as follows:
Given $e$ and $n$, we know that $\cfind{e}(n) \defined$ iff there is
an~$s$ such that $T(e, x, s)$, where $T$ is Kleene's predicate from
\olref[cmp][rec][nft]{thm:kleene-nf}. Since $\Char{T}$ is primitive
recursive it is representable in~$\Th{Q}$ by a formula
$!B_T(e,x,s,y)$, that is, $\Th{Q} \Proves !B_T(\num{e}, \num{x},
\num{s}, \num{1})$ iff $T(e, x, s)$.  If $\Th{Q} \Proves !B_T(\num{e},
\num{x}, \num{s}, \num{1})$ then also $\Th{Q} \Proves \Th{Q} \Proves
\lexists[y][!B_T(\num{e}, \num{x}, y, \num{1})]$. If no such $s$
exists, then $\Th{Q} \Proves \lnot !B_T(\num{e}, \num{x}, \num{s},
\num{1})$ for every~$s$. Since $\Th{Q}$ is $\omega$-consistent,
$\Th{Q} \Proves/ \lexists[y][!B_T(\num{e}, \num{x}, y, \num{1})]$.  In
other words, $\Th{Q} \Proves \lexists[y][!B_T(\num{e}, \num{x}, y,
  \num{1})]$ iff there is an $s$ such that $T(e, n, s)$, i.e., iff
$\cfind{e}(n) \defined$.  From $e$ and~$n$ we can compute
$\Gn{\lexists[y][!B_T(\num{e}, \num{n}, y, \num{1})]}$, let $g(e, n)$
  be the primitive recursive function which does that. So
\[
h(e, n) = 
\begin{cases}
1 & \text{if $\fn{Pr}_{\Th{Q}}(g(e, n))$}\\
0 & \text{otherwise}.
\end{cases}
\]
This would show that $h$ is recursive if $\fn{Pr}_{\Th{Q}}$ is. But $h$ is not
recursive, by \olref[cmp][rec][hlt]{thm:halting-problem}, so $\fn{Pr}_{\Th{Q}}$
cannot be either.
\end{proof}

\begin{cor}
First-order logic is undecidable.
\end{cor}

\begin{proof}
If first-order logic were decidable, provability in~$\Th{Q}$ would be
as well, since $\Th{Q} \Proves !A$ iff $\Proves !O \lif !A$, where
$!O$ is the conjunction of the axioms of~$\Th{Q}$.
\end{proof}

\end{document}
