% Part: propositional-logic
% Chapter: propositional-logic 
% Section: completeness

\documentclass[../../../include/open-logic-section]{subfiles}

\begin{document}

\olfileid{pl}{prp}{com}

\olsection{Completeness of Propositional Logic}

\begin{defn}
\ollabel{def:MaxCon}
A set $\Gamma$ of !!{formula}s is \emph{maximally consistent} if it is
consistent and if $\Delta$ is a consistent set such that $\Gamma
\subseteq \Delta$ then $\Gamma = \Delta$.
\end{defn}

\begin{lem}[Truth Lemma]
  \ollabel{lem:truth}
Let $\Gamma$ be maximally consistent; then:
\begin{enumerate}
\item \ollabel{lem:truth-1} $!A \in \Gamma$ if and only if $\lnot !A
  \notin \Gamma$;
\item \ollabel{lem:truth-2} $\Gamma \Proves !A$ if and only if $!A \in \Gamma$; 
\item \ollabel{lem:truth-3} $!A \lif !B \in \Gamma$ if and only if
  either $!A \notin \Gamma$ or $!B \in \Gamma$.
\end{enumerate}
\end{lem}

\begin{proof}
\begin{enumerate}
\item Suppose $!A \in \Gamma$. If also $\lnot !A \in \Gamma$, then
  $\Gamma$ is inconsistent; and if neither $!A$ nor $\lnot!A$ is in
  $\Gamma$ then by \olref[axd]{prop:phi} one of $\Gamma\cup\{!A\}$ and
  $\Gamma \cup \{\lnot !A\}$ is consistent, which means that $\Gamma$
  is not maximal. The other direction is proved the same way.
\item If $!A \in \Gamma$ then of course $\Gamma \Proves !A$.  On the
  other hand, suppose $\Gamma \Proves !A$. If $!A \notin \Gamma$ then,
  by \olref{lem:truth-1}, $\lnot !A \in \Gamma$.  So $\Gamma \Proves
  !A$ and $\Gamma \Proves \lnot !A$, a contradiction.
\item Exercise.
\end{enumerate}
\end{proof}

The left-to-right direction of \olref{lem:truth}\olref{lem:truth-2} shows that
$\Gamma$ is \emph{deductively closed}, i.e., if $\Gamma \Proves !A$ then $!A
\in \Gamma$ for all~$!A$.

\begin{prob}
  Complete the proof of \olref{lem:truth}.
\end{prob}

\begin{thm}[Completeness] 
\ollabel{thm:completeness} 
If $\Gamma$ is consistent then $\Gamma$ is satisfiable.
\end{thm}

\begin{proof}
Let $!A_0$, $!A_1$,~\dots be an exhaustive listing of all the
!!{formula}s of the language. We recursively define an increasing
sequence of sets of !!{formula}s $\Gamma_0$, $\Gamma_1$,~\dots, by
putting:
\begin{align*}
\Gamma_0 & = \Gamma\\
\Gamma_{n+1} &  =
\begin{cases}
  \Gamma_n \cup \{!A_n\} & \text{if $\Gamma_n \cup \{!A_n\}$ is
    consistent;}\\
  \Gamma_n \cup \{\lnot !A_n\} & \text{otherwise}.
\end{cases}
\end{align*}
Then define:
\[
\Gamma^* = \bigcup_{0\le n}\Gamma_n.
\]
The proof now proceeds by establishing, in turn, the following
facts:
\begin{enumerate}
\item For each $n$, the set $\Gamma_n$ is consistent (by induction on
  $n$, using \olref[axd]{prop:phi});
\item $\Gamma^*$ is consistent;
\item $\Gamma^*$ is maximal.
\end{enumerate}
Then define a valuation $v$ by putting $v(p_i) = \True$ if and
only if $p_i \in \Gamma^*$. By induction on $!A$ it is then shown that
membership in $\Gamma^*$ coincides with truth according to $v$ also
for more complex !!{sentence}s: $\overline{v}(!A) = \True$ if and
only if $!A \in \Gamma^*$. In particular, $v$ satisfies $\Gamma^*$, and
since $\Gamma \subseteq \Gamma^*$, also $\Gamma$ is satisfiable, as
desired.
\end{proof}

\begin{cor}
If $\Gamma \Entails !A$ then $\Gamma \Proves !A$.
\end{cor}

\begin{proof}
If $\Gamma\Proves/ !A$ then $\Gamma \cup \{\lnot !A\}$ is consistent,
by \olref[axd]{prop:prov-incons}. So by the completeness theorem, $\Gamma \cup
\{\lnot !A\}$ is satisfiable, and $\Gamma \Entails/ !A$.
\end{proof}

\begin{prop}[Compactness Theorem]
\ollabel{prop:compactness}
$\Gamma$ is satisfiable if and only if every \emph{finite} subset
$\Gamma_0$ of $\Gamma$ is satisfiable.
\end{prop}

\begin{proof}
$\Gamma$ is unsatisfiable if and only if it is inconsistent, if and
  only if some finite subset $\Gamma_0$ of $\Gamma$ is inconsistent,
  if and only if some finite subset $\Gamma_0$ of $\Gamma$ is
  unsatisfiable.
\end{proof}

\begin{cor}
$\Gamma \Entails !A$ if and only if for some finite subset
$\Gamma_0$ of $\Gamma$, $\Gamma_0 \Entails !A$.
\end{cor}

\end{document}
