% Part: propositional-logic
% Chapter: syntax-and-semantics
% Section: semantic-notions

\documentclass[../../../include/open-logic-section]{subfiles}

\begin{document}

\olfileid{pl}{syn}{sem}

\olsection{Semantic Notions}

We define the following semantic notions:

\begin{defn} 
\begin{enumerate}
\item !!^a{formula} $!A$ is \emph{satisfiable} if for
  some~$\pAssign{v}$, $\pSat{v}{!A}$; it is
  \emph{unsatisfiable} if for no $\pAssign{v}$, $\pSat{v}{!A}$;
\item !!^a{formula} $!A$ is a \emph{tautology} if $\pSat{v}{!A}$ for
  all !!{valuation}s~$v$;
\item !!^a{formula} $!A$ is \emph{contingent} if it is satisfiable but
  not a tautology;
\item If $\Gamma$ is a set of !!{formula}s, $\Gamma \Entails !A$ (``$\Gamma$
  entails $!A$'') if and only if $\pSat{v}{!A}$ for every
  !!{valuation}~$\pAssign{v}$ for which $\pSat{v}{\Gamma}$.
\item If $\Gamma$ is a set of !!{formula}s, $\Gamma$ is
  \emph{satisfiable} if there is !!a{valuation}~$\pAssign{v}$ for which
  $\pSat{v}{\Gamma}$, and $\Gamma$ is
  \emph{unsatisfiable} otherwise.
\end{enumerate} 
\end{defn}

\begin{prop}
\ollabel{prop:semanticalfacts} 
\begin{enumerate} 
\item $!A$ is a tautology if and only if
  $\emptyset \Entails !A$; 
\item If $\Gamma \Entails !A$ and $\Gamma \Entails !A \lif !B$ then
  $\Gamma \Entails !B$;
\item If $\Gamma$ is satisfiable then every finite subset of $\Gamma$
  is also satisfiable; 
\item \ollabel{def:Monotony} Monotony: if $\Gamma \subseteq \Delta$
  and $\Gamma \Entails !A$ then also $\Delta \Entails !A$;
\item \ollabel{def:Cut} Transitivity: if $\Gamma \Entails !A$ and
  $\Delta \cup \{ !A\} \Entails !B$ then $\Gamma \cup \Delta \Entails
  !B$;
\end{enumerate}
\end{prop}

\begin{proof}
Exercise.
\end{proof}

\begin{prob}
Prove \olref[pl][syn][sem]{prop:semanticalfacts}
\end{prob}

\begin{prop}\ollabel{prop:entails-unsat}
  $\Gamma \Entails !A$ if and only if $\Gamma \cup \{\lnot !A\}$
  is unsatisfiable; 
\end{prop}

\begin{proof}
Exercise.
\end{proof}

\begin{prob}
Prove \olref[pl][syn][sem]{prop:entails-unsat}
\end{prob}

\begin{thm}[Semantic Deduction Theorem]
  \ollabel{thm:sem-deduction} $\Gamma \Entails !A \lif !B$ if and only
  if $\Gamma \cup \{!A\} \Entails !B$.
\end{thm}

\begin{proof}
Exercise.
\end{proof}

\begin{prob}
Prove \olref[pl][syn][sem]{thm:sem-deduction}
\end{prob}

We write $!A \Entails !B$ for $\Gamma \Entails !B$ when $\Gamma = \{!A\}$ is a singleton and say that two !!{formula}s are semantically equivalent, $!A \cong !B$, if $!A \Entails !B$ and $!B \Entails !A$.

By looking at the following equivalences, known as the De Morgan Laws, we can see that the connectives $\land$ and $\lor$ seem to behave in a similar way. 
  \begin{align*}
    ({!A} \land {!B}) &\cong \lnot(\lnot{!A} \lor \lnot{!B})\\
    ({!A} \lor {!B}) &\cong \lnot(\lnot{!A} \land \lnot{!B})
  \end{align*}
We will now make this tight symmetry between conjunction and disjuntion precise in the following result.

\begin{defn}
Define, by recursion on formulas, the mapping that takes a !!{formula} with only occurrences of $\lnot$, $\land$, and $\lor$ to another such !!{formula} such that:
\begin{itemize}
  \item $!A^d \ident !A$ when $!A$ is atomic,
  \item $(\lnot !A )^d \ident \lnot !A^d$,
  \item $(!A \land !B)^d \ident !A^d \lor !B^d$,
  \item $(!A \lor !B)^d \ident !A^d \land !B^d$.
\end{itemize}
\end{defn}

\begin{prop}\ollabel{thm:dual}
$!A \cong !B$ iff $!A^d \cong !B^d$ whenever the dual mapping is defined.
\end{prop}
\begin{proof}
Exercise.
\end{proof}

\begin{prob}
Prove \olref[pl][syn][sem]{thm:dual} by introducting an axuilliary mapping $!A^n$ just as $!A^d$ except for atomic !!{formula}s where $!A^n$ is defined to be $\lnot !A$ and proving that $!A^n \cong \lnot !A$.
\end{prob}

\end{document}
