\documentclass[../../../include/open-logic-section]{subfiles}

\begin{document}

\olfileid{his}{set}{pathology}
\olsection{Pathologies}

However, the definition of a \emph{limit} turned out to allow for some rather ``pathological'' constructions. 

Around the 1830s, Bolzano discovered a function which was
\emph{continuous everywhere}, but \emph{differentiable nowhere}.
(Unfortunately, Bolzano never published this; the idea was first
encountered by mathematicians in 1872, thanks to Weierstrass's
independent discovery of the same idea.)\footnote{The history is
documented in extremely thorough footnotes to the Wikipedia article on
\href{http://en.wikipedia.org/wiki/Weierstrass_function}{the
Weierstrass function}.} This was, to say the least, rather surprising.
It is easy to find functions, such as $|x|$, which are continuous
everywhere but not differentiable at a particular point. But a
function which is continuous everywhere but differentiable
\emph{nowhere} is a very different beast. Consider, for a moment, how
you might try to draw such a function. To ensure it is continuous, you
must be able to draw it without ever removing your pen from the page;
but to ensure it is differentiable nowhere, you would have to abruptly
change the direction of your pen, constantly.

Further ``pathologies'' followed. In January 5 1874, Cantor wrote a
letter to Dedekind, posing the problem:
\begin{quote}
Can a surface (say a square including its boundary) be one-to-one
correlated to a line (say a straight line including its endpoints) so
that to every point of the surface there corresponds a point of the
line, and conversely to every point of the line there corresponds a
point of the surface?

It still seems to me at the moment that the answer to this question is
very difficult---although here too one is so impelled to say \emph{no}
that one would like to hold the proof to be almost superfluous.
[Quoted in \citealt{Gouvea2011}]
\end{quote}
But, in 1877, Cantor proved that he had been wrong. In fact, a line
and a square have exactly the same number of points. He wrote on 29
June 1877 to Dedekind ``\emph{je le vois, mais je ne le crois pas}'';
that is, ``I see it, but I don't believe it''. In the ``received
history'' of mathematics, this is often taken to indicate just how
\emph{literally incredible} these new results were to the
mathematicians of the time. (The correspondence is presented in
\citet{Gouvea2011}, and we return to it in
\olref[his][set][mythology]{sec}. Cantor's proof is outlined in
\olref[his][set][cantorplane]{sec}.) 

Inspired by Cantor's result, Peano started to consider whether it
might be possible to map a line \emph{smoothly} onto a plane. This
would be a \emph{curve which fills space}. In \citeyear{Peano1890},
Peano constructed just such a curve. This is truly counter-intuitive:
Euclid had defined a line as ``breadthless length'' (Book I,
Definition 2), but Peano had shown that, by curling up a line
appropriately, its length can be turned into breadth. In
\citeyear{Hilbert1891}, Hilbert described a slightly more intuitive
space-filling curve, together with some pictures illustrating it. The
curve is constructed in sequence, and here are the first six stages of
the construction:
\begin{center}
\begin{tikzpicture}
\begin{scope}[xshift=20pt, yshift=20pt]
	\draw [oldiagcolorC, l-system={Hilbert curve, step=40pt, angle=90, axiom=L, order=1}] lindenmayer system;
\end{scope}
\begin{scope}[xshift=110pt, yshift=10pt]
	\draw [oldiagcolorC, l-system={Hilbert curve, step=20pt, angle=90, axiom=L, order=2}] lindenmayer system;
\end{scope}
\begin{scope}[xshift=205pt, yshift=5pt]
	\draw [oldiagcolorC, l-system={Hilbert curve, step=10pt, angle=90, axiom=L, order=3}] lindenmayer system;
\end{scope}
\begin{scope}[xshift=2.5pt, yshift=-97.5pt]
	\draw [oldiagcolorC, l-system={Hilbert curve, step=5pt, angle=90, axiom=L, order=4}] lindenmayer system;
\end{scope}
\begin{scope}[xshift=101.25pt, yshift=-98.75pt]
	\draw [oldiagcolorC, l-system={Hilbert curve, step=2.5pt, angle=90, axiom=L, order=5}] lindenmayer system;
\end{scope}
\begin{scope}[xshift=200.625pt, yshift=-99.375pt]
	\draw [oldiagcolorC, l-system={Hilbert curve, step=1.25pt, angle=90, axiom=L, order=6}] lindenmayer system;
\end{scope}
\draw[gray] (0pt,0pt) rectangle (80pt, 80pt);
\draw[gray] (100pt,0pt) rectangle (180pt, 80pt);
\draw[gray] (200pt,0pt) rectangle (280pt, 80pt);
\draw[gray] (0pt,-100pt) rectangle (80pt, -20pt);
\draw[gray] (100pt,-100pt) rectangle (180pt, -20pt);
\draw[gray] (200pt,-100pt) rectangle (280pt, -20pt);
\end{tikzpicture}  
\end{center}
In the limit---a notion which had, by now, received rigorous
definition---the entire square is filled in solid !!{colorC}. And, in
passing, Hilbert's curve is continuous everywhere but differentiable
nowhere; intuitively because, in the infinite limit, the function
abruptly changes direction at every moment. (We will outline Hilbert's
construction in more detail in \olref[his][set][hilbertcurve]{sec}.)

For better or worse, these ``pathological'' geometric constructions
were treated as a reason to doubt appeals to geometric intuition. They
became something approaching \emph{propaganda} for a new way of doing
mathematics, which would culminate in set theory. In the later
myth-building of the subject, it was repeated, often, that these
results were both perfectly rigorous and perfectly shocking. They
therefore served a dual purpose: as a warning against relying upon
geometric intuition, and as a demonstration of the fertility of new
ways of thinking. 

\end{document}
