% Part: first-order-logic
% Chapter: semantics
% Section: intro-semantics

\documentclass[../../../include/open-logic-section]{subfiles}

\begin{document}

\olfileid{fol}{syn}{its}

\olsection{Introduction}

Giving the meaning of expressions is the domain of semantics.  The
central concept in semantics is that of satisfaction in
!!a{structure}. !!^a{structure} gives meaning to the building blocks
of the language: !!a{domain} is a non-empty set of objects. The
quantifiers are interpreted as ranging over this domain, !!{constant}s
are assigned elements in the domain, !!{function}s are assigned
functions from the !!{domain} to itself, and !!{predicate}s are
assigned relations on the !!{domain}.  The !!{domain} together with
assignments to the basic vocabulary constitutes !!a{structure}.
!!^{variable}s may appear in !!{formula}s, and in order to give a
semantics, we also have to assign !!{element}s of the !!{domain} to
them---this is a variable assignment. The satisfaction relation,
finally, brings these together. !!^a{formula} may be satisfied in
!!a{structure}~$\Struct{M}$ relative to !!a{variable} assignment~$s$,
written as $\Sat{M}{!A}[s]$. This relation is also defined by
induction on the structure of~$!A$, using the truth tables for the
logical connectives to define, say, satisfaction of $(!A \land !B)$ in
terms of satisfaction (or not) of $!A$ and ~$!B$.  It then turns out
that the !!{variable} assignment is irrelevant if the !!{formula}~$!A$
is !!a{sentence}, i.e., has no free variables, and so we can talk of
!!{sentence}s being simply satisfied (or not) in !!{structure}s.

On the basis of the satisfaction relation $\Sat{M}{!A}$ for !!{sentence}s
we can then define the basic semantic notions of validity, entailment,
and satisfiability. !!^a{sentence} is valid, $\Entails !A$, if every
!!{structure} satisfies it. It is entailed by a set of !!{sentence}s,
$\Gamma \Entails !A$, if every !!{structure} that satisfies all the
!!{sentence}s in~$\Gamma$ also satisfies~$!A$. And a set of !!{sentence}s
is satisfiable if some !!{structure} satisfies all !!{sentence}s in it
at the same time.  Because !!{formula}s are inductively defined, and
satisfaction is in turn defined by induction on the structure of
!!{formula}s, we can use induction to prove properties of our
semantics and to relate the semantic notions defined.

\end{document}
