% Part: normal-modal-logic
% Chapter: frame-correspondence
% Section: properties-accessibility

\documentclass[../../../include/open-logic-section]{subfiles}

\begin{document}

\olfileid{nml}{frd}{acc}

\olsection{Properties of Accessibility Relations}

Many modal !!{formula}s turn out to be characteristic of simple, and
even familiar, properties of the accessibility relation. In one
direction, that means that any model that has a given property makes a
corresponding !!{formula} (and all its substitution instances)
true. We begin with five classical examples of kinds of accessibility
relations and the formulas the truth of which they guarantee.

\begin{thm}\ollabel{thm:soundschemas}
  Let $\mModel{M} = \tuple{W, R, V}$ be a model.  If $R$ has the
  property on the left side of \olref{tab:five}, every instance of the
  !!{formula} on the right side is true in~$\mModel{M}$.
\end{thm}

\begin{table}[t]
    \begin{tabular}{| l || l |}
      \hline
      {\emph{If $R$ is \dots}} & {\emph{then \dots is true in~$\mModel{M}$:}} \\
      \hline \hline
      \emph{serial}: $\forall u \exists v Ruv$ & \hbox to.43\textwidth
           {$\Box p \lif \Diamond p$ \hfill (\Ax{D})} \\
      \hline
      \emph{reflexive}: $\forall w Rww$  
      & \hbox to.43\textwidth
          {$\Box p \lif p$ \hfill (\Ax{T})} \\
      \hline
      \emph{symmetric}: &  \hbox to .43\textwidth
           {$p \lif \Box\Diamond p$ \hfill (\Ax{B})} \\
           $\forall u\forall v(Ruv \lif Rvu)$ & \\
      \hline
      \emph{transitive}: & \hbox to.43\textwidth
           {$\Box p \lif \Box \Box p$ \hfill (\Ax{4})} \\
      $\forall u \forall v \forall w ((Ruv \land Rvw) \lif Ruw)$ & \\
      \hline 
      \emph{euclidean}: & \hbox to .43\textwidth
        {$\Diamond p \lif \Box\Diamond p$ \hfill (\Ax{5})} \\
      $\forall w \forall u \forall v ((Rwu \land Rwv) \lif Ruv)$ &\\
      \hline
    \end{tabular}
    \caption{Five correspondence facts.}
    \ollabel{tab:five}
  \end{table} 


\begin{proof}
  Here is the case for \Ax{B}: to show that the schema is true in a
  model we need to show that all of its instances are true at all worlds
  in the model. So let $!A \lif \Box\Diamond !A$ be a given instance
  of \Ax{B}, and let $w \in W$ be an arbitrary world. Suppose the
  antecedent $!A$ is true at $w$, in order to show that $\Box \Diamond
  !A$ is true at $w$. So we need to show that $\Diamond !A$ is true at
  all $w'$ accessible from $w$. Now, for any $w'$ such that $Rww'$ we
  have, using the hypothesis of symmetry, that also $Rw'w$ (see
  \olref{fig:Bsymm}). Since $\mSat{M}{!A}[w]$, we have
  $\mSat{M}{\Diamond !A}[w']$. Since $w'$ was an arbitrary world such
  that $Rww'$, we have $\mSat{M}{\Box\Diamond!A}[w]$.

  We leave the other cases as exercises.
\end{proof}

\begin{prob}
  Complete the proof of \olref[nml][frd][acc]{thm:soundschemas}.
\end{prob}

\begin{figure}
  \begin{center}
    \begin{tikzpicture}[modal]
      \node[world] (w1) [label={below:$\mSat{{}}{\formula{A}}$\\
          $\mSat{{}}{\Box\Diamond \formula{A}}$}] {$w$}; 
      \node[world] (w2) [label=below:{$\mSat{{}}{\Diamond \formula{A}}$},
        right=of w1] {$w'$}; 
      \draw[->, bend left] (w1) to (w2); 
      \draw[->, bend left] (w2) to (w1); 
    \end{tikzpicture}
  \end{center}
\caption{The argument from symmetry.}
\ollabel{fig:Bsymm}
\end{figure}

Notice that the converse implications of \olref{thm:soundschemas} do
not hold: it's not true that if a model verifies a schema, then the
accessibility relation of that model has the corresponding property.
In the case of \Ax{T} and reflexive models, it is easy to give an
example of a model in which \Ax{T} itself fails: let $W = \{w\}$ and
$V(p) = \emptyset$. Then $R$ is not reflexive, but $\mSat{M}{\Box
  p}[w]$ and $\mSat/{M}{p}[w]$. But here we have just a single
instance of \Ax{T} that fails in~$\mModel{M}$; other instances, e.g.,
$\Box \lnot p \lif \lnot p$ are true. It is harder to give examples
where \emph{every substitution instance} of~\Ax{T} is true
in~$\mModel{M}$ and $\mModel{M}$ is not reflexive. But there are such
models, too:

\begin{prop}\ollabel{prop:reflexive}
  Let $\mModel{M} = \tuple{W, R, V}$ be a model such that $W = \{u, v
  \}$, where worlds $u$ and $v$ are related by $R$: i.e., both $Ruv$
  and $Rvu$.  Suppose that for all $p$: $u \in V(p) \Leftrightarrow v
  \in V(p)$. Then:
  \begin{enumerate}
  \item For all $!A$: $\mSat{M}{!A}[u]$ if and only if
    $\mSat{M}{!A}[v]$ (use induction on $!A$).
  \item Every instance of~\Ax{T} is true in $\mModel{M}$.
  \end{enumerate}
  Since $\mModel{M}$ is not reflexive (it is, in fact,
  \emph{irreflexive}), the converse of \olref{thm:soundschemas}
  fails in the case of \Ax{T} (similar arguments can be given for
  some---though not all---the other schemas mentioned in
  \olref{thm:soundschemas}).
\end{prop}

\begin{prob}
  Prove the claims in \olref[nml][frd][acc]{prop:reflexive}.
\end{prob}

Although we will focus on the five classical !!{formula}s \Ax{D},
\Ax{T}, \Ax{B}, \Ax{4}, and \Ax{5}, we record in
\olref{tab:anotherfive} a few more properties of accessibility
relations. The accessibility relation~$R$ is partially functional, if
from every world at most one world is accessible. If it is the case
that from every world exactly one world is accessible, we call it
functional. (Thus the functional relations are precisely those that
are both serial and partially functional). They are called
``functional'' because the accessibility relation operates like a
(partial) function. A relation is weakly dense if whenever $Ruv$,
there is a~$w$ ``between'' $u$ and~$v$. So weakly dense relations are
in a sense the opposite of transitive relations: in a transitive
relation, whenever you can reach $v$ from~$u$ by a detour via~$w$, you
can reach $v$ from $u$ directly; in a weakly dense relation, whenever
you can reach $v$ from~$u$ directly, you can also reach it by a detour
via some~$w$. A relation is weakly directed if whenever you can reach
worlds $u$ and~$v$ from some world~$w$, you can reach a single
world~$t$ from both $u$ and~$v$---this is sometimes called the
``diamond property'' or ``confluence.''

\begin{table}[t]
    \begin{tabular}{| p{.50\textwidth} || p{.43\textwidth} |}
      \hline
      {\emph{If $R$ is \dots}} & {\emph{then \dots is true in~$\mModel{M}$:}} \\
      \hline \hline
      \emph{partially functional}: & 
      \multirow{2}{*}{$\Diamond p \lif \Box p$} \\
      $ \forall w \forall u \forall v ((Rwu \land Rwv) \lif u =v )$ & \\
      \hline
      \multirow{2}{*}{\emph{functional}: $\forall w \exists v \forall u(Rwu \liff \eq[u][v]) $} 
      & \multirow{2}{*}{$\Diamond p \liff \Box p$} \\
      & \\
      \hline
      \emph{weakly dense}: &  \multirow{2}{*}{$\Box \Box p \lif
        \Box p$} \\
      $\forall u\forall v(Ruv \lif \exists w(Ruw \land Rwv))$ & \\
      \hline
      \emph{weakly connected}: &
      \multirow{3}{*}{\hbox to.43\textwidth{$
        \begin{array}{@{}l@{}}
          \Box ((p \land \Box p) \lif q) \lor {} \\
          \qquad \Box ((q \land \Box q) \lif p)
        \end{array}$ \hfill (\Ax{L})}
      } \\
      $\forall w \forall u \forall v ((Rwu \land Rwv) \lif {}$ &\\
      \qquad $(Ruv \lor u=v \lor Rvu))$ & \\
      \hline 
      \emph{weakly directed}: & \multirow{3}{*}{\hbox to .43\textwidth
        {$\Diamond\Box p \lif \Box\Diamond p$\hfill (\Ax{G})}} \\
      $\forall w \forall u \forall v ((Rwu \land Rwv) \lif {}$ & \\
      \qquad $\exists t (Rut \land Rvt))$ & \\
      \hline
    \end{tabular}
    \caption{Five more correspondence facts.}
    \ollabel{tab:anotherfive}
  \end{table} 

\begin{prob}
  Let $\mModel{M} = \tuple{W, R, V}$ be a model. Show that if $R$
  satisfies the left-hand properties of
  \olref[nml][frd][acc]{tab:anotherfive}, every instance of the
  corresponding right-hand formula is true in~$\mModel{M}$.
\end{prob}



\end{document}
