% Part: first-order-logic
% Chapter: tableaux
% Section: soundness.tex

\documentclass[../../../include/open-logic-section]{subfiles}

\begin{document}

\iftag{FOL}
      {\olfileid{fol}{tab}{sou}}
      {\olfileid{pl}{tab}{sou}}
      
\olsection{Soundness}

\begin{explain}
!!^a{derivation} system, such as tableaux, is \emph{sound}
if it cannot !!{derive} things that do not actually hold.  Soundness is
thus a kind of guaranteed safety property for !!{derivation} systems.
Depending on which proof theoretic property is in question, we would
like to know for instance, that
\begin{enumerate}
\item every !!{derivable}~$!A$ is \iftag{FOL}{valid}{a tautology};
\item if !!a{sentence} is !!{derivable} from some others, it is also a
  consequence of them;
\item if a set of !!{sentence}s is inconsistent, it is unsatisfiable.
\end{enumerate}
These are important properties of !!a{derivation} system.  If any of them do
not hold, the !!{derivation} system is deficient---it would !!{derive} too much.
Consequently, establishing the soundness of !!a{derivation} system is of the
utmost importance.

Because all these proof-theoretic properties are defined via closed
!!{tableau}s of some kind or other, proving (1)--(3) above requires
proving something about the semantic properties of closed
!!{tableau}s.  We will first define what it means for !!a{signed
  formula} to be satisfied in a structure, and then show that if a
!!{tableau} is closed, no structure satisfies all its assumptions.
(1)--(3) then follow as corollaries from this result.
\end{explain}

\begin{defn}
\iftag{FOL}{!!^a{structure}}{!!^a{valuation}}~$\iftag{FOL}{\Struct{M}}{\pAssign{v}}$
\emph{satisfies} !!a{signed formula} $\sFmla{\True}{!A}$ iff
$\iftag{FOL}{\Sat{M}}{\pSat{v}}{!A}$, and it satisfies
$\sFmla{\False}{!A}$ iff
$\iftag{FOL}{\Sat/{M}}{\pSat/{v}}{!A}$. $\iftag{FOL}{\Struct{M}}{\pAssign{v}}$
satisfies a set of !!{signed formula}s~$\Gamma$ iff it satisfies every
$\sFmla{S}{!A} \in \Gamma$. $\Gamma$~is \emph{satisfiable} if there is
\iftag{FOL}{!!a{structure}}{!!a{valuation}} that satisfies it, and
\emph{unsatisfiable} otherwise.
\end{defn}

\begin{thm}[Soundness]
  \ollabel{thm:tableau-soundness}
  If $\Gamma$ has a closed !!{tableau}, $\Gamma$ is unsatisfiable.
\end{thm}

\begin{proof}
Let's call a branch of !!a{tableau} satisfiable iff the set of
!!{signed formula}s on it is satisfiable, and let's call !!a{tableau}
satisfiable if it contains at least one satisfiable branch.

We show the following: Extending a satisfiable !!{tableau} by one of
the rules of inference always results in a satisfiable !!{tableau}.
This will prove the theorem: any closed !!{tableau} results by
applying rules of inference to the !!{tableau} consisting only of
assumptions from~$\Gamma$. So if $\Gamma$ were satisfiable, any
!!{tableau} for it would be satisfiable. A closed !!{tableau},
however, is clearly not satisfiable: every branch contains both
$\sFmla{\True}{!A}$ and $\sFmla{\False}{!A}$, and no structure can
both satisfy and not satisfy~$!A$.

Suppose we have a satisfiable !!{tableau}, i.e., !!a{tableau} with at
least one satisfiable branch. Applying a rule of inference either adds
!!{signed formula}s to a branch, or splits a branch in two. If the
!!{tableau} has a satisfiable branch which is not extended by the rule
application in question, it remains a satisfiable branch in the
extended !!{tableau}, so the extended tableau is satisfiable. So we
only have to consider the case where a rule is applied to a
satisfiable branch.

Let $\Gamma$ be the set of !!{signed formula}s on that branch, and
let $\sFmla{S}{!A} \in \Gamma$ be the !!{signed formula} to which the
rule is applied. If the rule does not result in a split branch, we
have to show that the extended branch, i.e., $\Gamma$ together with
the conclusions of the rule, is still satisfiable. If the rule results
in a split branch, we have to show that at least one of the two
resulting branches is satisfiable.
  
First, we consider the possible inferences that do not result in a split branch.
\begin{enumerate}
\item The branch is expanded by applying $\TRule{\True}{\lnot}$ to
  $\sFmla{\True}{\lnot !B} \in \Gamma$. Then the extended branch
  contains the !!{signed formula}s $\Gamma \cup
  \{\sFmla{\False}{!B}\}$.  Suppose
  $\iftag{FOL}{\Sat{M}}{\pSat{v}}{\Gamma}$. In particular,
  $\iftag{FOL}{\Sat{M}}{\pSat{v}}{\lnot !B}$. Thus,
  $\iftag{FOL}{\Sat/{M}}{\pSat/{v}}{!B}$, i.e.,
  $\iftag{FOL}{\Struct{M}}{\pAssign{v}}$ satisfies
  $\sFmla{\False}{!B}$.
\item The branch is expanded by applying $\TRule{\False}{\lnot}$ to
  $\sFmla{\False}{\lnot !B} \in \Gamma$: Exercise.
\item The branch is expanded by applying $\TRule{\True}{\land}$ to
  $\sFmla{\True}{!B \land !C} \in \Gamma$, which results in two new
  !!{signed formula}s on the branch: $\sFmla{\True}{!B}$ and
  $\sFmla{\True}{!C}$. Suppose
  $\iftag{FOL}{\Sat{M}}{\pSat{v}}{\Gamma}$, in particular
  $\iftag{FOL}{\Sat{M}}{\pSat{v}}{!B \land !C}$. Then
  $\iftag{FOL}{\Sat{M}}{\pSat{v}}{!B}$ and
  $\iftag{FOL}{\Sat{M}}{\pSat{v}}{!C}$. This means that
  $\iftag{FOL}{\Struct{M}}{\pAssign{v}}$ satisfies both
  $\sFmla{\True}{!B}$ and $\sFmla{\True}{!C}$.
\item The branch is expanded by applying $\TRule{\False}{\lor}$ to
  $\sFmla{\False}{!B \lor !C} \in \Gamma$: Exercise.
\item The branch is expanded by applying $\TRule{\False}{\lif}$ to
  $\sFmla{\False}{!B \lif !C} \in \Gamma$: This results in two new
  !!{signed formula}s on the branch: $\sFmla{\True}{!B}$ and
  $\sFmla{\False}{!C}$. Suppose
  $\iftag{FOL}{\Sat{M}}{\pSat{v}}{\Gamma}$, in particular
  $\iftag{FOL}{\Sat/{M}}{\pSat/{v}}{!B \lif !C}$. Then
  $\iftag{FOL}{\Sat{M}}{\pSat{v}}{!B}$ and
  $\iftag{FOL}{\Sat/{M}}{\pSat/{v}}{!C}$. This means that
  $\iftag{FOL}{\Struct{M}}{\pAssign{v}}$ satisfies both
  $\sFmla{\True}{!B}$ and $\sFmla{\False}{!C}$.

\iftag{FOL}{%
\item The branch is expanded by applying $\TRule{\True}{\lforall}$ to
  $\sFmla{\True}{\lforall[x][!B(x)]} \in \Gamma$: This results in a
  new !!{signed formula}~$\sFmla{\True}{!A(t)}$ on the branch.
  Suppose $\Sat{M}{\Gamma}$, in particular,
  $\Sat{M}{\lforall[x][!A(x)]}$.  By
  \olref[syn][sem]{prop:quant-terms}, $\Sat{M}{!A(t)}$.  Consequently,
  $\Struct{M}$ satisfies $\sFmla{\True}{!A(t)}$.

\item The branch is expanded by applying $\TRule{\False}{\lforall}$ to
  $\sFmla{\False}{\lforall[x][!B(x)]} \in \Gamma$: This results in a
  new !!{signed formula}~$\sFmla{\False}{!A(a)}$ where $a$ is
  !!a{constant} not occurring in~$\Gamma$. Since $\Gamma$ is
  satisfiable, there is a $\Struct{M}$ such that $\Sat{M}{\Gamma}$, in
  particular $\Sat/{M}{\lforall[x][!B(x)]}$. We have to show that
  $\Gamma \cup \{\sFmla{\False}{!A(a)}\}$ is satisfiable. To do this,
  we define a suitable~$\Struct{M'}$ as follows.

  By \olref[syn][ass]{prop:sat-quant}, $\Sat/{M}{\lforall[x][!B(x)]}$
  iff for some $s$, $\Sat/{M}{!B(x)}[s]$. Now let $\Struct{M'}$ be
  just like $\Struct{M}$, except $\Assign{a}{M'} = s(x)$.  By
  \olref[syn][ext]{cor:extensionality-sent}, for any
  $\sFmla{\True}{!C} \in \Gamma$, $\Sat{M'}{!C}$, and for any
  $\sFmla{\False}{!C} \in \Gamma$, $\Sat/{M'}{!C}$, since $a$ does not
  occur in~$\Gamma$.

  By \olref[syn][ext]{prop:extensionality}, $\Sat/{M'}{!A(x)}[s]$.  By
  \olref[syn][ext]{prop:ext-formulas}, $\Sat/{M'}{!A(a)}[s]$.  Since
  $!A(a)$ is !!a{sentence}, by
  \olref[syn][ass]{prop:sentence-sat-true}, $\Sat/{M'}{!A(a)}$, i.e.,
  $\Struct{M'}$ satisfies $\sFmla{\False}{!A(a)}$.
\item The branch is expanded by applying $\TRule{\True}{\lexists}$ to
  $\sFmla{\True}{\lexists[x][!B(x)]} \in \Gamma$: Exercise.
\item The branch is expanded by applying $\TRule{\False}{\lexists}$ to
  $\sFmla{\False}{\lexists[x][!B(x)]} \in \Gamma$: Exercise.}{}
\end{enumerate}
Now let's consider the possible inferences that result in a split branch.
\begin{enumerate}
\item The branch is expanded by applying $\TRule{\False}{\land}$ to
  $\sFmla{\False}{!B \land !C} \in \Gamma$, which results in two
  branches, a left one continuing through $\sFmla{\False}{!B}$ and a
  right one through $\sFmla{\False}{!C}$. Suppose
  $\iftag{FOL}{\Sat{M}}{\pSat{v}}{\Gamma}$, in particular
  $\iftag{FOL}{\Sat/{M}}{\pSat/{v}}{!B \land !C}$.  Then
  $\iftag{FOL}{\Sat/{M}}{\pSat/{v}}{!B}$ or
  $\iftag{FOL}{\Sat/{M}}{\pSat/{v}}{!C}$. In the former case,
  $\iftag{FOL}{\Struct{M}}{\pAssign{v}}$ satisfies
  $\sFmla{\False}{!B}$, i.e., $\iftag{FOL}{\Struct{M}}{\pAssign{v}}$
  satisfies the formulas on the left branch. In the latter,
  $\iftag{FOL}{\Struct{M}}{\pAssign{v}}$ satisfies
  $\sFmla{\False}{!C}$, i.e., $\iftag{FOL}{\Struct{M}}{\pAssign{v}}$
  satisfies the formulas on the right branch.
\item The branch is expanded by applying $\TRule{\True}{\lor}$ to
  $\sFmla{\True}{!B \lor !C} \in \Gamma$: Exercise.
\item The branch is expanded by applying $\TRule{\True}{\lif}$ to
  $\sFmla{\True}{!B \lif !C} \in \Gamma$: Exercise.
\item The branch is expanded by \Cut: This results in two branches,
  one containing $\sFmla{\True}{!B}$, the other containing
  $\sFmla{\False}{!B}$. Since $\iftag{FOL}{\Sat{M}}{\pSat{v}}{\Gamma}$
  and either $\iftag{FOL}{\Sat{M}}{\pSat{v}}{!B}$ or
  $\iftag{FOL}{\Sat/{M}}{\pSat/{v}}{!B}$,
  $\iftag{FOL}{\Struct{M}}{\pAssign{v}}$ satisfies either the left or
  the right branch.
\end{enumerate}
\end{proof}

\tagprob{FOL}
\begin{prob}
Complete the proof of \olref[fol][tab][sou]{thm:tableau-soundness}.
\end{prob}
\tagendprob

\tagprob{notFOL}
\begin{prob}
Complete the proof of \olref[pl][tab][sou]{thm:tableau-soundness}.
\end{prob}
\tagendprob

\begin{cor}
\ollabel{cor:weak-soundness}
If $\Proves !A$ then $!A$ is \iftag{FOL}{valid}{a tautology}.
\end{cor}

\begin{cor}
\ollabel{cor:entailment-soundness}
If $\Gamma \Proves !A$ then $\Gamma \Entails !A$.
\end{cor}

\begin{proof}
  If $\Gamma \Proves !A$ then for some $!B_1$, \dots, $!B_n \in
  \Gamma$, $\{\sFmla{\False}{!A}, \sFmla{\True}{!B_1}, \dots,
  \sFmla{\True}{!B_n}\}$ has a closed !!{tableau}. By
  \olref{thm:tableau-soundness}, every
  \iftag{FOL}{!!{structure}}{!!{valuation}}~$\iftag{FOL}{\Struct{M}}{\pAssign{v}}$
  either makes some $!B_i$ false or makes $!A$ true.  Hence, if
  $\iftag{FOL}{\Sat{M}}{\pSat{v}}{\Gamma}$ then also
  $\iftag{FOL}{\Sat{M}}{\pSat{v}}{!A}$.
\end{proof}

\begin{cor}
\ollabel{cor:consistency-soundness}
If $\Gamma$ is satisfiable, then it is consistent.
\end{cor}

\begin{proof}
We prove the contrapositive.  Suppose that $\Gamma$ is not consistent.
Then there are $!B_1$, \dots, $!B_n \in \Gamma$ and a closed
!!{tableau} for $\{\sFmla{\True}{!B_1}, \dots, \sFmla{\True}{!B_n}\}$.  By
\olref{thm:tableau-soundness}, there is no
$\iftag{FOL}{\Struct{M}}{\pAssign{v}}$ such that
$\iftag{FOL}{\Sat{M}}{\pSat{v}}{!B_i}$ for all $i=1$, \dots,~$n$.  But
then $\Gamma$ is not satisfiable.
\end{proof}

\end{document}
