% Part: first-order-logic
% Chapter: completeness
% Section: compactness

\documentclass[../../../include/open-logic-section]{subfiles}

\begin{document}

\olfileid{fol}{com}{dcp}
\olsection{A Direct Proof of the Compactness Theorem}

We can prove the Compactness Theorem directly, without appealing to
the Completeness Theorem, using the same ideas as in the proof of the
completeness theorem.  In the proof of the Completeness Theorem we
started with a consistent set~$\Gamma$ of !!{sentence}s, expanded it to a
consistent, saturated, and !!{complete} set~$\Gamma^*$ of
!!{sentence}s, and then showed that in the term
model~$\Struct{M(\Gamma^*)}$ constructed from $\Gamma^*$, all
!!{sentence}s of~$\Gamma$ are true, so $\Gamma$ is satisfiable.

We can use the same method to show that a finitely satisfiable set of
sentences is satisfiable. We just have to prove the corresponding
versions of \olref[ccs]{prop:ccs} and \olref[lin]{lem:lindenbaum}
where we replace ``consistent'' with ``finitely satisfiable.''

\begin{thm}
  \ollabel{thm:fsat-henkin}
  Every finitely satisfiable set~$\Gamma$ can be extended to a
  saturated finitely satisfiable set~$\Gamma'$.
\end{thm}


\begin{lem}
  \ollabel{lem:fsat-lindenbaum}
  Every finitely satisfiable set~$\Gamma$ can be extended to a
  !!{complete} and finitely satisfiable, saturated set~$\Gamma^*$.
\end{lem}

\begin{prop}
\ollabel{prop:fsat-ccs}
Suppose $\Gamma$ is !!{complete} and finitely satisfiable. Then:
\begin{enumerate}
\item \ollabel{prop:complete-prov-in} If $\Gamma \Proves !A$, then $!A \in
  \Gamma$.

\tagitem{prvAnd}{\ollabel{prop:complete-and} $(!A \land !B) \in \Gamma$
  iff both $!A \in \Gamma$ and $!B \in \Gamma$.}{}

\tagitem{prvOr}{\ollabel{prop:complete-or} $(!A \lor !B) \in \Gamma$ iff
  either $!A \in \Gamma$ or $!B \in \Gamma$.}{}

\tagitem{prvIf}{\ollabel{prop:complete-if} $(!A \lif !B) \in \Gamma$ iff
  either $!A \notin \Gamma$ or $!B \in \Gamma$.}{}
\end{enumerate}
\end{prop}


\begin{thm}[Compactness Theorem]
\ollabel{thm:compactness}
The following hold for any sentences $\Gamma$ and $!A$:
\begin{enumerate}
  \item $\Gamma \Entails !A$ iff there is a finite $\Gamma_0
    \subseteq \Gamma$ such that $\Gamma_0 \Entails !A$.
  \item $\Gamma$ is satisfiable if and only if it is finitely
    satisfiable.
\end{enumerate}
\end{thm}

\begin{proof}
We prove (2).  If $\Gamma$ is satisfiable, then there is a
!!{structure}~$\Struct{M}$ such that $\Sat{M}{!A}$ for all $!A \in
\Gamma$.  Of course, this $\Struct M$ also satisfies every finite
subset of~$\Gamma$, so $\Gamma$ is finitely satisfiable.

Now suppose that $\Gamma$ is finitely satisfiable.  Then every finite
subset~$\Gamma_0 \subseteq \Gamma$ is satisfiable.  By soundness,
every finite subset is consistent.  Then $\Gamma$ itself must be
consistent.  For assume it is not, i.e., $\Gamma \Proves \bot$.  But
!!{derivation}s are finite, and so already some finite
subset~$\Gamma_0 \subseteq \Gamma$ must be inconsistent
(cf.\ \tagrefs{prfSC/{fol:seq:ptn:prop:proves-compact},prfND/{fol:ntd:ptn:prop:proves-compact}}).
But we just showed they are all consistent, a contradiction.  Now by
completeness, since $\Gamma$ is consistent, it is satisfiable.
\end{proof}

\begin{prob}
Prove (1) of \olref[fol][com][com]{thm:compactness}.
\end{prob}

\begin{ex}
In every model~$\Struct{M}$ of a theory~$\Gamma$, each term~$t$ of
course picks out !!a{element} of~$\Domain{M}$. Can we guarantee that
it is also true that every !!{element} of~$\Domain{M}$ is picked out
by some term or other? In other words, are there theories~$\Gamma$ all
models of which are covered?  The compactness theorem shows that this
is not the case if $\Gamma$ has infinite models. Here's how to see
this: Let $\Struct{M}$ be an infinite model of~$\Gamma$, and let $c$
be !!a{constant} not in the language of~$\Gamma$. Let $\Delta$ be the
set of all sentences $\eq/[c][t]$ for $t$ a term in the
language~$\Lang{L}$ of~$\Gamma$, i.e.,
  \[
  \Delta = \Setabs{\eq/[c][t]}{t \in \Trm[L]}.
  \]
A finite subset of $\Gamma \cup \Delta$ can be written as $\Gamma'
\cup \Delta'$, with $\Gamma' \subseteq \Gamma$ and $\Delta' \subseteq
\Delta$. Since $\Delta'$ is finite, it can contain only finitely many
terms. Let $a \in \Domain{M}$ be !!a{element} of $\Domain{M}$ not
picked out by any of them, and let $\Struct{M'}$ be the !!{structure}
that is just like $\Struct{M}$, but also $\Assign{c}{M'} = a$. Since
$a \neq \Value{t}{M}$ for all~$t$ occuring in~$\Delta'$,
$\Sat{M'}{\Delta'}$. Since $\Sat{M}{\Gamma}$, $\Gamma' \subseteq
\Gamma$, and $c$ does not occur in~$\Gamma$, also
$\Sat{M'}{\Gamma'}$. Together, $\Sat{M'}{\Gamma' \cup \Delta'}$ for
every finite subset $\Gamma' \cup \Delta'$ of $\Gamma \cup \Delta$. So
every finite subset of $\Gamma \cup \Delta$ is satisfiable. By
compactness, $\Gamma \cup \Delta$ itself is satisfiable. So there are
models~$\Sat{M}{\Gamma \cup \Delta}$. Every such $\Struct{M}$ is a
model of~$\Gamma$, but is not covered, since $\Value{c}{M} \neq
\Value{t}{M}$ for all terms~$t$ of~$\Lang{L}$.
\end{ex}

\begin{prob}
Use the compactness theorem to show that any set of sentences in the
language of arithmetic which are true in the standard model of
arithmetic $\Struct{N}$ are also true in !!a{structure}~$\Struct{N'}$
that contains an element greater than all natural
numbers~$\Assign{\num{n}}{N'}$ ($\num{n}$ is
$\Obj{0}^{\prime\dots\prime}$ with $n$ $\prime$'s).
\end{prob}

\begin{ex}
We know that first-order logic with !!{identity} can express that the
size of the !!{domain} must have some minimal size: The
sentence~$!A_{\ge n}$ (which says ``there are at least $n$ distinct
objects'') is true only in structures where $\Domain{M}$ has at
least~$n$ objects. So if we take
  \[
  \Delta = \Setabs{!A_{\ge n}}{n \ge 1}
  \]
then any model of $\Delta$ must be infinite. Thus, we can guarantee
that a theory only has infinite models by adding~$\Delta$ to it: the
models of $\Gamma \cup \Delta$ are all and only the infinite models
of~$\Gamma$.

So first-order logic can express infinitude. The compactness theorem
shows that it cannot express finitude, however. For suppose some set
of sentences $\Lambda$ were satisfied in all and only finite
!!{structure}s. Then $\Delta \cup \Lambda$ is finitely
satisfiable. Why? Suppose $\Delta' \cup \Lambda' \subseteq \Delta \cup
\Lambda$ is finite with $\Delta' \subseteq \Delta$ and $\Lambda'
\subseteq \Lambda$. Let $n$ be the largest number such that $A_{\ge n}
\in \Delta'$. $\Lambda$, being satisfied in all finite structures, has
a model~$\Struct{M}$ with finitely many but $\ge n$ !!{element}s. But
then $\Sat{M}{\Delta' \cup \Lambda'}$.  By compactness, $\Delta \cup
\Lambda$ has an infinite model, contradicting the assumption that
$\Lambda$ is satisfied only in finite !!{structure}s.
\end{ex}

\end{document}
