% Part: first-order-logic
% Chapter: completeness
% Section: complete-sets

% Definition of complete consistent sets. Properties of complete sets required
% for completeness proved are in provability.tex in the
% chapter on the proof system used.

\documentclass[../../../include/open-logic-section]{subfiles}

\begin{document}

\iftag{FOL}
      {\olfileid{fol}{com}{ccs}}
      {\olfileid{pl}{com}{ccs}}
      
\olsection{Complete Consistent Sets of \usetoken{P}{sentence}}

\begin{defn}[Complete set]
\ollabel{def:complete-set} A set~$\Gamma$ of !!{sentence}s is
\emph{!!{complete}} iff for any !!{sentence}~$!A$, either $!A \in
\Gamma$ or $\lnot !A \in \Gamma$.
\end{defn}

\begin{explain}
!!^{complete} sets of sentences leave no questions unanswered. For
any !!{sentence}~$!A$, $\Gamma$ ``says'' if $!A$ is true or false.  The
importance of !!{complete} sets extends beyond the proof of the
completeness theorem. A theory which is !!{complete} and
axiomatizable, for instance, is always decidable.
\end{explain}

\begin{explain}
!!^{complete} consistent sets are important in the completeness proof
since we can guarantee that every consistent set of
!!{sentence}s~$\Gamma$ is contained in !!a{complete} consistent
set~$\Gamma^*$.  !!^a{complete} consistent set contains, for each
!!{sentence}~$!A$, either $!A$ or its negation $\lnot !A$, but not
both. This is true in particular for \iftag{FOL}{atomic
!!{sentence}s}{propositional variables}, so from !!a{complete}
consistent set\iftag{FOL}{ in a language suitably expanded by
!!{constant}s}{}, we can construct
\iftag{FOL}{!!a{structure}}{!!a{valuation}} where the
\iftag{FOL}{interpretation of !!{predicate}s}{truth value assigned to
propositional variables} is defined according to which
\iftag{FOL}{atomic !!{sentence}s}{!!{propositional variable}s} are
in~$\Gamma^*$. This \iftag{FOL}{!!{structure}}{!!{valuation}} can then
be shown to make all !!{sentence}s in~$\Gamma^*$ (and hence also all
those in~$\Gamma$) true. The proof of this latter fact requires that
$\lnot !A \in \Gamma^*$ iff $!A \notin \Gamma^*$, $(!A \lor !B) \in
\Gamma^*$ iff $!A \in \Gamma^*$ or $!B \in \Gamma^*$, etc.
\end{explain}

In what follows, we will often tacitly use the properties of
reflexivity, monotonicity, and transitivity of $\Proves$ (see
\iftag{FOL}{%
  \tagrefs{prfSC/{fol:seq:ptn:sec},prfND/{fol:ntd:ptn:sec},prfAX/{fol:axd:ptn:sec},prfTab/{fol:tab:ptn:sec}}}{%
  \tagrefs{prfSC/{pl:seq:ptn:sec},prfND/{pl:ntd:ptn:sec},prfAX/{pl:axd:ptn:sec},prfTab/{pl:tab:ptn:sec}}}).

\begin{prop}
\ollabel{prop:ccs}
Suppose $\Gamma$ is !!{complete} and consistent. Then:
\begin{enumerate}
\item \ollabel{prop:ccs-prov-in} If $\Gamma \Proves !A$, then $!A \in
  \Gamma$.

\tagitem{prvAnd}{\ollabel{prop:ccs-and} $!A \land !B \in \Gamma$
  iff both $!A \in \Gamma$ and $!B \in \Gamma$.}{}

\tagitem{prvOr}{\ollabel{prop:ccs-or} $!A \lor !B \in \Gamma$ iff
  either $!A \in \Gamma$ or $!B \in \Gamma$.}{}

\tagitem{prvIf}{\ollabel{prop:ccs-if} $!A \lif !B \in \Gamma$ iff
  either $!A \notin \Gamma$ or $!B \in \Gamma$.}{}
\end{enumerate}
\end{prop}

\begin{proof}
Let us suppose for all of the following that $\Gamma$ is !!{complete} and
consistent.
\begin{enumerate}
\item If $\Gamma \Proves !A$, then $!A \in \Gamma$.

Suppose that $\Gamma \Proves !A$. Suppose to the contrary that $!A
\notin \Gamma$. Since $\Gamma$ is !!{complete}, $\lnot !A \in \Gamma$.
By
\iftag{FOL}{%
  \tagrefs{prfSC/{fol:seq:prv:prop:explicit-inc},
    prfND/{fol:ntd:prv:prop:explicit-inc},
    prfAX/{fol:axd:prv:prop:explicit-inc},
    prfTab/{fol:tab:prv:prop:explicit-inc}}}{%
  \tagrefs{prfSC/{pl:seq:prv:prop:explicit-inc},
    prfND/{pl:ntd:prv:prop:explicit-inc},
    prfAX/{pl:axd:prv:prop:explicit-inc},
    prfTab/{pl:tab:prv:prop:explicit-inc}}},
$\Gamma$ is inconsistent. This contradicts the assumption that
$\Gamma$ is consistent. Hence, it cannot be the case that $!A \notin
\Gamma$, so $!A \in \Gamma$.

\tagitem{defAnd}{}{%
\iftag{probAnd}{Exercise.}{%
$!A \land !B \in \Gamma$ iff both $!A \in \Gamma$ and $!B \in \Gamma$:

For the forward direction, suppose $!A \land !B \in \Gamma$. Then
by
\iftag{FOL}{%
  \tagrefs{prfSC/{fol:seq:ppr:prop:provability-land},
    prfND/{fol:ntd:ppr:prop:provability-land},
    prfAX/{fol:axd:ppr:prop:provability-land},
    prfTab/{fol:tab:ppr:prop:provability-land}}}{%
  \tagrefs{prfSC/{pl:seq:ppr:prop:provability-land},
    prfND/{pl:ntd:ppr:prop:provability-land},
    prfAX/{pl:axd:ppr:prop:provability-land},
    prfTab/{pl:tab:ppr:prop:provability-land}}}, item~(1),
$\Gamma \Proves !A$ and $\Gamma \Proves !B$. By
\olref{prop:ccs-prov-in}, $!A \in \Gamma$ and $!B \in \Gamma$, as
required.

For the reverse direction, let $!A \in \Gamma$ and $!B \in
\Gamma$. By
\iftag{FOL}{%
  \tagrefs{prfSC/{fol:seq:ppr:prop:provability-land},
    prfND/{fol:ntd:ppr:prop:provability-land},
    prfAX/{fol:axd:ppr:prop:provability-land},
    prfTab/{fol:tab:ppr:prop:provability-land}}}{%
  \tagrefs{prfSC/{pl:seq:ppr:prop:provability-land},
    prfND/{pl:ntd:ppr:prop:provability-land},
    prfAX/{pl:axd:ppr:prop:provability-land},
    prfTab/{pl:tab:ppr:prop:provability-land}}}, item~(2),
$\Gamma \Proves !A \land !B$. By \olref{prop:ccs-prov-in}, $!A \land
!B \in \Gamma$.}}

\tagitem{defOr}{}{%
\iftag{probOr}{Exercise.}{%
First we show that if $!A \lor !B \in \Gamma$, then either $!A \in
\Gamma$ or $!B \in \Gamma$. Suppose $!A \lor !B \in \Gamma$ but $!A
\notin \Gamma$ and $!B \notin \Gamma$.  Since $\Gamma$ is
!!{complete}, $\lnot !A \in \Gamma$ and $\lnot !B \in \Gamma$. By
\iftag{FOL}{%
  \tagrefs{prfSC/{fol:seq:ppr:prop:provability-lor},
    prfND/{fol:ntd:ppr:prop:provability-lor},
    prfAX/{fol:axd:ppr:prop:provability-lor},
    prfTab/{fol:tab:ppr:prop:provability-lor}}}{%
  \tagrefs{prfSC/{pl:seq:ppr:prop:provability-lor},
    prfND/{pl:ntd:ppr:prop:provability-lor},
    prfAX/{pl:axd:ppr:prop:provability-lor},
    prfTab/{pl:tab:ppr:prop:provability-lor}}},
item (1), $\Gamma$ is inconsistent, a contradiction. Hence, either $!A
\in \Gamma$ or $!B \in \Gamma$.

For the reverse direction, suppose that $!A \in \Gamma$ or $!B \in
\Gamma$. By
\iftag{FOL}{%
  \tagrefs{prfSC/{fol:seq:ppr:prop:provability-lor},
    prfND/{fol:ntd:ppr:prop:provability-lor},
    prfAX/{fol:axd:ppr:prop:provability-lor},
    prfTab/{fol:tab:ppr:prop:provability-lor}}}{%
  \tagrefs{prfSC/{pl:seq:ppr:prop:provability-lor},
    prfND/{pl:ntd:ppr:prop:provability-lor},
    prfAX/{pl:axd:ppr:prop:provability-lor},
    prfTab/{pl:tab:ppr:prop:provability-lor}}}, item (2),
$\Gamma \Proves !A \lor !B$. By \olref{prop:ccs-prov-in}, $!A \lor
!B \in \Gamma$, as required.}}

\tagitem{defIf}{}{%
\iftag{probIf}{Exercise.}{%
For the forward direction, suppose $!A \lif !B \in \Gamma$, and suppose
to the contrary that $!A \in \Gamma$ and $!B \notin \Gamma$. On these
assumptions, $!A \lif !B \in \Gamma$ and $!A \in \Gamma$. By
\iftag{FOL}{%
  \tagrefs{prfSC/{fol:seq:ppr:prop:provability-lif},
    prfND/{fol:ntd:ppr:prop:provability-lif},
    prfAX/{fol:axd:ppr:prop:provability-lif},
    prfTab/{fol:tab:ppr:prop:provability-lif}}}{%
  \tagrefs{prfSC/{pl:seq:ppr:prop:provability-lif},
    prfND/{pl:ntd:ppr:prop:provability-lif},
    prfAX/{pl:axd:ppr:prop:provability-lif},
    prfTab/{pl:tab:ppr:prop:provability-lif}}}, item (1),
$\Gamma \Proves !B$. But then by \olref{prop:ccs-prov-in}, $!B \in
\Gamma$, contradicting the assumption that $!B \notin \Gamma$.

For the reverse direction, first consider the case where $!A \notin
\Gamma$. Since $\Gamma$ is !!{complete}, $\lnot !A \in \Gamma$.  By
\iftag{FOL}{%
  \tagrefs{prfSC/{fol:seq:ppr:prop:provability-lif},
    prfND/{fol:ntd:ppr:prop:provability-lif},
    prfAX/{fol:axd:ppr:prop:provability-lif},
    prfTab/{fol:tab:ppr:prop:provability-lif}}}{%
  \tagrefs{prfSC/{pl:seq:ppr:prop:provability-lif},
    prfND/{pl:ntd:ppr:prop:provability-lif},
    prfAX/{pl:axd:ppr:prop:provability-lif},
    prfTab/{pl:tab:ppr:prop:provability-lif}}}, item (2),
$\Gamma \Proves !A \lif !B$. Again by \olref{prop:ccs-prov-in}, we get
that $!A \lif !B \in \Gamma$, as required.

Now consider the case where $!B \in \Gamma$.  By
\iftag{FOL}{%
  \tagrefs{prfSC/{fol:seq:ppr:prop:provability-lif},
    prfND/{fol:ntd:ppr:prop:provability-lif},
    prfTab/{fol:tab:ppr:prop:provability-lif},
    prfAX/{fol:axd:ppr:prop:provability-lif}}}{%
  \tagrefs{prfSC/{pl:seq:ppr:prop:provability-lif},
    prfND/{pl:ntd:ppr:prop:provability-lif},
    prfTab/{pl:tab:ppr:prop:provability-lif},
    prfAX/{pl:axd:ppr:prop:provability-lif}}},
item (2) again, $\Gamma \Proves !A \lif !B$. By
\olref{prop:ccs-prov-in}, $!A \lif !B \in \Gamma$.}}
\end{enumerate}
\end{proof}

\tagprob{FOL}
\begin{prob}
Complete the proof of \olref[fol][com][ccs]{prop:ccs}.
\end{prob}
\tagendprob

\tagprob{notFOL}
\begin{prob}
Complete the proof of \olref[pl][com][ccs]{prop:ccs}.
\end{prob}
\tagendprob

\end{document}
