% Part: first-order-logic
% Chapter: completeness
% Section: construction-of-model

\documentclass[../../../include/open-logic-section]{subfiles}

\begin{document}

\iftag{FOL}
      {\olfileid{fol}{com}{mod}}
      {\olfileid{pl}{com}{mod}}

\olsection{Construction of a Model}

\begin{explain}
\iftag{FOL}{Right now we are not concerned about $\eq$, i.e., we only
  want to show that a consistent set~$\Gamma$ of !!{sentence}s not
  containing~$\eq$ is satisfiable. We first extend~$\Gamma$ to a
  consistent, !!{complete}, and saturated set~$\Gamma^*$.  In this
  case, the definition of a model~$\Struct{M(\Gamma^*)}$ is simple: We
  take the set of closed terms of~$\Lang{L'}$ as the domain. We assign
  every !!{constant} to itself, and make sure that more generally, for
  every closed term~$t$, $\Value{t}{M(\Gamma^*)} = t$.  The
  !!{predicate}s are assigned extensions in such a way that an atomic
  !!{sentence} is true in $\Struct{M(\Gamma^*)}$ iff it is
  in~$\Gamma^*$.  This will obviously make all the atomic
  !!{sentence}s in~$\Gamma^*$ true in $\Struct{M(\Gamma^*)}$. The rest
  are true provided the $\Gamma^*$ we start with is consistent,
  complete, and saturated.}{We are now ready to define !!a{valuation}
  that makes all $!A \in \Gamma$ true. To do this, we first apply
  Lindenbaum's Lemma: we get a complete consistent $\Gamma^* \supseteq
  \Gamma$. We let the !!{propositional variable}s in~$\Gamma^*$
  determine~$\pAssign v(\Gamma^*)$.}
\end{explain}

\iftag{FOL}{%
\begin{defn}[Term model]
\ollabel{defn:termmodel}
Let $\Gamma^*$ be !!a{complete} and consistent,
saturated set of !!{sentence}s in a language~$\Lang L$. The \emph{term
  model}~$\Struct M(\Gamma^*)$ of $\Gamma^*$ is the !!{structure}
defined as follows:
\begin{enumerate}
\item The !!{domain}~$\Domain{M(\Gamma^*)}$ is the set of all closed
  terms of~$\Lang L$.
\item The interpretation of !!a{constant} $c$ is $c$ itself:
  $\Assign{c}{M(\Gamma^*)} = c$.
\item The !!{function}~$f$ is assigned the function which, given as
  arguments the closed terms $t_1$, \dots, $t_n$, has as value the
  closed term $f(t_1, \dots, t_n)$:
\[
\Assign{f}{M(\Gamma^*)}(t_1, \dots, t_n) = f(t_1,\dots, t_n)
\]
\item If $R$ is an $n$-place !!{predicate}, then
  \[
  \tuple{t_1, \dots,
  t_n} \in \Assign{R}{M(\Gamma^*)} \text{ iff } \Atom{R}{t_1, \dots,
    t_n} \in \Gamma^*.
  \]
\end{enumerate}
\end{defn}
}{%
\begin{defn}
\ollabel{defn:termmodel}
Suppose $\Gamma^*$ is a complete consistent set of !!{formula}s. Then we let
\[
\pAssign v(\Gamma^*)(p) = \begin{cases}
  \True & \text{if $p \in \Gamma^*$}\\
  \False & \text{if $p \notin \Gamma^*$}
  \end{cases}
\]
\end{defn}
}

\iftag{FOL}{%
We will now check that we indeed have $\Value{t}{M(\Gamma^*)} = t$.

\begin{lem} 
\ollabel{lem:val-in-termmodel} Let $\Struct M(\Gamma^*)$ be the term model
of \olref{defn:termmodel}, then $\Value{t}{M(\Gamma^*)} = t$.
\end{lem}
\begin{proof} 
 The proof is by induction on $t$, where the base case, when $t$
 is !!a{constant}, follows directly from the definition of the term
 model. For the induction step assume $t_1, \ldots, t_n$ are closed terms
 such that $\Value{t_i}{M(\Gamma^*)} = t_i$ and that $f$ is an $n$-ary
 !!{function}. Then 
\begin{align*}
\Value{f(t_1,\ldots,t_n)}{M(\Gamma^*)} &= \Assign{f}{M(\Gamma^*)}(\Value{t_1}
 {M(\Gamma^*)}, 
\ldots, \Value{t_n}{M(\Gamma^*)}) \\
&= \Assign{f}{M(\Gamma^*)}(t_1, \dots, t_n) \\
&= f(t_1,\dots, t_n),
\end{align*} 
and so by induction this holds for every closed term~$t$.
\end{proof}
}

\iftag{FOL}{%
\begin{explain}
  \iftag{prvEx}{!!^a{structure}~$\Struct{M}$ may make an existentially
    quantified !!{sentence}~$\lexists[x][!A(x)]$ true without there
    being an instance~$!A(t)$ that it makes true.}{}
  \iftag{prvAll}{!!^a{structure}~$\Struct{M}$ may make all
    instances~$!A(t)$ of a universally quantified
    !!{sentence}~$\lforall[x][!A(x)]$ true, without
    making~$\lforall[x][!A(x)]$ true.}{} This is because in general
  not every !!{element} of~$\Domain{M}$ is the value of a closed term
  ($\Struct{M}$ may not be covered).  This is the reason the
  satisfaction relation is defined via variable assignments. However,
  for our term model~$\Struct{M(\Gamma^*)}$ this wouldn't be
  necessary---because it is covered. This is the content of the next
  result.
\end{explain}

\begin{prop}
\ollabel{prop:quant-termmodel}
Let $\Struct M(\Gamma^*)$ be the term model of \olref{defn:termmodel}.
\begin{tagenumerate}{prvEx,prvAll}
\tagitem{prvEx}{$\Sat{M(\Gamma^*)}{\lexists[x][!A(x)]}$ iff
  $\Sat{M(\Gamma^*)}{!A(t)}$ for at least one closed term~$t$.}{}
\tagitem{prvAll}{$\Sat{M(\Gamma^*)}{\lforall[x][!A(x)]}$ iff
  $\Sat{M(\Gamma^*)}{!A(t)}$ for all closed terms~$t$.}{}
\end{tagenumerate}
\end{prop}

\begin{proof}
\begin{tagenumerate}{prvEx,prvAll}
\tagitem{prvEx}{%
  \iftag{probEx}{Exercise.}{By \olref[syn][ass]{prop:sat-quant},
    $\Sat{M(\Gamma^*)}{\lexists[x][!A(x)]}$ iff for at least one
    variable assignment~$s$, $\Sat{M(\Gamma^*)}{!A(x)}[s]$. As
    $\Domain{M(\Gamma^*)}$ consists of the closed terms of~$\Lang{L}$,
    this is the case iff there is at least one closed term~$t$ such
    that $s(x) = t$ and $\Sat{M(\Gamma^*)}{!A(x)}[s]$.  By
    \olref[fol][syn][ext]{prop:ext-formulas},
    $\Sat{M(\Gamma^*)}{!A(x)}[s]$ iff $\Sat{M(\Gamma^*)}{!A(t)}[s]$,
    where $s(x) = t$.  By \olref[fol][syn][ass]{prop:sentence-sat-true},
    $\Sat{M(\Gamma^*)}{!A(t)}[s]$ iff $\Sat{M(\Gamma^*)}{!A(t)}$,
    since $!A(t)$ is a sentence.}}{}
\tagitem{prvAll}{%
  \iftag{probAll}{Exercise.}{By \olref[syn][ass]{prop:sat-quant},
    $\Sat{M(\Gamma^*)}{\lforall[x][!A(x)]}$ iff for every variable
    assignment $s$, $\Sat{M(\Gamma^*)}{!A(x)}[s]$. Recall that
    $\Domain{M(\Gamma^*)}$ consists of the closed terms of~$\Lang{L}$,
    so for every closed term~$t$, $s(x) = t$ is such a variable
    assignment, and for any variable assignment, $s(x)$ is some closed
    term~$t$.  By \olref[fol][syn][ext]{prop:ext-formulas},
    $\Sat{M(\Gamma^*)}{!A(x)}[s]$ iff $\Sat{M(\Gamma^*)}{!A(t)}[s]$,
    where $s(x) = t$.  By \olref[fol][syn][ass]{prop:sentence-sat-true},
    $\Sat{M(\Gamma^*)}{!A(t)}[s]$ iff $\Sat{M(\Gamma^*)}{!A(t)}$,
    since $!A(t)$ is a sentence.}}{}
\end{tagenumerate}
\end{proof}
}{}

\tagprob[FOL]{probEx,probAll}
\begin{prob}
Complete the proof of \olref[fol][com][mod]{prop:quant-termmodel}.
\end{prob}
\tagendprob

\begin{lem}[Truth Lemma]
\ollabel{lem:truth} \iftag{FOL}{Suppose $!A$ does not contain~$\eq$. Then}{}
$\iftag{FOL}{\Sat{M(\Gamma^*)}}{\pSat{v(\Gamma^*)}}{!A}$ iff $!A \in \Gamma^*$.
\end{lem}

\begin{proof}
We prove both directions simultaneously, and by induction on $!A$.
\begin{enumerate}
\tagitem{prvFalse}{\indcase{!A}{\lfalse}
  {$\iftag{FOL}{\Sat/{M(\Gamma^*)}{\lfalse}}{\pSat/{v(\Gamma^*)}{\lfalse}}$
    by definition of satisfaction. On the other hand, $\lfalse \notin
    \Gamma^*$ since $\Gamma^*$ is consistent.}}{}

\tagitem{prvTrue}{\indcase{!A}{\ltrue}
  {$\iftag{FOL}{\Sat{M(\Gamma^*)}}{\pSat{v(\Gamma^*)}}{\ltrue}$
    by definition of satisfaction. On the other hand, $\ltrue \in
    \Gamma^*$ since $\Gamma^*$ is consistent and !!{complete}, and
    $\Gamma^* \Proves \ltrue$.}}{}

\tagitem{FOL}{\indcase{!A}{R(t_1, \dots, t_n)}
  {$\Sat{M(\Gamma^*)}{\Atom{R}{t_1, \dots, t_n}}$ iff $\tuple{t_1,
      \dots, t_n} \in \Assign{R}{M(\Gamma^*)}$ (by the definition of
    satisfaction) iff $R(t_1, \dots, t_n) \in \Gamma^*$ (by the
    construction of $\Struct
    M(\Gamma^*)$).}}{\indcase{!A}{p}{$\pSat{v(\Gamma^*)}{p}$ iff
    $\pAssign v(\Gamma^*)(p) = \True$ (by the definition of
    satisfaction) iff $p \in \Gamma^*$ (by the construction of
    $\pAssign v(\Gamma^*)$).}}

\tagitem{prvNot}{%
  \indcase{!A}{\lnot !B}
    {$\iftag{FOL}{\Sat{M(\Gamma^*)}}{\pSat{v(\Gamma^*)}}{\indfrm}$ iff
    $\iftag{FOL}{\Sat/{M(\Gamma^*)}{!B}}{\pSat/{v(\Gamma^*)}{!B}}$ (by
    definition of satisfaction). By induction hypothesis,
    $\iftag{FOL}{\Sat/{M(\Gamma^*)}{!B}}{\pSat/{v(\Gamma^*)}{!B}}$ iff
    $!B \notin \Gamma^*$. Since $\Gamma^*$ is consistent and
    !!{complete}, $!B
    \notin \Gamma^*$ iff $\lnot !B \in \Gamma^*$.}}{}

\tagitem{prvAnd}{%
\iftag{probAnd}{%
  \indcase!{!A}{!B \land !C}{}}{%
  \indcase{!A}{!B \land
    !C}{$\iftag{FOL}{\Sat{M(\Gamma^*)}}{\pSat{v(\Gamma^*)}}{\indfrm}$
    iff we have both
    $\iftag{FOL}{\Sat{M(\Gamma^*)}}{\pSat{v(\Gamma^*)}}{!B}$ and
    $\iftag{FOL}{\Sat{M(\Gamma^*)}}{\pSat{v(\Gamma^*)}}{!C}$ (by
    definition of satisfaction) iff both $!B \in \Gamma^*$ and $!C \in
    \Gamma^*$ (by the induction hypothesis). By
    \olref[ccs]{prop:ccs}\olref[ccs]{prop:ccs-and}, this is the case
    iff $(!B \land !C) \in \Gamma^*$.}}}{}

\tagitem{prvOr}{%
\iftag{probOr}{%
  \indcase!{!A}{!B \lor !C}{}}{%
  \indcase{!A}{!B \lor !C}
    {$\iftag{FOL}{\Sat{M(\Gamma^*)}}{\pSat{v(\Gamma^*)}}{\indfrm}$
    iff $\iftag{FOL}{\Sat{M(\Gamma^*)}}{\pSat{v(\Gamma^*)}}{!B}$ or
    $\iftag{FOL}{\Sat{M(\Gamma^*)}}{\pSat{v(\Gamma^*)}}{!C}$ (by
    definition of satisfaction) iff $!B \in \Gamma^*$ or $!C \in
    \Gamma^*$ (by induction hypothesis). This is the case iff $(!B
    \lor !C) \in \Gamma^*$ (by
    \olref[ccs]{prop:ccs}\olref[ccs]{prop:ccs-or}).}}}{}
  
\tagitem{prvIf}{%
\iftag{probIf}{%
  \indcase!{!A}{!B \lif !C}{}}{%
  \indcase{!A}{!B \lif !C}
    {$\iftag{FOL}{\Sat{M(\Gamma^*)}}{\pSat{v(\Gamma^*)}}{\indfrm}$
    iff $\iftag{FOL}{\Sat/{M(\Gamma^*)}}{\pSat/{v(\Gamma^*)}}{!B}$ or $\iftag{FOL}{\Sat{M (\Gamma^*)}}{\pSat{v(\Gamma^*)}}{!C}$ (by
    definition of satisfaction) iff $!B \notin \Gamma^*$ or $!C \in
    \Gamma^*$ (by induction hypothesis). This is the case iff $(!B
    \lif !C) \in \Gamma^*$ (by
    \olref[ccs]{prop:ccs}\olref[ccs]{prop:ccs-if}).}}}{}

\iftag{FOL}{%
\tagitem{prvAll}{%
  \iftag{probAll}{%
    \indcase!{!A}{\lforall[x][!B(x)]}{}}{%
    \indcase{!A}{\lforall[x][!B(x)]}{$\Sat{M(\Gamma^*)}{\indfrm}$ iff
      $\Sat{M(\Gamma^*)}{!B(t)}$ for all terms~$t$
      (\olref{prop:quant-termmodel}).  By induction hypothesis, this
      is the case iff $!B(t) \in \Gamma^*$ for all terms~$t$, by
      \olref[hen]{prop:saturated-instances}, this in turn is the case
      iff $\lforall[x][!A(x)] \in \Gamma^*$.}}}{}

\tagitem{prvEx}{%
  \iftag{probEx}{%
    \indcase!{!A}{\lexists[x][!B(x)]}{}}{%
    \indcase{!A}{\lexists[x][!B(x)]}{$\Sat{M(\Gamma^*)}{\indfrm}$ iff
      $\Sat{M(\Gamma^*)}{!B(t)}$ for at least one term~$t$
      (\olref{prop:quant-termmodel}).  By induction hypothesis, this
      is the case iff $!B(t) \in \Gamma^*$ for at least one term~$t$.
      By \olref[hen]{prop:saturated-instances}, this in turn is the
      case iff $\lexists[x][!B(x)] \in \Gamma^*$.}}}{}
}{}
\end{enumerate}
\end{proof}


\tagprob[FOL]{probOr,probAnd,probIf,probEx,probAll}

\begin{prob}
Complete the proof of \olref[fol][com][mod]{lem:truth}.
\end{prob}

\tagendprob

\tagprob[notFOL]{probOr,probAnd,probIf}

\begin{prob}
Complete the proof of \olref[pl][com][mod]{lem:truth}.
\end{prob}

\tagendprob

\end{document}
