% Part: first-order-logic
% Chapter: axiomatic-deduction
% Section: soundness

\documentclass[../../../include/open-logic-section]{subfiles}

\begin{document}

\iftag{FOL}
      {\olfileid{fol}{axd}{sou}}
      {\olfileid{pl}{axd}{sou}}
      
\olsection{Soundness}

\begin{explain}
!!^a{derivation} system, such as axiomatic deduction, is \emph{sound}
if it cannot !!{derive} things that do not actually hold.  Soundness is
thus a kind of guaranteed safety property for !!{derivation} systems.
Depending on which proof theoretic property is in question, we would
like to know for instance, that
\begin{enumerate}
\item every !!{derivable}~$!A$ is valid;
\item if $!A$ is !!{derivable} from some others~$\Gamma$, it is also a
  consequence of them;
\item if a set of !!{formula}s~$\Gamma$ is inconsistent, it is
  unsatisfiable.
\end{enumerate}
These are important properties of !!a{derivation} system.  If any of them do
not hold, the !!{derivation} system is deficient---it would !!{derive} too much.
Consequently, establishing the soundness of !!a{derivation} system is of the
utmost importance.
\end{explain}

\begin{prop}
  If $!A$ is an axiom, then
  \iftag{FOL}
    {$\Sat{M}{!A}[s]$ for each !!{structure}~$\Struct{M}$ and assignment~$s$.}
    {$\pSat{v}{!A}$ for each !!{valuation}~$\pAssign{v}$.}
\end{prop}

\begin{proof}
\iftag{FOL}{We have to verify that all the axioms are valid. For
  instance, here is the case for \olref[qua]{ax:q1}: suppose $t$ is
  !!{free for} $x$ in $!A$, and assume
  $\Sat{M}{\lforall[x][!A]}[s]$. Then by definition of satisfaction,
  for each $\varAssign{s'}{s}{x}$, also $\Sat{M}{!A}[s']$, and in particular
  this holds when $s'(x) = \Value{t}{M}[s]$. By
  \olref[syn][ext]{prop:ext-formulas},
  $\Sat{M}{\Subst{!A}{t}{x}}[s]$. This shows that
  $\Sat{M}{(\lforall[x][!A] \lif \Subst{!A}{t}{x})}[s]$.}{Do truth
  tables for each axiom to verify that they are tautologies.}
\end{proof}

\begin{thm}[Soundness]
\ollabel{thm:soundness}
If $\Gamma \Proves !A$ then $\Gamma \Entails !A$.
\end{thm}

\begin{proof}
By induction on the length of the !!{derivation} of $!A$ from
$\Gamma$. If there are no steps justified by inferences, then all
!!{formula}s in the derivation are either instances of axioms or are
in~$\Gamma$. By the previous proposition, all the axioms are
\iftag{FOL}{valid}{tautologies}, and hence if $!A$ is an axiom then
$\Gamma \Entails !A$. If $!A \in \Gamma$, then trivially $\Gamma
\Entails !A$.

If the last step of the derivation of~$!A$ is justified by modus
ponens, then there are !!{formula}s $!B$ and $!B \lif !A$ in the
!!{derivation}, and the induction hypothesis applies to the part of
the !!{derivation} ending in those !!{formula}s (since they contain at
least one fewer steps justified by an inference).  So, by induction
hypothesis, $\Gamma \Entails !B$ and $\Gamma \Entails !B \lif
!A$. Then $\Gamma \Entails !A$ by
\iftag{FOL}
      {\olref[syn][sem]{thm:sem-deduction}}
      {\olref[pl][syn][sem]{thm:sem-deduction}}.

\iftag{FOL}{Now suppose the last step is justified by
\QR. Then that step has the form $!C \lif \lforall[x][B(x)]$ and
there is a preceding step $!C \lif !B(c)$ with $c$ not in $\Gamma$,
$!C$, or $\lforall[x][B(x)]$. By induction hypothesis, $\Gamma
\Entails !C \lif \lforall[x][B(x)]$. By
\olref[syn][sem]{thm:sem-deduction}, $\Gamma \cup \{!C\} \Entails
!B(c)$.

Consider some structure~$\Struct{M}$ such that $\Sat{M}{\Gamma \cup
  \{!C\}}$.  We need to show that $\Sat{M}{\lforall[x][!B(x)]}$. Since
$\lforall[x][!B(x)]$ is !!a{sentence}, this means we have to show that
for every variable assignment~$s$, $\Sat{M}{!B(x)}[s]$
(\olref[syn][ass]{prop:sat-quant}). Since $\Gamma \cup \{!C\}$
consists entirely of sentences, $\Sat{M}{!D}[s]$ for all $!D \in
\Gamma$ by \olref[syn][sat]{defn:satisfaction}.  Let $\Struct{M'}$ be
like $\Struct{M}$ except that $\Assign{c}{M'} = s(x)$.  Since $c$ does
not occur in~$\Gamma$ or~$!C$, $\Sat{M'}{\Gamma \cup \{!C\}}$ by
\olref[syn][ext]{cor:extensionality-sent}. Since $\Gamma \cup \{!C\}
\Entails !B(c)$, $\Sat{M'}{B(c)}$.  Since $!B(c)$ is !!a{sentence},
$\Sat{M}{!B(c)}[s]$ by
\olref[syn][ass]{prop:sentence-sat-true}. $\Sat{M'}{!B(x)}[s]$ iff
$\Sat{M'}{!B(c)}$ by \olref[syn][ext]{prop:ext-formulas} (recall that
$!B(c)$ is just $\Subst{!B(x)}{c}{x}$). So,
$\Sat{M'}{!B(x)}[s]$. Since $c$ does not occur in~$!B(x)$, by
\olref[syn][ext]{prop:extensionality}, $\Sat{M}{!B(x)}[s]$. But $s$
was an arbitrary variable assignment, so
$\Sat{M}{\lforall[x][!B(x)]}$. Thus $\Gamma \cup \{!C\} \Entails
\lforall[x][!B(x)]$. By \olref[syn][sem]{thm:sem-deduction}, $\Gamma
\Entails !C \lif \lforall[x][!B(x)]$.

The case where $!A$ is justified by \QR{} but is of the form
$\lexists[x][!B(x)] \lif !C$ is left as an exercise.}{}
\end{proof}

\tagprob{FOL}
\begin{prob}
  Complete the proof of \olref[fol][axd][sou]{thm:soundness}.
\end{prob}
\tagendprob

\begin{cor}
\ollabel{cor:weak-soundness}
If $\Proves !A$, then $!A$ is \iftag{FOL}{valid}{a tautology}.
\end{cor}

\begin{cor}
\ollabel{cor:consistency-soundness}
If $\Gamma$ is satisfiable, then it is consistent.
\end{cor}

\begin{proof}
We prove the contrapositive.  Suppose that $\Gamma$ is not consistent.
Then $\Gamma \Proves \lfalse$, i.e., there is !!a{derivation} of
$\lfalse$ from~$\Gamma$. By \olref{thm:soundness}, any
\iftag{FOL}{!!{structure}~$\Struct{M}$}{!!{valuation}~$\pAssign{v}$}
that satisfies $\Gamma$ must satisfy~$\lfalse$.  Since
\iftag{FOL}{$\Sat/{M}{\lfalse}$ for every
  !!{structure}~$\Struct{M}$}{$\pSat/{v}{\lfalse}$ for every
  !!{valuation}~$\pAssign{v}$}, no
\iftag{FOL}{$\Struct{M}$}{$\pAssign{v}$} can satisfy $\Gamma$, i.e.,
$\Gamma$ is not satisfiable.
\end{proof}


\end{document}
