% Part: sets-functions-relations
% Chapter: functions
% Section: basics

\documentclass[../../include/open-logic-section]{subfiles}


\begin{document}

\olfileid{sfr}{fun}{bas}
\olsection{Basics}

\begin{explain}
A function is a relation in which each object is related to a unique partner. Many functions are familiar to us from basic arithmetic. For instance, addition and multiplication are functions. They take in two numbers and return a third. A function, more generally, is something that takes one or more things as input and returns some kind of output. A function is a \emph{black box}: what matters is only what output is paired with what input, not the method for calculating the output. 
\end{explain}

\begin{defn}
A \emph{function} $f : X \rightarrow Y$ goes
    from a set $X$ to a set $Y$. We call $X$ the \emph{domain} of $f$ and $Y$ the \emph{codomain} of $f$. The \emph{range} of $f$ is the subset of the codomain that is actually output by $f$ for some input.
\end{defn}

\begin{ex}
Multiplication goes from $\mathbb{N} \times \mathbb{N}$ (the domain) to $\mathbb{N}$ (the codomain). As it turns out, the range is also $\mathbb{N}$.
\end{ex}

\begin{explain}
Multiplication is a function because it pairs each input---each pair of natural numbers---with a single output. In contrast, the square root operation applied to the domain $\mathbb{N}$ is not functional, since each positive integer $n$ has two square roots: $\sqrt{n}$ and $-\sqrt{n}$. The relation that pairs each student in a class with their final grade is a function---no student can get two different final grades in the same class. The relation that pairs each student in a class with their parents is not a function---generally each student will have at least two parents.
\end{explain}

\begin{ex}
Let $f: \mathbb{N} \rightarrow \mathbb{N}$ be defined such that $f(x) = x+1$. This tells us that $f$ is a function which takes in natural numbers and outputs natural numbers. It then tells us that, given a natural number, $f$ will output its successor.
\end{ex}

\begin{explain}
In this case, the codomain $\mathbb{N}$ is not the range of $f$, since the natural number $0$ is not the successor of any other natural number. The range is the set of all positive integers, $\mathbb{Z}^{+}$.
\end{explain}

\begin{ex}
Let $g: \mathbb{N} \rightarrow \mathbb{N}$ be defined such that $g(x) = x-1+2$. This tells us that $g$ is a function which takes in natural numbers and outputs natural numbers. It then tells us that, given a natural number, $g$ will output the successor of the successor of its predecessor. Despite their different definitions, $g$ and $f$ are the same function.
\end{ex}

\begin{explain}
Functions $f$ and $g$ defined above are the same because for any natural number $x$, $x-1+2 = x+1$. $f$ and $g$ pair each natural number with the same output. Functions, just like relations more generally, can be treated as just sets of pairs. The definitions for $f$ and $g$ specify the same set by means of different equations, and as we know, sets are independent of how they are specified.
\end{explain}

\begin{ex}
We can also define functions by cases. For instance, we could define
    $f: \mathbb{N} \rightarrow \mathbb{N}$  by 
\[
f(x) = \left\{
                                                \begin{array}{ll}
                                                  \frac{x}{2}, & \hbox{if $x$ is even;} \\
                                                  \frac{x+1}{2}, & \hbox{if $x$ is odd.}
                                                \end{array}
                                              \right.
\]
This is fine, since every natural number is either even or odd, and the output of this function will always be a natural number. Just remember that if you define a function by cases, every possible input must fall into exactly one case.
\end{ex}

\end{document}
