% Part: propositional-logic
% Chapter: syntax-and-semantics
% Section: semantics-properties

\documentclass[../../../include/open-logic-section]{subfiles}

\begin{document}

\olfileid{pl}{syn}{smp}

\olsection{Properties of validity}

\begin{prop}
\ollabel{prop:semanticalfacts} 
\begin{enumerate} 
  \item $!A$ is a tautology if and only if $\emptyset \Entails !A$; 
  \item $\Gamma \Entails !A$ if and only if $\Gamma \cup \{\lnot !A\}$ is unsatisfiable; 
  \item If $\Gamma$ is satisfiable then every finite subset of $\Gamma$
  is also satisfiable.
\end{enumerate}
\end{prop}

\begin{proof}
Exercise.
\end{proof}

\begin{prob}
Prove \olref[pl][syn][smp]{prop:semanticalfacts}
\end{prob}

\begin{prop}
\ollabel{prop:structural}
\begin{enumerate}
  \item \ollabel{def:Reflexivity} \emph{Reflexivity}. $\Gamma, !A \Entails !A$. 
  \item \ollabel{def:Monotonicity} \emph{Monotonicity}. If $\Gamma \Entails !A$ and $\Gamma \subseteq \Delta$ then also $\Delta \Entails !A$;
  \item \ollabel{def:Cut} \emph{Transitivity} (or Semantic Cut). If $\Gamma \Entails !A$ and $\Delta, !A \Entails !B$ then $\Gamma \cup \Delta \Entails !B$.
\end{enumerate}
\end{prop}

\begin{proof}
Exercise.
\end{proof}

\begin{prob}
Prove \olref[pl][syn][smp]{prop:structural}
\end{prob}

\begin{prop}[Local Determination]
\ollabel{prop:localdetermination}
\emph{Local determination}. If two !!{valuation}s $\pAssign{v}$ and $\pAssign{v'}$ agree on every !!{propositional variable} in $!A$ then they agree on $!A$. That is, if $\pValue{v}(!B)=\pValue{v}(!B)$ for very !!{propositional variable} $!B$ that appears in $!A$ then $\pValue{v}(!B)=\pValue{v}(!B)$.
\end{prop}

\begin{proof}
Exercise.
\end{proof}

\begin{prob}
Prove \olref[pl][syn][smp]{prop:localdetermination}
\end{prob}

\begin{prop}
\ollabel{prop:bivalence} \emph{Bivalence}. For every !!{valuation}s $\pAssign{v}$ and every formula $!A$, $\pValue{v}(!A)=\True$ or $\pValue{v}(!A)=\False$ but not both.
\end{prop}

\begin{proof}
Exercise.
\end{proof}

\begin{prob}
Prove \olref[pl][syn][smp]{prop:localdetermination}
\end{prob}

\begin{prop}
\ollabel{prop:substitutionrules}
\begin{enumerate}
  \item \emph{Rule of tautological instance}. If $\Entails !A$ then also $\Entails \Subst{!A}{!C}{!B}$ for any $!B$, $!C$. That is, if $\Entails !A$ then for every !!{propositional variable} $!B$ and formula $!C$, the result of replacing every occurrence of $!B$ in $!A$ with $!C$, that is $\Subst{!A}{!C}{!B}$ is a tautology too.
  \item \emph{Rule of substitution of logical equivalents}. If $!A$ and $!A$ are logical equivalents then for any !!{valuation} $\pAssign{v}$ and any !!{formula} $!C$, $\pAssign{v}{!C}=\pAssign{v}{\Subst{!C}{!B}{!A}}$
\end{enumerate}
\end{prop}

\begin{proof}
Exercise.
\end{proof}

\begin{prob}
Prove \olref[pl][syn][smp]{prop:substitutionrules}
\end{prob}

\paragraph*{Validity and the connectives}

\begin{thm}[Semantic Deduction Theorem]
  \ollabel{thm:sem-deduction} $\Gamma \Entails !A \lif !B$ if and only
  if $\Gamma, !A \Entails !B$.
\end{thm}

\begin{proof}
Exercise.
\end{proof}

\begin{prob}
Prove \olref[pl][syn][smp]{thm:sem-deduction}
\end{prob}

\end{document}
