% Part: computability
% Chapter: machines-computations
% Section: halting-problem

\documentclass[../../../include/open-logic-section]{subfiles}

\begin{document}

\olfileid{tur}{und}{hal}
\olsection{The Halting Problem}

\begin{explain}
Assume we have fixed some enumeration of Turing machine
descriptions.  Each Turing machine thus receives an \emph{index}: its
place in the enumeration $M_1$, $M_2$, $M_3$, \dots{} of Turing machine
descriptions.

We know that there must be non-Turing-computable functions: the set of
Turing machine descriptions---and hence the set of Turing machines---is
!!{enumerable}, but the set of all functions from $\Nat$ to $\Nat$ is
not. But we can find specific examples of non-computable functions as
well. One such function is the halting function.
\end{explain}

\begin{defn}[Halting function]
 The \emph{halting function}~$h$ is defined as
\[
h(e,n) =
\begin{cases}
  \text{0} & \text{if machine~$M_e$ does not halt for input $n$} \\
  \text{1} & \text{if machine~$M_e$ halts for input $n$}
\end{cases}
\]
\end{defn}

\begin{defn}[Halting problem]
The \emph{Halting Problem} is the problem of determining (for any $e$, $n$)
whether the Turing machine~$M_e$ halts for an input of~$n$ strokes.
\end{defn}

\begin{explain}
We show that~$h$ is not Turing-computable by showing that a related
function,~$s$, is not Turing-computable. This proof relies on the fact
that anything that can be computed by a Turing machine can be computed
by a disciplined Turing machine (\olref[mac][dis]{sec}), and the fact
that two Turing machines can be hooked together to create a single
machine (\olref[mac][cmb]{sec}).
\end{explain}

\begin{defn} The function~$s$ is defined as
\[
s(e) =
\begin{cases}
  \text{0} & \text{if machine~$M_e$ does not halt for input $e$} \\
  \text{1} & \text{if machine~$M_e$ halts for input $e$}
\end{cases}
\]
\end{defn}

\begin{lem}
The function~$s$ is not Turing computable.
\end{lem}

\begin{proof}
We suppose, for contradiction, that the function~$s$ is Turing
computable.  Then there would be a Turing machine~$S$ that
computes~$s$. We may assume, without loss of generality, that when $S$
halts, it does so while scanning the first square (i.e., that it is
disciplined).  This machine can be ``hooked up'' to another
machine~$J$, which halts if it is started on input~$0$ (i.e., if it
reads $\TMblank$ in the initial state while scanning the square to the
right of the end-of-tape symbol), and otherwise wanders off to the
right, never halting. $S \concat J$, the machine created by hooking
$S$ to~$J$, is a Turing machine, so it is $M_e$ for some~$e$ (i.e., it
appears somewhere in the enumeration). Start $M_e$ on an input of~$e$
$\TMstroke$s. There are two possibilities: either $M_e$ halts or it
does not halt.
\begin{enumerate}
\item Suppose $M_e$ halts for an input of $e$ $\TMstroke$s. Then $s(e)
  = 1$. So $S$, when started on~$e$, halts with a single $\TMstroke$
  as output on the tape.  Then $J$ starts with a $\TMstroke$ on the
  tape. In that case $J$ does not halt. But $M_e$ is the machine $S
  \concat J$, so it should do exactly what $S$ followed by $J$ would
  do (i.e., in this case, wander off to the right and never halt).  So
  $M_e$ cannot halt for an input of $e$ $\TMstroke$'s.

\item Now suppose $M_e$ does not halt for an input of $e$
  $\TMstroke$s.  Then $s(e) = 0$, and $S$, when started on input~$e$,
  halts with a blank tape.  $J$,~when started on a blank tape,
  immediately halts.  Again, $M_e$ does what $S$ followed by~$J$ would
  do, so $M_e$ must halt for an input of $e$ $\TMstroke$'s.
\end{enumerate}
In each case we arrive at a contradiction with our assumption. This
shows there cannot be a Turing machine~$S$: $s$~is not Turing
computable.
\end{proof}

\begin{thm}[Unsolvability of the Halting Problem]
\ollabel{thm:halting-problem} The halting problem is unsolvable, i.e.,
the function~$h$ is not Turing computable.
\end{thm}

\begin{proof}
Suppose $h$ were Turing computable, say, by a Turing machine~$H$. We
could use $H$ to build a Turing machine that computes~$s$: First, make
a copy of the input (separated by a~$\TMblank$ symbol). Then move back
to the beginning, and run~$H$.  We can clearly make a machine that
does the former (see \cref{tur:mac:dis:prob:copier}), and if $H$
existed, we would be able to ``hook it up'' to such a copier machine
to get a new machine which would determine if $M_e$ halts on
input~$e$, i.e., computes~$s$. But we've already shown that no such
machine can exist. Hence, $h$~is also not Turing computable.
\end{proof}

\begin{prob}
The Three Halting (3-Halt) problem is the problem of giving a decision
procedure to determine whether or not an arbitrarily chosen Turing
Machine halts for an input of three $\TMstroke$'s on an otherwise
blank tape. Prove that the 3-Halt problem is unsolvable.
\end{prob}

\begin{prob}
Show that if the halting problem is solvable for Turing machine and
input pairs $M_e$ and~$n$ where $e \neq n$, then it is also solvable
for the cases where $e = n$.
\end{prob}

\begin{prob}
We proved that the halting problem is unsolvable if the input is a
number~$e$, which identifies a Turing machine~$M_e$ via an enumeration
of all Turing machines.  What if we allow the description of Turing
machines from \olref[tur][und][enu]{sec} directly as input?  Can there
be a Turing machine which decides the halting problem but takes as
input descriptions of Turing machines rather than indices?  Explain
why or why not.
\end{prob}

\begin{prob} Show that the \emph{partial} function~$s'$ is defined as
  \[
  s'(e) =
  \begin{cases}
    \text{1} & \text{if machine~$M_e$ halts for input $e$}\\
    \text{undefined} & \text{if machine~$M_e$ does not halt for input $e$}
  \end{cases}
  \]
  \emph{is} Turing computable.
\end{prob}

\end{document}
