% Part: first-order-logic
% Chapter: sequent-calculus
% Section: soundness.tex

\documentclass[../../../include/open-logic-section]{subfiles}

\begin{document}

\iftag{FOL}
      {\olfileid{fol}{seq}{sou}}
      {\olfileid{pl}{seq}{sou}}

\olsection{Soundness}

\begin{explain}
!!^a{derivation} system, such as the sequent calculus, is \emph{sound}
if it cannot !!{derive} things that do not actually hold.  Soundness is
thus a kind of guaranteed safety property for !!{derivation} systems.
Depending on which proof theoretic property is in question, we would
like to know for instance, that
\begin{enumerate}
\item every !!{derivable}~$!A$ is \iftag{FOL}{valid}{a tautology};
\item if !!a{sentence} is !!{derivable} from some others, it is also a
  consequence of them;
\item if a set of !!{sentence}s is inconsistent, it is unsatisfiable.
\end{enumerate}
These are important properties of !!a{derivation} system.  If any of
them do not hold, the !!{derivation} system is deficient---it would
!!{derive} too much.  Consequently, establishing the soundness of a
!!{derivation} system is of the utmost importance.

Because all these proof-theoretic properties are defined via
!!{derivability} in the sequent calculus of certain sequents, proving
(1)--(3) above requires proving something about the semantic
properties of !!{derivable} sequents.  We will first define what it
means for a sequent to be \emph{valid}, and then show that every
!!{derivable} sequent is valid.  (1)--(3) then follow as corollaries
from this result.
\end{explain}

\begin{defn}
\iftag{FOL}{!!^a{structure}~$\Struct
  M$}{!!^a{valuation}~$\pAssign{v}$} \emph{satisfies} a sequent
$\Gamma \Sequent \Delta$ iff either
$\iftag{FOL}{\Sat/{M}}{\pSat/{v}}{!A}$ for some $!A \in \Gamma$ or
$\iftag{FOL}{\Sat{M}}{\pSat{v}}{!A}$ for some $!A \in \Delta$.

A sequent is \emph{valid} iff every \iftag{FOL}{!!{structure}~$\Struct
  M$}{!!{valuation}~$\pAssign{v}$} satisfies it.
\end{defn}

\begin{thm}[Soundness]
\ollabel{thm:sequent-soundness} If $\Log{LK}$ !!{derive}s $\Theta
\Sequent \Xi$, then $\Theta \Sequent \Xi$ is valid.
\end{thm}

\begin{proof}
Let $\pi$ be !!a{derivation} of $\Theta \Sequent \Xi$. We proceed by
induction on the number of inferences~$n$ in~$\pi$.

If the number of inferences is~$0$, then $\pi$ consists only of an
initial sequent. Every initial sequent $!A \Sequent !A$ is obviously
valid, since for every \iftag{FOL}{$\Struct M$}{$\pAssign{v}$}, either
$\iftag{FOL}{\Sat/{M}}{\pSat/{v}}{!A}$ or
$\iftag{FOL}{\Sat{M}}{\pSat{v}}{!A}$.

If the number of inferences is greater than~0, we distinguish cases
according to the type of the lowermost inference. By induction
hypothesis, we can assume that the premises of that inference are
valid, since the number of inferences in the !!{derivation} of any premise is
smaller than~$n$.

First, we consider the possible inferences with only one premise.
\begin{enumerate}
\item The last inference is a weakening.  Then $\Theta \Sequent \Xi$
  is either $!A, \Gamma \Sequent \Delta$ (if the last inference is
  \LeftR{\Weakening}) or $\Gamma \Sequent \Delta, !A$ (if it's
  \RightR{\Weakening}), and the !!{derivation} ends in one of
  \begin{prooftree}
    \AxiomC{}
    \Deduce$\Gamma \fCenter \Delta$
    \RightLabel{\LeftR{\Weakening}}
    \UnaryInf$!A, \Gamma \fCenter \Delta$
    \DisplayProof\qquad\bottomAlignProof
    \AxiomC{}
    \Deduce$\Gamma \fCenter \Delta$
    \RightLabel{\RightR{\Weakening}}
    \UnaryInf$\Gamma \fCenter \Delta, !A$
  \end{prooftree}
  By induction hypothesis, $\Gamma \Sequent \Delta$ is valid, i.e.,
  for every
  \iftag{FOL}{!!{structure}~$\Struct{M}$}{!!{valuation}~$\pAssign{v}$},
  either there is some $!C \in \Gamma$ such that
  $\iftag{FOL}{\Sat/{M}}{\pSat/{v}}{!C}$ or there is some $!C \in
  \Delta$ such that $\iftag{FOL}{\Sat{M}}{\pSat{v}}{!C}$.
  
  If $\iftag{FOL}{\Sat/{M}}{\pSat/{v}}{!C}$ for some $!C \in \Gamma$,
  then $!C \in \Theta$ as well since $\Theta = !A, \Gamma$, and so
  $\iftag{FOL}{\Sat/{M}}{\pSat/{v}}{!C}$ for some $!C \in \Theta$.
  Similarly, if $\iftag{FOL}{\Sat{M}}{\pSat{v}}{!C}$ for some $!C \in
  \Delta$, as $!C \in \Xi$, $\iftag{FOL}{\Sat{M}}{\pSat{v}}{!C}$ for
  some $!C \in \Xi$. Consequently, $\Theta \Sequent \Xi$ is valid.
\item The last inference is \LeftR{\lnot}: Then the premise of the
  last inference is $\Gamma \Sequent \Delta, !A$ and the conclusion is
  $\lnot !A, \Gamma \Sequent \Delta$, i.e., the !!{derivation} ends in
  \begin{prooftree}
    \AxiomC{}
    \Deduce$\Gamma \fCenter \Delta, !A$
    \RightLabel{\LeftR{\lnot}}
    \UnaryInf$\lnot !A, \Gamma \fCenter \Delta$
  \end{prooftree}
  and $\Theta = \lnot !A, \Gamma$ while $\Xi = \Delta$.

  The induction hypothesis tells us that $\Gamma \Sequent \Delta, !A$
  is valid, i.e., for every \iftag{FOL}{$\Struct{M}$}{$\pAssign{v}$},
  either (a) for some $!C \in \Gamma$,
  $\iftag{FOL}{\Sat/{M}}{\pSat/{v}}{!C}$, or (b) for some $!C \in
  \Delta$, $\iftag{FOL}{\Sat{M}}{\pSat{v}}{!C}$, or (c)
  $\iftag{FOL}{\Sat{M}}{\pSat{v}}{!A}$. We want to show that $\Theta
  \Sequent \Xi$ is also valid.  Let \iftag{FOL}{$\Struct{M}$ be
    !!a{structure}}{$\pAssign{v}$ be !!a{valuation}}. If (a) holds,
  then there is $!C \in \Gamma$ so that
  $\iftag{FOL}{\Sat/{M}}{\pSat/{v}}{!C}$, but $!C \in \Theta$ as
  well. If (b) holds, there is $!C \in \Delta$ such that
  $\iftag{FOL}{\Sat{M}}{\pSat{v}}{!C}$, but $!C \in \Xi$ as
  well. Finally, if $\iftag{FOL}{\Sat{M}}{\pSat{v}}{!A}$, then
  $\iftag{FOL}{\Sat/{M}}{\pSat/{v}}{\lnot !A}$. Since $\lnot !A \in
  \Theta$, there is $!C \in \Theta$ such that
  $\iftag{FOL}{\Sat/{M}}{\pSat/{v}}{!C}$. Consequently, $\Theta
  \Sequent \Xi$ is valid.
\item The last inference is \RightR{\lnot}: Exercise.
\item The last inference is \LeftR{\land}: There are two variants: $!A
  \land !B$ may be inferred on the left from $!A$ or from $!B$ on the
  left side of the premise.  In the first case, the $\pi$ ends in
  \begin{prooftree}
    \AxiomC{}
    \Deduce$!A, \Gamma \fCenter \Delta$
    \RightLabel{\LeftR{\land}}
    \UnaryInf$!A \land !B, \Gamma \fCenter \Delta$
  \end{prooftree}
  and $\Theta = !A \land !B, \Gamma$ while $\Xi = \Delta$.  Consider
  \iftag{FOL}{!!a{structure}~$\Struct
    M$}{!!a{valuation}~$\pAssign{v}$}.  Since by induction hypothesis,
  $!A, \Gamma \Sequent \Delta$ is valid, (a)
  $\iftag{FOL}{\Sat/{M}}{\pSat/{v}}{!A}$, (b)
  $\iftag{FOL}{\Sat/{M}}{\pSat/{v}}{!C}$ for some $!C \in \Gamma$, or
  (c)~$\iftag{FOL}{\Sat{M}}{\pSat{v}}{!C}$ for some $!C \in \Delta$.
  In case (a), $\iftag{FOL}{\Sat/{M}}{\pSat/{v}}{!A \land !B}$, so
  there is $!C \in \Theta$ (namely, $!A \land !B$) such that
  $\iftag{FOL}{\Sat/{M}}{\pSat/{v}}{!C}$.  In case (b), there is $!C
  \in \Gamma$ such that $\iftag{FOL}{\Sat/{M}}{\pSat/{v}}{!C}$, and
  $!C \in \Theta$ as well. In case (c), there is $!C \in \Delta$ such
  that $\iftag{FOL}{\Sat{M}}{\pSat{v}}{!C}$, and $!C \in \Xi$ as well
  since $\Xi = \Delta$. So in each case, \iftag{FOL}{$\Struct
    M$}{$\pAssign{v}$} satisfies $!A \land !B, \Gamma \Sequent
  \Delta$.  Since \iftag{FOL}{$\Struct{M}$}{$\pAssign{v}$} was
  arbitrary, $\Gamma \Sequent \Delta$ is valid.  The case where $!A
  \land !B$ is inferred from $!B$ is handled the same, changing $!A$
  to $!B$.
\item The last inference is \RightR{\lor}: There are two variants: $!A
  \lor !B$ may be inferred on the right from $!A$ or from $!B$ on the
  right side of the premise.  In the first case, $\pi$ ends in
  \begin{prooftree}
    \AxiomC{}
    \Deduce$\Gamma \fCenter \Delta, !A$
    \RightLabel{\RightR{\lor}}
    \UnaryInf$\Gamma \fCenter \Delta, !A \lor !B$
  \end{prooftree}
  Now $\Theta = \Gamma$ and $\Xi = \Delta, !A \lor !B$.  Consider
  \iftag{FOL}{!!a{structure}~$\Struct{M}$}{!!a{valuation}~$\pAssign{v}$}.
  Since $\Gamma \Sequent \Delta, !A$ is valid, (a)
  $\iftag{FOL}{\Sat{M}}{\pSat{v}}{!A}$, (b)
  $\iftag{FOL}{\Sat/{M}}{\pSat/{v}}{!C}$ for some $!C \in \Gamma$, or
  (c)~$\iftag{FOL}{\Sat{M}}{\pSat{v}}{!C}$ for some $!C \in \Delta$.
  In case (a), $\iftag{FOL}{\Sat{M}}{\pSat{v}}{!A \lor !B}$.  In case
  (b), there is $!C \in \Gamma$ such that
  $\iftag{FOL}{\Sat/{M}}{\pSat/{v}}{!C}$.  In case (c), there is $!C
  \in \Delta$ such that $\iftag{FOL}{\Sat{M}}{\pSat{v}}{!C}$. So in
  each case, \iftag{FOL}{$\Struct{M}$}{$\pAssign{v}$} satisfies
  $\Gamma \Sequent \Delta, !A \lor !B$, i.e., $\Theta \Sequent \Xi$.
  Since \iftag{FOL}{$\Struct{M}$}{$\pAssign{v}$} was arbitrary,
  $\Theta \Sequent \Xi$ is valid.  The case where $!A \lor !B$ is
  inferred from $!B$ is handled the same, changing $!A$ to $!B$.
\item The last inference is \RightR{\lif}: Then $\pi$ ends in
  \begin{prooftree}
    \AxiomC{}
    \Deduce$!A, \Gamma \fCenter \Delta, !B$
    \RightLabel{\RightR{\lif}}
    \UnaryInf$\Gamma \fCenter \Delta, !A \lif !B$
  \end{prooftree}
  Again, the induction hypothesis says that the premise is valid; we
  want to show that the conclusion is valid as well. Let
  \iftag{FOL}{$\Struct{M}$}{$\pAssign{v}$} be arbitrary. Since $!A,
  \Gamma \Sequent \Delta, !B$ is valid, at least one of the following
  cases obtains: (a) $\iftag{FOL}{\Sat/{M}}{\pSat/{v}}{!A}$, (b)
  $\iftag{FOL}{\Sat{M}}{\pSat{v}}{!B}$, (c)
  $\iftag{FOL}{\Sat/{M}}{\pSat/{v}}{!C}$ for some~$~!C \in \Gamma$, or
  (d) $\iftag{FOL}{\Sat{M}}{\pSat{v}}{!C}$ for some $!C \in \Delta$.
  In cases (a) and (b), $\iftag{FOL}{\Sat{M}}{\pSat{v}}{!A \lif !B}$
  and so there is a $!C \in \Delta, !A \lif !B$ such that
  $\iftag{FOL}{\Sat{M}}{\pSat{v}}{!C}$.  In case (c), for some $!C \in
  \Gamma$, $\iftag{FOL}{\Sat/{M}}{\pSat/{v}}{!C}$. In case (d), for
  some $!C \in \Delta$, $\iftag{FOL}{\Sat{M}}{\pSat{v}}{!C}$.  In each
  case, \iftag{FOL}{$\Struct{M}$}{$\pAssign{v}$} satisfies $\Gamma
  \Sequent \Delta, !A \lif !B$.  Since
  \iftag{FOL}{$\Struct{M}$}{$\pAssign{v}$} was arbitrary, $\Gamma
  \Sequent \Delta, !A \lif !B$ is valid.  \iftag{FOL}{%
\item The last inference is \LeftR{\lforall}: Then there is
  !!a{formula}~$!A(x)$ and a closed term~$t$ such that $\pi$ ends in
  \begin{prooftree}
    \AxiomC{}
    \Deduce$!A(t), \Gamma \fCenter \Delta$
    \RightLabel{\LeftR{\lforall}}
    \UnaryInf$\lforall[x][!A(x)], \Gamma \fCenter \Delta$
  \end{prooftree}
  We want to show that the conclusion $\lforall[x][!A(x)], \Gamma
  \Sequent \Delta$ is valid.  Consider !!a{structure}~$\Struct M$.
  Since the premise $!A(t), \Gamma \Sequent \Delta$ is valid, (a)
  $\Sat/{M}{!A(t)}$, (b) $\Sat/{M}{!C}$ for some $!C \in \Gamma$, or
  (c)~$\Sat{M}{!C}$ for some $!C \in \Delta$.  In case (a), by
  \olref[syn][sem]{prop:quant-terms}, if
  $\Sat{M}{\lforall[x][!A(x)]}$, then $\Sat{M}{!A(t)}$. Since
  $\Sat/{M}{!A(t)}$, $\Sat/{M}{\lforall[x][!A(x)]}$ . In case (b) and
  (c), $\Struct{M}$ also satisfies $\lforall[x][!A(x)], \Gamma
  \Sequent \Delta$.  Since $\Struct M$ was arbitrary,
  $\lforall[x][!A(x)], \Gamma \Sequent \Delta$ is valid.
\item The last inference is \RightR{\lexists}: Exercise.
\item The last inference is \RightR{\lforall}: Then there is
  !!a{formula}~$!A(x)$ and !!a{constant}~$a$ such that $\pi$ ends in
  \begin{prooftree}
    \AxiomC{}
    \Deduce$\Gamma \fCenter \Delta, !A(a)$
    \RightLabel{\RightR{\lforall}}
    \UnaryInf$\Gamma \fCenter \Delta, \lforall[x][!A(x)]$
  \end{prooftree}
  where the eigenvariable condition is satisfied, i.e., $a$ does not
  occur in $!A(x)$, $\Gamma$, or $\Delta$.  By induction hypothesis,
  the premise of the last inference is valid.  We have to show that
  the conclusion is valid as well, i.e., that for any
  !!{structure}~$\Struct M$, (a) $\Sat{M}{\lforall[x][!A(x)]}$, (b)
  $\Sat/{M}{!C}$ for some $!C \in \Gamma$, or (c)~$\Sat{M}{!C}$ for
  some $!C \in \Delta$.

  Suppose $\Struct{M}$ is an arbitrary !!{structure}. If (b) or (c)
  holds, we are done, so suppose neither holds: for all $!C \in
  \Gamma$, $\Sat{M}{!C}$, and for all $!C \in \Delta$,
  $\Sat/{M}{!C}$. We have to show that (a) holds, i.e.,
  $\Sat{M}{\lforall[x][!A(x)]}$.  By
  \olref[syn][ass]{prop:sat-quant}, if suffices to show that
  $\Sat{M}{!A(x)}[s]$ for all variable assignments~$s$.  So let $s$ be
  an arbitrary variable assignment. Consider the
  structure~$\Struct{M'}$ which is just like~$\Struct{M}$ except
  $\Assign{a}{M'} = s(x)$. By
  \olref[syn][ext]{cor:extensionality-sent}, for any $!C \in \Gamma$,
  $\Sat{M'}{!C}$ since $a$ does not occur in~$\Gamma$, and for any
  $!C \in \Delta$, $\Sat/{M'}{!C}$. But the premise is valid, so
  $\Sat{M'}{!A(a)}$. By \olref[syn][ass]{prop:sentence-sat-true},
  $\Sat{M'}{!A(a)}[s]$, since $!A(a)$ is a sentence. Now $\varAssign{s}{s}{x}$
  with $s(x) = \Value{a}{M'}[s]$, since we've defined $\Struct{M'}$ in
  just this way. So \olref[syn][ext]{prop:ext-formulas} applies, and
  we get $\Sat{M'}{!A(x)}[s]$. Since $a$ does not occur in~$!A(x)$, by
  \olref[syn][ext]{prop:extensionality}, $\Sat{M}{!A(x)}[s]$. Since
  $s$ was arbitrary, we've completed the proof that
  $\Sat{M}{!A(x)}[s]$ for all variable assignments.
\item The last inference is \LeftR{\lexists}: Exercise.
}{}
\end{enumerate}
Now let's consider the possible inferences with two premises.
\begin{enumerate}
\item The last inference is a cut: then $\pi$ ends in
  \begin{prooftree}
    \AxiomC{}
    \Deduce$\Gamma \fCenter \Delta, !A$
    \AxiomC{}
    \Deduce$!A, \Pi \fCenter \Lambda$
    \RightLabel{\Cut}
    \BinaryInf$\Gamma, \Pi \fCenter \Delta, \Lambda$
  \end{prooftree}
  Let \iftag{FOL}{$\Struct{M}$ be !!a{structure}}{$\pAssign{v}$ be
    !!a{valuation}}.  By induction hypothesis, the premises are valid,
  so \iftag{FOL}{$\Struct{M}$}{$\pAssign{v}$} satisfies both premises.
  We distinguish two cases: (a) $\iftag{FOL}{\Sat/{M}}{\pSat/{v}}{!A}$
  and (b) $\iftag{FOL}{\Sat{M}}{\pSat{v}}{!A}$.  In case (a), in order
  for \iftag{FOL}{$\Struct{M}$}{$\pAssign{v}$} to satisfy the left
  premise, it must satisfy $\Gamma \Sequent \Delta$.  But then it also
  satisfies the conclusion.  In case (b), in order for
  \iftag{FOL}{$\Struct{M}$}{$\pAssign{v}$} to satisfy the right
  premise, it must satisfy $\Pi \setminus \Lambda$.  Again,
  \iftag{FOL}{$\Struct{M}$}{$\pAssign{v}$} satisfies the conclusion.
\item The last inference is \RightR{\land}.  Then $\pi$ ends in
  \begin{prooftree}
    \AxiomC{}
    \Deduce$\Gamma \fCenter \Delta, !A$
    \AxiomC{}
    \Deduce$\Gamma \fCenter \Delta, !B$
    \RightLabel{\RightR{\land}}
    \BinaryInf$\Gamma \fCenter \Delta, !A \land !B$
  \end{prooftree}
  Consider \iftag{FOL}{!!a{structure}~$\Struct
    M$}{!!a{valuation}~$\pAssign{v}$}. If
  \iftag{FOL}{$\Struct{M}$}{$\pAssign{v}$} satisfies $\Gamma \Sequent
  \Delta$, we are done. So suppose it doesn't. Since $\Gamma \fCenter
  \Delta, !A$ is valid by induction hypothesis,
  $\iftag{FOL}{\Sat{M}}{\pSat{v}}{!A}$. Similarly, since $\Gamma
  \Sequent \Delta, !B$ is valid,
  $\iftag{FOL}{\Sat{M}}{\pSat{v}}{!B}$. But then
  $\iftag{FOL}{\Sat{M}}{\pSat{v}}{!A \land !B}$.
\item The last inference is \LeftR{\lor}: Exercise.
\item The last inference is \LeftR{\lif}.  Then $\pi$ ends in
  \begin{prooftree}
    \AxiomC{}
    \Deduce$\Gamma \fCenter \Delta, !A$
    \AxiomC{}
    \Deduce$!B, \Pi \fCenter \Lambda$
    \RightLabel{\LeftR{\lif}}
    \BinaryInf$!A \lif !B, \Gamma, \Pi \fCenter \Delta, \Lambda$
  \end{prooftree}
  Again, consider
  \iftag{FOL}{!!a{structure}~$\Struct{M}$}{!!a{valuation}~$\pAssign{v}$}
  and suppose \iftag{FOL}{$\Struct{M}$}{$\pAssign{v}$} doesn't satisfy
  $\Gamma, \Pi \Sequent \Delta, \Lambda$. We have to show that
  $\iftag{FOL}{\Sat/{M}}{\pSat/{v}}{!A \lif !B}$. If
  \iftag{FOL}{$\Struct{M}$}{$\pAssign{v}$} doesn't satisfy $\Gamma,
  \Pi \Sequent \Delta, \Lambda$, it satisfies neither $\Gamma \Sequent
  \Delta$ nor $\Pi \Sequent \Lambda$. Since, $\Gamma \Sequent \Delta,
  !A$ is valid, we have $\iftag{FOL}{\Sat{M}}{\pSat{v}}{!A}$. Since
  $!B, \Pi \Sequent \Lambda$ is valid, we have
  $\iftag{FOL}{\Sat/{M}}{\pSat/{v}}{!B}$. But then
  $\iftag{FOL}{\Sat/{M}}{\pSat/{v}}{!A \lif !B}$, which is what we
  wanted to show.
\end{enumerate}
\end{proof}

\tagprob{FOL}
\begin{prob}
Complete the proof of \olref[fol][seq][sou]{thm:sequent-soundness}.
\end{prob}
\tagendprob

\tagprob{notFOL}
\begin{prob}
Complete the proof of \olref[pl][seq][sou]{thm:sequent-soundness}.
\end{prob}
\tagendprob

\begin{cor}
\ollabel{cor:weak-soundness}
If $\Proves !A$ then $!A$ is \iftag{FOL}{valid}{a tautology}.
\end{cor}

\begin{cor}
\ollabel{cor:entailment-soundness}
If $\Gamma \Proves !A$ then $\Gamma \Entails !A$.
\end{cor}

\begin{proof}
If $\Gamma \Proves !A$ then for some finite subset $\Gamma_0 \subseteq
\Gamma$, there is !!a{derivation} of $\Gamma_0 \Sequent !A$.  By
\olref{thm:sequent-soundness}, every
\iftag{FOL}{!!{structure}~$\Struct{M}$}{!!{valuation}~$\pAssign{v}$}
either makes some $!B \in \Gamma_0$ false or makes $!A$ true.  Hence,
if $\iftag{FOL}{\Sat{M}}{\pSat{v}}{\Gamma}$ then also
$\iftag{FOL}{\Sat{M}}{\pSat{v}}{!A}$.
\end{proof}

\begin{cor}
\ollabel{cor:consistency-soundness}
If $\Gamma$ is satisfiable, then it is consistent.
\end{cor}

\begin{proof}
We prove the contrapositive.  Suppose that $\Gamma$ is not consistent.
Then there is a finite $\Gamma_0 \subseteq \Gamma$ and !!a{derivation}
of $\Gamma_0 \Sequent \quad$.  By \olref{thm:sequent-soundness},
$\Gamma_0 \Sequent \quad$ is valid.  In other words, for every
\iftag{FOL}{!!{structure}~$\Struct{M}$}{!!{valuation}~$\pAssign{v}$},
there is $!C \in \Gamma_0$ so that
$\iftag{FOL}{\Sat/{M}}{\pSat/{v}}{!C}$, and since $\Gamma_0
\subseteq \Gamma$, that $!C$ is also in~$\Gamma$.  Thus, no
\iftag{FOL}{$\Struct{M}$}{$\pAssign{v}$} satisfies~$\Gamma$, and
$\Gamma$ is not satisfiable.
\end{proof}

\end{document}
