% Part: computability
% Chapter: computability-theory
% Section: complement-ce

\documentclass[../../../include/open-logic-section]{subfiles}

\begin{document}

\olfileid{cmp}{thy}{cmp}
\olsection{Computably Enumerable Sets not Closed under Complement}

Suppose $A$ is computably enumerable. Is the complement of $A$,
$\Complement{A} = \Nat \setminus A$, necessarily computably enumerable
as well? The following theorem and corollary show that the answer is
``no.''

\begin{thm}
\ollabel{thm:ce-comp}
Let $A$ be any set of natural numbers. Then $A$ is computable if and
only if both $A$ and $\Complement{A}$ are computably enumerable.
\end{thm}

\begin{proof}
The forwards direction is easy: if $A$ is computable, then
$\Complement{A}$ is computable as well ($\Char{A} = 1 \tsub
\Char{\Complement{A}}$), and so both are computably enumerable.

In the other direction, suppose $A$ and $\Complement{A}$ are both
computably enumerable. Let $A$ be the domain of $\cfind{d}$, and let
$\Complement{A}$ be the domain of $\cfind{e}$. Define $h$ by
\[
h(x) = \umin{s}{(T(d,x,s) \lor T(e,x,s))}.
\]
In other words, on input $x$, $h$ searches for either a halting
computation of~$\cfind{d}$ or a halting computation
of~$\cfind{e}$. Now, if $x \in A$, it will succeed in the first case,
and if $x \in \Complement{A}$, it will succeed in the second case. So,
$h$~is a total computable function. But now we have that for every
$x$, $x \in A$ if and only if $T(e, x, h(x))$, i.e., if $\cfind{e}$ is
the one that is defined. Since $T(e, x, h(x))$ is a computable relation,
$A$~is computable.
\end{proof}

\begin{explain}
It is easier to understand what is going on in informal computational
terms: to decide $A$, on input $x$ search for halting computations of
$\cfind{e}$ and $\cfind{f}$. One of them is bound to halt; if it is $\cfind{e}$,
then $x$ is in~$A$, and otherwise, $x$ is in~$\Complement{A}$.
\end{explain}

\begin{cor}
\ollabel{cor:comp-k}
$\Complement{K_0}$ is not computably enumerable.
\end{cor}

\begin{proof}
We know that $K_0$ is computably enumerable, but not computable. If
$\Complement{K_0}$ were computably enumerable, then $K_0$ would be
computable by \olref{thm:ce-comp}.
\end{proof}

\end{document}
