% Part: applied-modal-logics
% Chapter: epistemic-logic
% Section: public-announcement-logic-semantics

\documentclass[../../../include/open-logic-section]{subfiles}

\begin{document}

\olfileid{aml}{el}{psm}

\olsection{Semantics of Public Announcement Logic}

Relational models for public announcement logics are the same as they
were in epistemic logics. However, the semantics for the public
announcement operator are something new.

\begin{defn}\ollabel{defn:mmodels} \emph{Truth of !!a{formula}~$!A$
  at~$w$} in a~$\mModel M = \tuple{W, R, V}$, in symbols:
  $\mSat{M}{!A}[w]$, is defined inductively as follows:
  \begin{enumerate}
  \tagitem{prvFalse}{\indcase{!A}{\lfalse}{Never $\mSat{M}{\lfalse}[w]$}.}{}
  \tagitem{prvTrue}{\indcase{!A}{\ltrue}{Always $\mSat{M}{\ltrue}[w]$}.}{}
  \item $\mSat{M}{p}[w]$ iff $w \in V(p)$
  \tagitem{prvNot}{\indcase{!A}{\lnot !B}{$\mSat{M}{\indfrm}[w]$ iff
    $\mSat/{M}{!B}[w]$}.}{}
  \tagitem{prvAnd}{\indcase{!A}{(!B \land !C)}{$\mSat{M}{\indfrm}[w]$ iff
    $\mSat{M}{!B}[w]$ and $\mSat{M}{!C}[w]$}.}{}
  \tagitem{prvOr}{\indcase{!A}{(!B \lor !C)}{$\mSat{M}{\indfrm}[w]$ iff
    $\mSat{M}{!B}[w]$ or $\mSat{M}{!C}[w]$} (or both).}{}
  \tagitem{prvIf}{\indcase{!A}{(!B \lif !C)}{$\mSat{M}{\indfrm}[w]$ iff
    $\mSat/{M}{!B}[w]$ or $\mSat{M}{!C}[w]$}.}{}
  \tagitem{prvIff}{\indcase{!A}{(!B \liff !C)}{$\mSat{M}{\indfrm}[w]$ iff
    either both $\mSat{M}{!B}[w]$ and $\mSat{M}{!C}[w]$ or
    neither $\mSat{M}{!B}[w]$ nor $\mSat{M}{!C}[w]$}.}{}
  \item\ollabel{defn:sub:mmodels-box}
    \indcase{!A}{\Knows_a !B}{$\mSat{M}{\indfrm}[w]$ iff
    $\mSat{M}{!B}[w']$ for all $w' \in W$ with $R_a ww'$}
  \item\ollabel{defn:sub:mmodels-pal}
    \indcase{!A}{[!B] !C}{$\mSat{M}{\indfrm}[w]$ iff
    $\mSat{M}{!B}[w]$ implies $\mSat{M \mid !B}{!C}[w]$}

  
  Where $\mModel M \mid !B = \tuple{W', R', V'}$ is defined as
  follows:
  \begin{enumerate}
    \item $W' = \Setabs{ u \in W}{\mSat{M}{!B}[u]}$. So the worlds of
      $\mModel M \mid !B$ are the worlds in~$\mModel M$ at which $!B$
      holds.
    \item $R'_a = R_a \cap (W' \times W')$. Each agent's accessibility
      relation is simply restricted to the worlds that remain in~$W'$.
    \item $V'(p) = \Setabs{ u \in W'}{u \in V(p) }$. Similarly, the
      propositional valuations at worlds remain the same, representing
      the idea that informational events will not change the truth
      value of propositional variables.
  \end{enumerate}
  
  \end{enumerate} 
\end{defn}

What is distinctive, then, about public announcement logics, is that
the truth of !!a{formula} at~$\mModel M$ can sometimes only be decided
by referring to a model other than~$\mModel M$ itself.

Notice also that our semantics treats the announcement operator as a
$\Box$~operator, and so if !!a{formula}~$!A$ cannot be truthfully
announced at a world, then $[!A]B$ will hold there trivially, just as
all $\Box$ !!{formula}s hold at endpoints. 

\begin{figure}
  \begin{center}
    \begin{tikzpicture}[modal]
      \node[world] (w1) [label={below left:$p, \lnot q$}]{$w_1$}; 
      \node[world] (w2) [left=of w1, label={left:$\lnot p, \lnot q$}]{$w_2$}; 
      \node[world] (w3) [above=of w1, label={right:$p, q$}] {$w_3$};
      \draw[<->] (w1) to [swap] node {$b$} (w2);
      \draw[->,loop below] (w1) to node {$a, b$} (w1);
      \draw[<->] (w1) to [swap] node {$a$} (w3);
      \draw[->,loop above] (w2) to node {$a, b$} (w2) ;
      \draw[->,loop above] (w3) to node {$a, b$} (w3) ;
      
      \node[world](v1)[right of=w1, xshift=1.25in, label={right:$p, \lnot q$}]{$w'_1$};
      \node[world](v3)[above of=v1,label={right:$p,q$}]{$w'_3$};
      \draw[<->] (v1) to node {$a$} (v3);
      \draw[->,loop below] (v1) to node {$a,b$} (v1);
      \draw[->,loop above] (v3) to node {$a,b$} (v3) ;
      
      \node(m)[below=of w1]{$\mModel M$};
      \node(m')[below=of v1]{$\mModel M \mid p$}; 

      \draw[-,dotted] (w1) to node {\footnotesize announcement of $p$}(v1) ;
     
    \end{tikzpicture}
  \end{center}
  \caption{Before and after the public announcement of $p$.}
  \ollabel{fig:announcement-example}
\end{figure}

We can see the public announcement of !!a{formula} as shrinking a
model, or restricting it to the worlds at which the !!{formula} was
true. \olref{fig:announcement-example} gives an example of the effects
of publicly announcing~$p$. One notable thing about that model is that
agent~$b$ learns that~$p$ as a result of the announcement, while
agent~$a$ does not (since $a$ already knew that~$p$ was true).

More formally, we have $\mSat{M}{\lnot \Knows_b p}[w_1]$ but $\mSat{M
\mid p}{\Knows_b p}[w'_1]$. This implies that $\mSat{M}{[p] \Knows_b
p}[w_1]$. But we have some even stronger claims that we can make about
the result of the announcement. In fact, it is the case that
$\mSat{M}{[p]\CKnows_{\{a,b\}} p}[w_1]$. In other words, after $p$ is
announced, it becomes \emph{common knowledge}.

We might wonder, though, whether this holds in the general case, and
whether a truthful announcement of~$!A$ will \emph{always} result in
$!A$ becoming common knowledge. It may be surprising that the answer
is no. And in fact, it is possible to truthfully announce !!{formula}s
that will no longer be true once they are announced. For example,
consider the effects of announcing $p \land \lnot \Knows_b p$ at~$w_1$
in \olref{fig:announcement-example}. In fact, $\mModel M \mid p$ and
$\mModel M \mid (p \land \lnot \Knows_b p)$ are the same model.
However, as we have already noted, $\mSat{M \mid p}{\Knows_b
p}[w'_1]$. Therefore, $\mSat{M \mid (p \land \lnot \Knows_b p)}{\lnot
(p \land \lnot \Knows_b p)}[w'_1]$, so this is !!a{formula} that
becomes false once it has been announced. 

\end{document}
