% Part: methods
% Chapter: proofs
% Section: example-2

\documentclass[../../../include/open-logic-section]{subfiles}

\begin{document}

\olfileid{mth}{prf}{ex2}

\olsection{Another Example} 

\begin{prop}
If $X \subseteq Z$, then $X \cup (Z \setminus X) = Z$.
\end{prop}  

\begin{proof}
  Suppose that $X \subseteq Z$.  We want to show that $X \cup (Z
  \setminus X) = Z$.
  \begin{quote}
    We begin by observing that this is a conditional statement. It is
    tacitly universally quantified: the proposition holds for all sets
    $X$ and $Z$. So $X$ and $Z$ are variables for arbitrary sets. To
    prove such a statement, we assume the antecedent and prove the
    consequent.

    We continue by using the assumption that $X \subseteq Z$. Let's
    unpack the definition of~$\subseteq$: the assumption means that
    all !!{element}s of~$X$ are also !!{element}s of~$Z$. Let's write
    this down---it's an important fact that we'll use throughout the
    proof.
  \end{quote}
  By the definition of~$\subseteq$, since $X \subseteq Z$, for all
  $z$, if $z \in X$, then $z \in Z$.
  \begin{quote}
    We've unpacked all the definitions that are given to us in the
    assumption. Now we can move onto the conclusion. We want to show
    that $X \cup (Z \setminus X) = Z$, and so we set up a proof
    similarly to the last example: we show that every !!{element} of
    $X \cup (Z \setminus X)$ is also !!a{element} of~$Z$ and,
    conversely, every !!{element} of $Z$ is !!a{element} of $X \cup (Z
    \setminus X)$. We can shorten this to: $X \cup (Z \setminus X)
    \subseteq Z$ and $Z \subseteq X \cup (Z \setminus X)$. (Here we're
    doing the opposite of unpacking a definition, but it makes the
    proof a bit easier to read.)  Since this is a conjunction, we have
    to prove both parts. To show the first part, i.e., that every
    !!{element} of $X \cup (Z \setminus X)$ is also !!a{element}
    of~$Z$, we assume that $z \in X \cup (Z \setminus X)$ for an
    arbitrary~$z$ and show that $z \in Z$. By the definition of
    $\cup$, we can conclude that $z \in X$ or $z \in Z \setminus X$
    from $z \in X \cup (Z \setminus X)$. You should now be getting the
    hang of this.
    \end{quote}
  $X \cup (Z \setminus X) = Z$ iff $X \cup (Z \setminus X) \subseteq
  Z$ and $Z \subseteq (X \cup (Z \setminus X)$.  First we prove that
  $X \cup (Z \setminus X) \subseteq Z$.  Let $z \in X \cup (Z
  \setminus X)$. So, either $z \in X$ or $z \in (Z \setminus X)$.
  \begin{quote}
    We've arrived at a disjunction, and from it we want to prove that
    $z \in Z$. We do this using proof by cases.
  \end{quote}
  Case 1: $z \in X$. Since for all $z$, if $z \in X$, $z \in Z$, we
  have that $z \in Z$.
  \begin{quote}
    Here we've used the fact recorded earlier which followed from the
    hypothesis of the proposition that $X \subseteq Z$.  The first
    case is complete, and we turn to the second case, $z \in (Z
    \setminus X)$.  Recall that $Z \setminus X$ denotes the
    \emph{difference} of the two sets, i.e., the set of all
    !!{element}s of~$Z$ which are not !!{element}s of~$X$.  But any
    !!{element} of $Z$ not in~$X$ is in particular !!a{element} of~$Z$.
  \end{quote}
  Case 2: $z \in (Z \setminus X)$.  This means that $z \in Z$ and $z
  \notin X$. So, in particular, $z \in Z$.
  \begin{quote}
    Great, we've proved the first direction. Now for the second
    direction. Here we prove that $Z \subseteq X \cup (Z \setminus
    X)$.  So we assume that $z \in Z$ and prove that $z \in X \cup (Z
    \setminus X)$.
  \end{quote}
  Now let $z \in Z$. We want to show that $z \in X$ or $z \in Z
  \setminus X$.
  \begin{quote}
    Since all !!{element}s of $X$ are also !!{element}s of $Z$, and $Z
    \setminus X$ is the set of all things that are !!{element}s of $Z$
    but not $X$, it follows that $z$ is either in $X$ or in $Z
    \setminus X$.  This may be a bit unclear if you don't already know
    why the result is true.  It would be better to prove it
    step-by-step.  It will help to use a simple fact which we can
    state without proof: $z \in X$ or $z \notin X$. This is called the
    ``principle of excluded middle:'' for any statement~$p$, either
    $p$ is true or its negation is true. (Here, $p$ is the statement
    that $z \in X$.)  Since this is a disjunction, we can again use
    proof-by-cases.
  \end{quote}
  Either $z \in X$ or $z \notin X$. In the former case, $z \in X \cup
  (Z \setminus X)$. In the latter case, $z \in Z$ and $z \notin X$, so
  $z \in Z \setminus X$.  But then $z \in X \cup (Z \setminus X)$.
  \begin{quote}
    Our proof is complete: we have shown that $X \cup (Z \setminus X) = Z$.
    \qedhere
  \end{quote}
\end{proof}

\end{document}
