% Part: methods
% Chapter: proofs
% Section: resources

\documentclass[../../../include/open-logic-section]{subfiles}

\begin{document}

\olfileid{mth}{prf}{res}

\section{Other Resources}

There are many books on how to do proofs in mathematics which may be
useful. Check out
\href{https://www.amazon.ca/How-Read-Proofs-Introduction-Mathematical/dp/1118164024/ref=sr_1_1?ie=UTF8&qid=1474314676&sr=8-1&keywords=an+introduction+to+mathematical+thought+process}{\emph{How
  to Read and do Proofs: An Introduction to Mathematical Thought
  Processes}} by Daniel Solow and
\href{https://www.amazon.ca/How-Prove-Structured-Daniel-Velleman/dp/0521675995}{\emph{How
  to Prove It: A Structured Approach}} by Daniel Velleman in
particular.  The
\href{http://www.people.vcu.edu/~rhammack/BookOfProof/BookOfProof.pdf}{\emph{Book
  of Proof}} by Richard Hammack and
\href{http://scholarworks.gvsu.edu/cgi/viewcontent.cgi?article=1009&context=books}{\emph{Mathematical
  Reasoning}} by Ted Sundstrom are books on proof that are freely
available.  Philosophers might find
\href{https://www.amazon.ca/More-Precisely-Math-Need-Philosophy/dp/1551119099/ref=sr_1_1?ie=UTF8&qid=1474314425&sr=8-1&keywords=The+math+you+need+to+do+philosophy}{\emph{More
Precisely: The Math you need to do Philosophy}} by Eric Steinhart to be a
good primer on mathematical reasoning. 

There are also various shorter guides to proofs available on the
internet; e.g.,
\href{https://math.berkeley.edu/~hutching/teach/proofs.pdf}{``Introduction
  to Mathematical Arguments''} by Michael Hutchings and
\href{http://cheng.staff.shef.ac.uk/proofguide/proofguide.pdf}{``How to
  write proofs''} by Eugenia Chang.

\subsection{Motivational Videos}

Feel like you have no motivation to do your homework? Feeling down? These
videos might help!

\begin{itemize}
\item \url{https://www.youtube.com/watch?v=ZXsQAXx_ao0}
\item \url{https://www.youtube.com/watch?v=BQ4yd2W50No}
\item \url{https://www.youtube.com/watch?v=StTqXEQ2l-Y}
\end{itemize}

\end{document}
