% part: intuitionistic-logic
% chapter: soundeness-completeness
% section: soundness-nd

\documentclass[../../../include/open-logic-section]{subfiles}

\begin{document}

\olfileid{int}{sc}{snd}

\olsection{Soundness of Natural Deduction}

We will now prove soundness of natural deduction with regards to the 
relational semantics, that is, showing that if a formula is deducible from
a set of assumptions then the set of assumptions entails the formula. Observe that
when proving that $\Gamma \Entails !A$ we may use the definition
\begin{itemize}
  \item for every model~$\mModel{M}$ and every~$w$ such that
$\mSat{M}{\Gamma}[w]$, $\mSat{M}{!A}[w]$
\end{itemize}
or the result of \olref[int][sem][sem]{prop:sat-entails} and prove
\begin{itemize}
  \item if for every model~$\mModel{M}$ such
  that $\mSat{M}{\Gamma}$, $\mSat{M}{!A}$.
\end{itemize}

\begin{thm}[Soundness]
  \ollabel{thm:soundness}
  If $\Gamma \Proves !A$, then $ \Gamma \Entails !A$.
\end{thm}

\begin{proof}
  We prove that if $\Gamma \Proves !A$, then $\Gamma \Entails !A$.
  The proof is by induction on the !!{derivation} of~$!A$
  from~$\Gamma$.
  
  \begin{enumerate}
  \item If the !!{derivation} consists of just the assumption~$!A$, we
    have $!A \Proves !A$, and want to show that $!A \Entails !A$.
    Consider any model $\mModel{M}$ such that $\mSat{M}{!A}$. Then
    trivially $\mSat{M}{!A}$.

  \item The derivation ends in $\Intro{\land}$:
    \iftag{probAnd}{Exercise.}{The !!{derivation}s of the premises
      $!B$ from !!{undischarged} assumptions~$\Gamma$ and of~$!C$ from
      !!{undischarged} assumptions~$\Delta$ show that $\Gamma \Proves
      !B$ and $\Delta \Proves !C$. By induction hypothesis we have
      that $\Gamma \Entails !B$ and $\Delta \Entails !C$. We have to
      show that $\Gamma \cup \Delta \Entails !A \land !B$, since the
      !!{undischarged} assumptions of the entire derivation are
      $\Gamma$ together with $\Delta$.  So suppose $\mSat{M}{\Gamma
        \cup \Delta}$. Then also $\mSat{M}{\Gamma}$. Since $\Gamma
      \Entails !B$, $\mSat{M}{!B}$. Similarly, $\mSat{M}{!C}$. So
      $\mSat{M}{!B \land !C}$.}

  \item The !!{derivation} ends in $\Elim{\land}$:
    \iftag{probAnd}{Exercise.}{The !!{derivation} of the premise $!B
        \land !C$ from !!{undischarged} assumptions $\Gamma$ shows
        that $\Gamma \Proves !B \land !C$. By induction hypothesis,
        $\Gamma \Entails !B \land !C$. We have to show that $\Gamma
        \Entails !B$. So suppose $\mSat{M}{\Gamma}$. Since $\Gamma
        \Entails !B \land !C$, $\mSat{M}{!B \land !C}$. Then also
        $\mSat{M}{!B}$. Similarly if $\Elim\land$ ends in~$!C$, then
        $\Gamma \Entails !C$.}

  \item The !!{derivation} ends in $\Intro{\lor}$:
    \iftag{probOr}{Exercise.}{Suppose the premise is $!B$, and the
      !!{undischarged} assumptions of the !!{derivation} ending
      in~$!B$ are $\Gamma$. Then we have $\Gamma \Proves !B$ and by
      inductive hypothesis, $\Gamma \Entails !B$. We have to show that
      $\Gamma \Entails !B \lor !C$. Suppose $\mSat{M}{\Gamma}$. Since
      $\Gamma \Entails !B$, $\mSat{M}{!B}$. But then also $\mSat{M}{!B
        \lor !C}$. Similarly, if the premise is~$!C$, we have that
      $\Gamma \Entails !C$.}
    
  \item The !!{derivation} ends in~$\Elim{\lor}$:
    \iftag{probOr}{Exercise.}{The !!{derivation}s ending in the
      premises are of $!B \lor !C$ from !!{undischarged}
      assumptions~$\Gamma$, of $!D$ from !!{undischarged} assumptions
      $\Delta_1 \cup \{!B\}$, and of $!D$ from !!{undischarged}
      assumptions~$\Delta_2 \cup \{!C\}$. So we have $\Gamma \Proves
      !B \lor !C$, $\Delta_1 \cup \{!B\} \Proves !D$, and $\Delta_2
      \cup \{!C\} \Proves !D$.  By induction hypothesis, $\Gamma
      \Entails !B \lor !C$, $\Delta_1 \cup \{!B\} \Entails !D$, and
      $\Delta_2 \cup \{!C\} \Entails !D$.  We have to prove that
      $\Gamma \cup \Delta_1 \cup \Delta_2 \Entails !D$.

    Suppose $\mSat{M}{\Gamma \cup \Delta_1 \cup \Delta_2}$. Then
    $\mSat{M}{\Gamma}$ and since $\Gamma \Entails !B \lor !C$,
    $\mSat{M}{!B \lor !C}$. By definition of $\mSat{M}{}$, either
    $\mSat{M}{!B}$ or $\mSat{M}{!C}$. So we distinguish cases: (a)
    $\mSat{M}{!B}$. Then $\mSat{M}{\Delta_1 \cup \{!B\}}$. Since
    $\Delta_1 \cup !B \Entails !D$, we have $\mSat{M}{!D}$. (b)
    $\mSat{M}{!C}$. Then $\mSat{M}{\Delta_2 \cup \{!C\}}$. Since
    $\Delta_2 \cup !C \Entails !D$, we have $\mSat{M}{!D}$. So in
    either case, $\mSat{M}{!D}$, as we wanted to show.}

  \item The !!{derivation} ends with $\Intro{\lif}$ concluding~$!B\lif
    !C$. Then the premise is~$!C$, and the !!{derivation} ending in
    the premise has !!{undischarged} assumptions~$\Gamma \cup
    \{!B\}$. So we have that $\Gamma \cup \{!B\} \Proves !C$, and by
    induction hypothesis that $\Gamma \cup \{!B\} \Entails !C$. We
    have to show that $\Gamma \Entails !B \lif !C$.

    Suppose $\mSat{M}{\Gamma}[w]$. We want to show that for all
    $w'$ such that $Rww'$, if $\mSat{M}{!B}[w']$, then
    $\mSat{M}{!C}[w']$. So assume that $Rww'$ and
    $\mSat{M}{!B}[w']$. By \olref[sem][rel]{prop:true-monotonic},
    $\mSat{M}{\Gamma}[w']$. Since $\Gamma \cup \{!B\} \Entails !C$,
    $\mSat{M}{!C}[w']$, which is what we wanted to show. 

  \item The !!{derivation} ends in $\Elim{\lif}$ and
    conclusion~$!C$. The premises are $!B \lif !C$ and $!B$, with
    !!{derivation}s from !!{undischarged} assumptions $\Gamma$,
    $\Delta$. So we have $\Gamma \Proves !B \lif !C$ and $\Delta
    \Proves !B$. By inductive hypothesis, $\Gamma \Entails !B \lif !C$
    and $\Delta \Entails !B$. We have to show that $\Gamma \cup \Delta
    \Entails !C$.

    Suppose $\mSat{M}{\Gamma \cup \Delta}[w]$. Since
    $\mSat{M}{\Gamma}[w]$ and $\Gamma \Entails !B \lif !C$,
    $\mSat{M}{!B \lif !C}[w]$. By definition, this means that for all
    $w'$ such that $Rww'$, if $\mSat{M}{!B}[w']$ then
    $\mSat{M}{!C}[w']$. Since $R$ is reflexive, $w$ is among the $w'$
    such that $Rww'$, i.e., we have that if $\mSat{M}{!B}[w]$ then
    $\mSat{M}{!C}[w]$. Since $\mSat{M}{\Delta}[w]$ and $\Delta
    \Entails !B$, $\mSat{M}{!B}[w]$. So, $\mSat{M}{!C}[w]$, as we
    wanted to show.

  \item The !!{derivation} ends in $\FalseInt$, concluding~$!A$. The
    premise is $\lfalse$ and the !!{undischarged} assumptions of the
    !!{derivation} of the premise are~$\Gamma$. Then $\Gamma \Proves
    \lfalse$. By inductive hypothesis, $\Gamma \Entails \lfalse$. We
    have to show $\Gamma \Entails !A$.

    We proceed indirectly. If $\Gamma \Entails/ !A$ there is a
    model~$\mModel{M}$ and world~$w$ such that $\mSat{M}{\Gamma}[w]$
    and $\mSat/{M}{!A}[w]$.  Since $\Gamma \Entails \lfalse$,
    $\mSat{M}{\lfalse}[w]$. But that's impossible, since by
    definition, $\mSat/{M}{\lfalse}[w]$. So $\Gamma \Entails !A$.
  \item The derivation ends in $\Intro\lnot$: Exercise.
  \item The derivation ends in $\Elim\lnot$: Exercise.
  \end{enumerate}
\end{proof}

\begin{prob}
  Complete the proof of \olref[int][sc][snd]{thm:soundness}. For the
  cases for $\Intro\lnot$ and $\Elim\lnot$, use the definition of
  $\mSat{M}{\lnot !A}[w]$ in \olref[int][sem][rel]{defn:true-at-w}, i.e.,
  don't treat $\lnot !A$ as defined by $!A \lif \lfalse$.
\end{prob}

\begin{prob}
Show that the following formulas are not derivable in intuitionistic logic:
\begin{enumerate}
  \item $(!A \lif !B) \lor (!B \lif !A)$
  \item$(\lnot\lnot !A \lif !A) \lif (!A \lor \lnot !A)$
  \item $(!A \lif !B \lor !C) \lif \bigl((!A \lif !B)\lor(!A \lif !C)\bigr)$
\end{enumerate}
\end{prob}

\end{document}
