% Part: turing-machines
% Chapter: machines-computations
% Section: variants

\documentclass[../../../include/open-logic-section]{subfiles}

\begin{document}

\olfileid{tur}{mac}{var}
\olsection{Variants of Turing Machines}

There are in fact many possible ways to define Turing machines, of
which ours is only one.  In some ways, our definition is more liberal
than others. We allow arbitrary finite alphabets, a more restricted
definition might allow only two tape symbols, $\TMstroke$
and~$\TMblank$.  We allow the machine to write a symbol to the tape
and move at the same time, other definitions allow either writing or
moving.  We allow the possibility of writing without moving the tape
head, other definitions leave out the $\TMstay$ ``instruction.''  In
other ways, our definition is more restrictive. We assumed that the
tape is infinite in one direction only, other definitions allow the
tape to be infinite both to the left and the right. In fact, one can
even allow any number of separate tapes, or even an infinite grid of
squares.  We represent the instruction set of the Turing machine by a
transition function; other definitions use a transition relation where
the machine has more than one possible instruction in any given
situation.

This last relaxation of the definition is particularly interesting.
In our definition, when the machine is in state~$q$ reading
symbol~$\sigma$, $\delta(q, \sigma)$ determines what the new symbol,
state, and tape head position is.  But if we allow the instruction set
to be a relation between current state-symbol pairs $\tuple{q,
  \sigma}$ and new state-symbol-direction triples $\tuple{q', \sigma',
  D}$, the action of the Turing machine may not be uniquely
determined---the instruction relation may contain both $\tuple{q,
  \sigma, q', \sigma', D}$ and $\tuple{q, \sigma, q'', \sigma'', D'}$.
In this case we have a \emph{non-deterministic} Turing machine.  These
play an important role in computational complexity theory.

There are also different conventions for when a Turing machine halts:
we say it halts when the transition function is undefined, other
definitions require the machine to be in a special designated halting
state. We have explained in \olref[hal]{sec} why requiring a
designated halting state is not a restriction which impacts what
Turing machines can compute.  Since the tapes of our Turing machines
are infinite in one direction only, there are cases where a Turing
machine can't properly carry out an instruction: if it reads the
leftmost square and is supposed to move left. According to our
definition, it just stays put instead of ``falling off'', but we could
have defined it so that it halts when that happens. This definition is
also equivalent: we could simulate the behavior of a Turing machine
that halts when it attempts to move left from square~$0$ by deleting
every transition $\delta(q,\TMendtape) =
\tuple{q',\sigma,\TMleft}$---then instead of attempting to move left
on~$\TMendtape$ the machine halts.\footnote{This doesn't \emph{quite}
work, since nothing prevents us from writing and reading $\TMendtape$
on squares other than square~$0$ (see \olref[una]{ex:mover}). We can
get around that by adding a second~$\TMendtape'$ symbol to use instead
for such a purpose.}

There are also different ways of representing numbers (and hence the
input-output function computed by a Turing machine): we use unary
representation, but you can also use binary representation. This
requires two symbols in addition to $\TMblank$ and~$\TMendtape$.

Now here is an interesting fact: none of these variations matters as
to which functions are Turing computable. \emph{If a function is Turing
computable according to one definition, it is Turing computable
according to all of them.}

We won't go into the details of verifying this. Here's just one
example: we gain no additional computing power by allowing a tape that
is infinite in both directions, or multiple tapes. The reason is,
roughly, that a Turing machine with a single one-way infinite tape can
simulate multiple or two-way infinite tapes. E.g., using additional
states and instructions, we can ``translate'' a program for a
machine with multiple tapes or two-way infinite tape into one with a
single one-way infinite tape.  The translated machine can use the
even squares for the squares of tape~$1$ (or the ``positive'' squares
of a two-way infinite tape) and the odd squares for the squares of
tape~$2$ (or the ``negative'' squares).

\end{document}
